### 2015-03-23 tex/Book/gpbook-Tozier-2015-03-23/chapter.tex
merge Nic's LaTeX fixes in

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title*{Genetic Programming As If You Meant It}
% Use \titlerunning{} for an abbreviated version of
% your contribution title if the original one is too long
\author{Bill Tozier}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Name of First Author \at Name, Address of Institute}

\maketitle

\abstract{Each chapter should be preceded by an abstract (10--15 lines long) that summarizes the content. The abstract will appear \textit{online} at \url{www.SpringerLink.com} and be available with unrestricted access. This allows unregistered users to read the abstract as a teaser for the complete chapter. As a general rule the abstracts will not appear in the printed version of your book unless it is the style of your particular book or that of the series to which your book belongs.}

\begin{keywords}
keywords to your chapter, these words should also be indexed
\end{keywords}
\index{keywords to your chapter}
\index{these words should also be indexed}
\\
\section{Section Heading}
\label{sec:1}
% Always give a unique label
% and use \ref{<label>} for cross-references
% and \cite{<label>} for bibliographic references
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations. And please note that the first line of
text that follows a heading is not indented, whereas the first lines of
all subsequent paragraphs are.

\section{Section Heading}
\label{sec:2}
% Always give a unique label
% and use \ref{<label>} for cross-references
% and \cite{<label>} for bibliographic references
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations.

Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.

Use the standard \verb|equation| environment to typeset your equations, e.g.
%
\begin{equation}
a \times b = c\;,
\end{equation}
%
however, for multiline equations we recommend to use the \verb|eqnarray| environment.
\begin{eqnarray}
a \times b = c \nonumber\\
\vec{a} \cdot \vec{b}=\vec{c}
\label{eq:01}
\end{eqnarray}

\subsection{Subsection Heading}
\label{subsec:2}
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references\index{cross-references} and citations\index{citations}
as has already been described in Sect.~\ref{sec:2}.

\begin{quotation}
Please do not use quotation marks when quoting texts! Simply use the \verb|quotation| environment -- it will automatically render Springer's preferred layout.
\end{quotation}

\subsubsection{Subsubsection Heading}
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations as has already been described in
Sect.~\ref{subsec:2}, see also Fig.~\ref{fig:1}\footnote{Footnotes are easily added with this simple command.}

Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.

% For figures use
%
\begin{figure}[b] %[b] sets the image at the bottom of the page; t = top, b = bottom, h = here%
\sidecaption
% Use the relevant command for your figure-insertion program
% to insert the figure file.
% For example, with the graphicx style use
\includegraphics[scale=.65]{figure}
%
% If no graphics program available, insert a blank space i.e. use
%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
%
\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page. If the figure is positioned at the top of the page, align the sidecaption with the top of the figure -- to achieve this you simply need to use the optional argument \texttt{[t]} with the \texttt{sidecaption} command}
\label{fig:1}       % Give a unique label
\end{figure}

\paragraph{Paragraph Heading} %
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations as has already been described in
Sect.~\ref{sec:2}.

Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.

For typesetting numbered lists we recommend to use the \verb|enumerate| environment -- it will automatically render Springer's preferred layout.

\begin{enumerate}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\begin{enumerate}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\end{enumerate}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\end{enumerate}

\subparagraph{Subparagraph Heading} In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text. Use the \LaTeX\ automatism for all your cross-references and citations as has already been described in Sect.~\ref{sec:2}, see also Fig.~\ref{fig:2}.

For unnumbered list we recommend to use the \verb|itemize| environment -- it will automatically render Springer's preferred layout.

\begin{itemize}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development, cf. Table~\ref{tab:1}.}
\begin{itemize}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\end{itemize}
\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
\end{itemize}

\begin{figure}[t] %[t] sets the image at the top of the page; t = top, b = bottom, h = here%
\sidecaption[t]
% Use the relevant command for your figure-insertion program
% to insert the figure file.
% For example, with the option graphics use
\includegraphics[scale=.65]{figure}
%
% If no graphics program available, insert a blank space i.e. use
%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
%
%\caption{Please write your figure caption here}
\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page. If the figure is positioned at the top of the page, align the sidecaption with the top of the figure -- to achieve this you simply need to use the optional argument \texttt{[t]} with the \texttt{sidecaption} command}
\label{fig:2}       % Give a unique label
\end{figure}

% Use the \index{} command to code your index words
% Make sure to inlcude the indexed word inside and outside of the brackets if you want the text to show up in your paragraph:
% e.g. This book is about \index{genetic programming}genetic programming. 
% If the text is not entered outside the brackets it will appear as: This book is about .
%
% For tables use
%
\begin{table}
\caption{Please write your table caption here}
\label{tab:1}       % Give a unique label
%
% Follow this input for your own table layout
%
\begin{tabular}{p{2cm}p{2.4cm}p{2cm}p{4.9cm}}
\hline\noalign{\smallskip}
Classes & Subclass & Length & Action Mechanism  \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Translation & mRNA$^a$  & 22 (19--25) & Translation repression, mRNA cleavage\\
Translation & mRNA cleavage & 21 & mRNA cleavage\\
Translation & mRNA  & 21--22 & mRNA cleavage\\
Translation & mRNA  & 24--26 & Histone and DNA Modification\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
$^a$ Table foot note (with superscript)
\end{table}
%
\section{Section Heading}
\label{sec:3}
% Always give a unique label
% and use \ref{<label>} for cross-references
% and \cite{<label>} for bibliographic references
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations as has already been described in
Sect.~\ref{sec:2}.

Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.

If you want to list definitions or the like we recommend to use the Springer-enhanced \verb|description| environment -- it will automatically render Springer's preferred layout.

\begin{description}[Type 1]
\item[Type 1]{That addresses central themes pertainng to migration, health, and disease. In Sect.~\ref{sec:1}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
\item[Type 2]{That addresses central themes pertainng to migration, health, and disease. In Sect.~\ref{subsec:2}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
\end{description}

\subsection{Subsection Heading} %
In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text. Use the \LaTeX\ automatism for all your cross-references and citations citations as has already been described in Sect.~\ref{sec:2}.

Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.

\begin{svgraybox}
If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option \verb|graybox| and the newly defined environment \verb|svgraybox|. This will produce a 15 percent screened box 'behind' your text.

If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option and environment \verb|svgraybox|. This will produce a 15 percent screened box 'behind' your text.
\end{svgraybox}

\subsubsection{Subsubsection Heading}
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations as has already been described in
Sect.~\ref{sec:2}.

Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.

\begin{theorem}
Theorem text goes here.
\end{theorem}
%
% or
%
\begin{definition}
Definition text goes here.
\end{definition}

\begin{proof}
%\smartqed
Proof text goes here.
\qed
\end{proof}

\paragraph{Paragraph Heading} %
Instead of simply listing headings of different levels we recommend to
let every heading be followed by at least a short passage of text.
Further on please use the \LaTeX\ automatism for all your
cross-references and citations as has already been described in
Sect.~\ref{sec:2}.

Note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
%
% For built-in environments use
%
\begin{theorem}
Theorem text goes here.
\end{theorem}
%
\begin{definition}
Definition text goes here.
\end{definition}
%
\begin{proof}
\smartqed
Proof text goes here.
\qed
\end{proof}
%
\begin{acknowledgement}
If you want to include acknowledgments of assistance and the like at the end of an individual chapter please use the \verb|acknowledgement| environment -- it will automatically render Springer's preferred layout.
\end{acknowledgement}
%
\paragraph{Referencing}
This is how to reference. \cite{Kadie}

\bibliographystyle{spbasic}
\bibliography{gp-bibliography,chapter}

### 2015-03-23 markdown/manuscript.tex
towards kramdown integration

\section{``TDD as if you meant it''}\hypertarget{tdd-as-if-you-meant-it}{}\label{tdd-as-if-you-meant-it}

As far as I can tell, Keith Braithwaite first described his \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{training exercise} for software developers in 2009. The target of the exercise is ``Pseudo-TDD'': the noted habit among software developers who claim to ``know'' and ``do'' test-driven development as part of their daily work towards a sort of thoughtless approximation of the technique.

I should note that a number of agile software development practices share informative relations to genetic programming's dynamics\footnote{I imagine there is an Engineering Studies thesis in this for some aspiring graduate student: Genetic programming and agile development practices arose in the same period and more or less the same culture, and both informed by the same currents in complex systems and emergent approaches to problem-solving.}, but in this work I'll focus on those of TDD. In particular, test-driven development (or more accurately ``test-driven design'') \emph{when done correctly} can break down the complex design space of a software project into a value-ordered set of incremental test cases, focus the developers' attention on those cases alone, inhibit unnecessary ``code bloat'' and feature creep, and produce low-complexity understandable and maintainable software.

TDD \emph{as such} is a rigorous process, to the point where it can be described as ``painful'' (though also ``useful'') by experienced programmers. The steps are deceptively easy to trivialize and misunderstand, especially for those whose habits of thinking about code are ingrained:

\begin{enumerate}
\item Add a little test
\item Run all tests and fail
\item Make a little change
\item Run the tests and succeed
\item Refactor to remove duplication
\end{enumerate}

Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the overarching flow of implementation (or ``design'') that it imposes: Each cycle, and also the overarching design process, begins with the choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced over all iterations of this cycle. The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working through them as if they could be automated easily. The \emph{mindfulness} of the process lives in the choice of next steps and (though somewhat less so) of standard refactoring operations.

Braithwaite's exercise does an interesting thing to surface the formal rigor of this approach. In it, the participants (willing, of course, because the exercise is a \emph{kata} or ``refresher'' for experienced software developers to hone their skills) are asked to implement a nominally simple project like the game of Tic-Tac-Toe, given an \emph{ordered} list of features to implement and the artificial restriction that they must go farther than normal TDD practice asks. Rather than producing a suite of tests and a self-contained codebase, they are forced to use \emph{only} refactoring of code added to tests to produce their eventual ``codebase''. In other words, no code can be ``produced'' until \emph{duplication in code added multiple passing tests} provides a warrant for refactoring it out. Further, a facilitator patrols ongoing work and deletes \emph{any and all code not called for by a pre-existing failing test}.

Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens\ldots{} as one might imagine. But as Gojko Adzik emphasizes in his descriptions of workshops he's run, the results of ``design'' of even simple algorithms in this artificial amplified setting seems much more \emph{open-ended} than it would if the software were built within the typical framing of habits and assumptions that an experienced programmer carries along with her to any project.

A number of contextually positive benefits are attributed to agile software development practices, and to TDD within that suite of practices. But the one that brings us here today is that aspect surfaced particularly in Adzik's account of Tic-tac-toe:

\begin{quotation}
By the end of the exercise, almost half the teams were coding towards something that was not a $3\times 3$ char/int grid. We did not have the time to finish the whole thing, but some interesting solutions in making were:

\begin{itemize}
\item a bag of fields that are literally taken by players---field objects start in the collection belonging to the game and move to collections belonging to players, which simply avoids edge cases such as taking an already taken field and makes checking for game end criteria very easy.
\item fields that have logic whether they are taken or not and by whom
\item game with a current state field that was recalculated as the actions were performed on it and methods that could set this externally to make it easy to test
\end{itemize}
\end{quotation}

In other words: innovative approaches to the problem at hand began to arise, though there wasn't enough time to finish them in the time alloted for the exercise. The analogy to our collective experience to date with genetic programming should start to peek through at this point---though the thoughtful reader will hopefully wonder what utility there is in an analogy drawn between two similarly unsatisfying outcomes.

\section{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}

\section{What it can mean to ``use'' genetic programming}\hypertarget{what-it-can-mean-to-use-genetic-programming}{}\label{what-it-can-mean-to-use-genetic-programming}

\section{Cases for and against ``usable GP'': cost, features, reliability, simplicity}\hypertarget{cases-for-and-against-usable-gp-cost-features-reliability-simplicity}{}\label{cases-for-and-against-usable-gp-cost-features-reliability-simplicity}

\section{``Artificial Intelligence investigated by Special Crimes Unit''; or, Even Athena needed the head of Zeus}\hypertarget{artificial-intelligence-investigated-by-special-crimes-unit-or-even-athena-needed-the-head-of-zeus}{}\label{artificial-intelligence-investigated-by-special-crimes-unit-or-even-athena-needed-the-head-of-zeus}

\section{The mangle of practice and the mangle in practice}\hypertarget{the-mangle-of-practice-and-the-mangle-in-practice}{}\label{the-mangle-of-practice-and-the-mangle-in-practice}

\section{The yardstick and the bamboo hand}\hypertarget{the-yardstick-and-the-bamboo-hand}{}\label{the-yardstick-and-the-bamboo-hand}

\section{Leveraging ``resistance'': the problem of the dots and lines}\hypertarget{leveraging-resistance-the-problem-of-the-dots-and-lines}{}\label{leveraging-resistance-the-problem-of-the-dots-and-lines}

\section{``Batch'' and ``interactive'' styles}\hypertarget{batch-and-interactive-styles}{}\label{batch-and-interactive-styles}

\section{Expressing ``resistance'': the problem of Bertrand's Paradox}\hypertarget{expressing-resistance-the-problem-of-bertrands-paradox}{}\label{expressing-resistance-the-problem-of-bertrands-paradox}

\section{Exploration and exploitation interfaces and affordances}\hypertarget{exploration-and-exploitation-interfaces-and-affordances}{}\label{exploration-and-exploitation-interfaces-and-affordances}

\section{The deeper question: What should it mean to \emph{act intelligently}?}\hypertarget{the-deeper-question-what-should-it-mean-to-act-intelligently}{}\label{the-deeper-question-what-should-it-mean-to-act-intelligently}

### 2015-03-23 markdown/readme.tex
towards kramdown integration

\section{MS Prep for Genetic Programming Theory \& Practice 2015}\hypertarget{ms-prep-for-genetic-programming-theory--practice-2015}{}\label{ms-prep-for-genetic-programming-theory--practice-2015}

\subsection{In brief}\hypertarget{in-brief}{}\label{in-brief}

``Genetic programming as if you meant it: real and imaginary user experience''

My title is a direct nod to \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{Keith Braithwaite's} exercise for software developers, ``TDD as if you meant it''. Like Keith's exercise, what I describe here for genetic programming will seem onerous and difficult to most experienced theoreticians or practitioners. I'll make the case that this approach suggests---and demonstrates---a broader systematic framework in which genetic programming is far more likely to \emph{provide value to its users} than we find in a quarter-decade of exploration. Along the way, I'm forced to address deep philosophical problems in the program of genetic progamming research and practice more broadly: \emph{viz.}, what it is ``for'' and when it is ``done''.

### 2015-03-25 markdown/manuscript.tex
\import package for \subimporting rendered markdown; some cleanup

@@
+\section{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}
+
+Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'', also abbreviated ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---least of all---optimization.
+A case could be made that the resistance we have all recounted towards GP from the technical and lay audience has little to do with the technical results so amply demonstrated over the last quarter-century, but rather from discomfort among that audience in GP's particular ``way of working'' on problems.
+There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are ``rigorous'', or even successful, only when they proceed through ordered phases of
+
+\begin{enumerate}
+\item conceptualization, the ``vision thing''
+\item planning
+\item design
+\item architecture
+\item implementation
+\item testing
+\item debugging
+\end{enumerate}
+
+Very few scientists or engineers would admit that any \emph{real} project has followed this narrative structure in practice, but in every phase of work life from fund-raising to published reports the story is framed in that way, especially in scientific work: Because of work gone before, an idea or insight was had.
+The idea was framed as a formal hypothesis.
+The hypothesis (and current Best Statistical Practices) suggested an experimental design obvious to any in the discipline.
+The experimental design was undertaken, the data collected, the hypothesis tested, and we can be confident of its veracity because, well, you just heard me say ``Best Statistical Practices'', right?
+
+Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe any institutional project management and public policy planning process as well.
+The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.
+
+Of course, nobody ``really believes'' this narrative who has ever done any of the work.
+It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and more today.
+
+Andrew Pickering's is the one I'd like to use here.
+
 \section{``TDD as if you meant it''}\hypertarget{tdd-as-if-you-meant-it}{}\label{tdd-as-if-you-meant-it}
@@
 
-\section{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}
-
 \section{What it can mean to ``use'' genetic programming}\hypertarget{what-it-can-mean-to-use-genetic-programming}{}\label{what-it-can-mean-to-use-genetic-programming}

### 2015-03-25 tex/Book/gpbook-Tozier-2015-03-23/chapter.tex
\import package for \subimporting rendered markdown; some cleanup

@@
 
-\title*{Genetic Programming As If You Meant It}
-% Use \titlerunning{} for an abbreviated version of
+\title*{GP As If You Meant It: Real and Imaginary User Experience}
+\titlerunning{GP As If You Meant It}
+
 % your contribution title if the original one is too long
-\author{Bill Tozier}
-% Use \authorrunning{Short Title} for an abbreviated version of
-% your contribution title if the original one is too long
-\institute{Name of First Author \at Name, Address of Institute}
+\author{William A. Tozier}
+\authorrunning{Bill Tozier}
+
+\institute{William A. Tozier}
 
@@
 \\
-\section{Section Heading}
-\label{sec:1}
-% Always give a unique label
-% and use \ref{<label>} for cross-references
-% and \cite{<label>} for bibliographic references
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations. And please note that the first line of
-text that follows a heading is not indented, whereas the first lines of
-all subsequent paragraphs are.
 
-\section{Section Heading}
-\label{sec:2}
-% Always give a unique label
-% and use \ref{<label>} for cross-references
-% and \cite{<label>} for bibliographic references
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations.
+\subimport{../../../markdown/}{manuscript}
 
-Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-
-Use the standard \verb|equation| environment to typeset your equations, e.g.
+%\section{Section Heading}
+%\label{sec:2}
+%% Always give a unique label
+%% and use \ref{<label>} for cross-references
+%% and \cite{<label>} for bibliographic references
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references and citations.
 %
-\begin{equation}
-a \times b = c\;,
-\end{equation}
+%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
 %
-however, for multiline equations we recommend to use the \verb|eqnarray| environment.
-\begin{eqnarray}
-a \times b = c \nonumber\\
-\vec{a} \cdot \vec{b}=\vec{c}
-\label{eq:01}
-\end{eqnarray}
-
-\subsection{Subsection Heading}
-\label{subsec:2}
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references\index{cross-references} and citations\index{citations}
-as has already been described in Sect.~\ref{sec:2}.
-
-\begin{quotation}
-Please do not use quotation marks when quoting texts! Simply use the \verb|quotation| environment -- it will automatically render Springer's preferred layout.
-\end{quotation}
-
-\subsubsection{Subsubsection Heading}
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations as has already been described in
-Sect.~\ref{subsec:2}, see also Fig.~\ref{fig:1}\footnote{Footnotes are easily added with this simple command.}
-
-Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-
-% For figures use
+%Use the standard \verb|equation| environment to typeset your equations, e.g.
+%%
+%\begin{equation}
+%a \times b = c\;,
+%\end{equation}
+%%
+%however, for multiline equations we recommend to use the \verb|eqnarray| environment.
+%\begin{eqnarray}
+%a \times b = c \nonumber\\
+%\vec{a} \cdot \vec{b}=\vec{c}
+%\label{eq:01}
+%\end{eqnarray}
 %
-\begin{figure}[b] %[b] sets the image at the bottom of the page; t = top, b = bottom, h = here%
-\sidecaption
-% Use the relevant command for your figure-insertion program
-% to insert the figure file.
-% For example, with the graphicx style use
-\includegraphics[scale=.65]{figure}
+%\subsection{Subsection Heading}
+%\label{subsec:2}
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references\index{cross-references} and citations\index{citations}
+%as has already been described in Sect.~\ref{sec:2}.
 %
-% If no graphics program available, insert a blank space i.e. use
-%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
+%\begin{quotation}
+%Please do not use quotation marks when quoting texts! Simply use the \verb|quotation| environment -- it will automatically render Springer's preferred layout.
+%\end{quotation}
 %
-\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page.
+%
+%\subsubsection{Subsubsection Heading}
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references and citations as has already been described in
+%Sect.~\ref{subsec:2}, see also Fig.~\ref{fig:1}\footnote{Footnotes are easily added with this simple command.}
+%
+%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
+%
+%% For figures use
+%%
+%\begin{figure}[b] %[b] sets the image at the bottom of the page; t = top, b = bottom, h = here%
+%\sidecaption
+%% Use the relevant command for your figure-insertion program
+%% to insert the figure file.
+%% For example, with the graphicx style use
+%\includegraphics[scale=.65]{figure}
+%%
+%% If no graphics program available, insert a blank space i.e. use
+%%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
+%%
+%\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page.
 If the figure is positioned at the top of the page, align the sidecaption with the top of the figure -- to achieve this you simply need to use the optional argument \texttt{[t]} with the \texttt{sidecaption} command}
-\label{fig:1}       % Give a unique label
-\end{figure}
-
-\paragraph{Paragraph Heading} %
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations as has already been described in
-Sect.~\ref{sec:2}.
-
-Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-
-For typesetting numbered lists we recommend to use the \verb|enumerate| environment -- it will automatically render Springer's preferred layout.
-
-\begin{enumerate}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\begin{enumerate}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\end{enumerate}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\end{enumerate}
-
-\subparagraph{Subparagraph Heading} In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text.
+%\label{fig:1}       % Give a unique label
+%\end{figure}
+%
+%
+%\paragraph{Paragraph Heading} %
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references and citations as has already been described in
+%Sect.~\ref{sec:2}.
+%
+%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
+%
+%For typesetting numbered lists we recommend to use the \verb|enumerate| environment -- it will automatically render Springer's preferred layout.
+%
+%\begin{enumerate}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\begin{enumerate}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\end{enumerate}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\end{enumerate}
+%
+%
+%\subparagraph{Subparagraph Heading} In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text.
 Use the \LaTeX\ automatism for all your cross-references and citations as has already been described in Sect.~\ref{sec:2}, see also Fig.~\ref{fig:2}.
-
-For unnumbered list we recommend to use the \verb|itemize| environment -- it will automatically render Springer's preferred layout.
-
-\begin{itemize}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development, cf. Table~\ref{tab:1}.}
-\begin{itemize}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\end{itemize}
-\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-\end{itemize}
-
-\begin{figure}[t] %[t] sets the image at the top of the page; t = top, b = bottom, h = here%
-\sidecaption[t]
-% Use the relevant command for your figure-insertion program
-% to insert the figure file.
-% For example, with the option graphics use
-\includegraphics[scale=.65]{figure}
 %
-% If no graphics program available, insert a blank space i.e. use
-%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
+%For unnumbered list we recommend to use the \verb|itemize| environment -- it will automatically render Springer's preferred layout.
 %
-%\caption{Please write your figure caption here}
-\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page.
+%\begin{itemize}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development, cf. Table~\ref{tab:1}.}
+%\begin{itemize}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\end{itemize}
+%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
+%\end{itemize}
+%
+%\begin{figure}[t] %[t] sets the image at the top of the page; t = top, b = bottom, h = here%
+%\sidecaption[t]
+%% Use the relevant command for your figure-insertion program
+%% to insert the figure file.
+%% For example, with the option graphics use
+%\includegraphics[scale=.65]{figure}
+%%
+%% If no graphics program available, insert a blank space i.e. use
+%%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
+%%
+%%\caption{Please write your figure caption here}
+%\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page.
 If the figure is positioned at the top of the page, align the sidecaption with the top of the figure -- to achieve this you simply need to use the optional argument \texttt{[t]} with the \texttt{sidecaption} command}
-\label{fig:2}       % Give a unique label
-\end{figure}
-
-% Use the \index{} command to code your index words
-% Make sure to inlcude the indexed word inside and outside of the brackets if you want the text to show up in your paragraph:
-% e.g. This book is about \index{genetic programming}genetic programming. 
-% If the text is not entered outside the brackets it will appear as: This book is about .
+%\label{fig:2}       % Give a unique label
+%\end{figure}
 %
-% For tables use
+%% Use the \index{} command to code your index words
+%% Make sure to inlcude the indexed word inside and outside of the brackets if you want the text to show up in your paragraph:
+%% e.g. This book is about \index{genetic programming}genetic programming. 
+%% If the text is not entered outside the brackets it will appear as: This book is about .
+%%
+%% For tables use
+%%
+%\begin{table}
+%\caption{Please write your table caption here}
+%\label{tab:1}       % Give a unique label
+%%
+%% Follow this input for your own table layout
+%%
+%\begin{tabular}{p{2cm}p{2.4cm}p{2cm}p{4.9cm}}
+%\hline\noalign{\smallskip}
+%Classes & Subclass & Length & Action Mechanism  \\
+%\noalign{\smallskip}\svhline\noalign{\smallskip}
+%Translation & mRNA$^a$  & 22 (19--25) & Translation repression, mRNA cleavage\\
+%Translation & mRNA cleavage & 21 & mRNA cleavage\\
+%Translation & mRNA  & 21--22 & mRNA cleavage\\
+%Translation & mRNA  & 24--26 & Histone and DNA Modification\\
+%\noalign{\smallskip}\hline\noalign{\smallskip}
+%\end{tabular}
+%$^a$ Table foot note (with superscript)
+%\end{table}
+%%
+%\section{Section Heading}
+%\label{sec:3}
+%% Always give a unique label
+%% and use \ref{<label>} for cross-references
+%% and \cite{<label>} for bibliographic references
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references and citations as has already been described in
+%Sect.~\ref{sec:2}.
 %
-\begin{table}
-\caption{Please write your table caption here}
-\label{tab:1}       % Give a unique label
+%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
 %
-% Follow this input for your own table layout
+%If you want to list definitions or the like we recommend to use the Springer-enhanced \verb|description| environment -- it will automatically render Springer's preferred layout.
 %
-\begin{tabular}{p{2cm}p{2.4cm}p{2cm}p{4.9cm}}
-\hline\noalign{\smallskip}
-Classes & Subclass & Length & Action Mechanism  \\
-\noalign{\smallskip}\svhline\noalign{\smallskip}
-Translation & mRNA$^a$  & 22 (19--25) & Translation repression, mRNA cleavage\\
-Translation & mRNA cleavage & 21 & mRNA cleavage\\
-Translation & mRNA  & 21--22 & mRNA cleavage\\
-Translation & mRNA  & 24--26 & Histone and DNA Modification\\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\end{tabular}
-$^a$ Table foot note (with superscript)
-\end{table}
+%\begin{description}[Type 1]
+%\item[Type 1]{That addresses central themes pertainng to migration, health, and disease.
+In Sect.~\ref{sec:1}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
+%\item[Type 2]{That addresses central themes pertainng to migration, health, and disease.
+In Sect.~\ref{subsec:2}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
+%\end{description}
 %
-\section{Section Heading}
-\label{sec:3}
-% Always give a unique label
-% and use \ref{<label>} for cross-references
-% and \cite{<label>} for bibliographic references
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations as has already been described in
-Sect.~\ref{sec:2}.
-
-Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-
-If you want to list definitions or the like we recommend to use the Springer-enhanced \verb|description| environment -- it will automatically render Springer's preferred layout.
-
-\begin{description}[Type 1]
-\item[Type 1]{That addresses central themes pertainng to migration, health, and disease.
-In Sect.~\ref{sec:1}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
-\item[Type 2]{That addresses central themes pertainng to migration, health, and disease.
-In Sect.~\ref{subsec:2}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
-\end{description}
-
-\subsection{Subsection Heading} %
-In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text.
+%\subsection{Subsection Heading} %
+%In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text.
 Use the \LaTeX\ automatism for all your cross-references and citations citations as has already been described in Sect.~\ref{sec:2}.
-
-Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-
-\begin{svgraybox}
-If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option \verb|graybox| and the newly defined environment \verb|svgraybox|.
+%
+%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
+%
+%\begin{svgraybox}
+%If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option \verb|graybox| and the newly defined environment \verb|svgraybox|.
 This will produce a 15 percent screened box 'behind' your text.
-
-If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option and environment \verb|svgraybox|.
+%
+%If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option and environment \verb|svgraybox|.
 This will produce a 15 percent screened box 'behind' your text.
-\end{svgraybox}
-
-\subsubsection{Subsubsection Heading}
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations as has already been described in
-Sect.~\ref{sec:2}.
-
-Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-
-\begin{theorem}
-Theorem text goes here.
-\end{theorem}
+%\end{svgraybox}
 %
-% or
 %
-\begin{definition}
-Definition text goes here.
-\end{definition}
-
-\begin{proof}
+%\subsubsection{Subsubsection Heading}
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references and citations as has already been described in
+%Sect.~\ref{sec:2}.
+%
+%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
+%
+%\begin{theorem}
+%Theorem text goes here.
+%\end{theorem}
+%%
+%% or
+%%
+%\begin{definition}
+%Definition text goes here.
+%\end{definition}
+%
+%\begin{proof}
+%%\smartqed
+%Proof text goes here.
+%\qed
+%\end{proof}
+%
+%\paragraph{Paragraph Heading} %
+%Instead of simply listing headings of different levels we recommend to
+%let every heading be followed by at least a short passage of text.
+%Further on please use the \LaTeX\ automatism for all your
+%cross-references and citations as has already been described in
+%Sect.~\ref{sec:2}.
+%
+%Note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
+%%
+%% For built-in environments use
+%%
+%\begin{theorem}
+%Theorem text goes here.
+%\end{theorem}
+%%
+%\begin{definition}
+%Definition text goes here.
+%\end{definition}
+%%
+%\begin{proof}
 %\smartqed
-Proof text goes here.
-\qed
-\end{proof}
-
-\paragraph{Paragraph Heading} %
-Instead of simply listing headings of different levels we recommend to
-let every heading be followed by at least a short passage of text.
-Further on please use the \LaTeX\ automatism for all your
-cross-references and citations as has already been described in
-Sect.~\ref{sec:2}.
-
-Note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-% For built-in environments use
-%
-\begin{theorem}
-Theorem text goes here.
-\end{theorem}
-%
-\begin{definition}
-Definition text goes here.
-\end{definition}
-%
-\begin{proof}
-\smartqed
-Proof text goes here.
-\qed
-\end{proof}
-%
-\begin{acknowledgement}
-If you want to include acknowledgments of assistance and the like at the end of an individual chapter please use the \verb|acknowledgement| environment -- it will automatically render Springer's preferred layout.
-\end{acknowledgement}
-%
-\paragraph{Referencing}
-This is how to reference. \cite{Kadie}
+%Proof text goes here.
+%\qed
+%\end{proof}
+%%
+%\begin{acknowledgement}
+%If you want to include acknowledgments of assistance and the like at the end of an individual chapter please use the \verb|acknowledgement| environment -- it will automatically render Springer's preferred layout.
+%\end{acknowledgement}
+%%
+%\paragraph{Referencing}
+%This is how to reference. \cite{Kadie}
 

### 2015-03-25 markdown/manuscript.tex
a bit more intro

@@
-\section{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}
+\section{Why}\hypertarget{why}{}\label{why}
 
-Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'', also abbreviated ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---least of all---optimization.
+More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved evolutionary algorithms and genetic programming.
+As we were chatting one day, Bill asked Rick what he'd most like to see as part of the research program of GP ``moving forward''.
+Rick's answer informs this contribution, as well as much of my work since.
+
+As I recall, he said he'd like to better understand the ``symptoms'' we often see when we run an evolutionary search process: premature convergence, failure to improve, the sense when we look at results or ongoing runs that the many parameters \emph{just aren't quite right}.
+Years later, I take this not only to mean that a benchmarking catalog listing conditions in which search operator $X$ under contingency $Y$ tends to produce outcome $Z$ might be helpful, but also that \emph{the symptoms themselves} are poorly understood.
+
+The ``field'' of GP has grown quite a bit in the intervening 15 years.
+All of us who follow it closely can see the expanding front of new methodologies and techniques---the ones we discuss in our little workshop, and also the newsworthy and admirable public successes.
+But those advances bring an accompanying dilution of the scope of the theoretical warrants we use to justify them.
+Each example of ``practice'' seems to pose unique domain-specific quirks, each implementation makes numerous contingent (and often arbitrary) design and architectural choices, and even the ontological foundations of ``individual'', ``fitness'', ``behavior'' and ``population'' start to get sloppy around the edges whenever we actually look at What People Actually Do.
+
+The lack of interest in GP among statisticians, planners, designers and mathematical programming aficionados is not just a matter of disciplinary boundary-setting.
+I imagine much of the skepticism arises from our own inability to say what's happening in a GP run (\emph{even when we look}), let alone why, or what to do in response.
+We have learned that \emph{more CPU time} can get us past the stage of ``not working'', but I argue that we have not been able to satisfactorily explain \emph{why} for any reasonable and interesting example.
+
+I also argue that this is not the \emph{fault} of GP.
+Rather it's the source of GP's long-term survivability as an approach to scientific and engineering projects.
+In making this argument, I'm forced to stray into philosophical territory before coming back to build a concrete example.
+But along that wandering path there are several points where we can productively address Rick's concerns.
+
+\section{How we treat GP (and how it treats us in turn)}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}
+
+Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'' and also abbreviate as ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---though least of all---optimization.
 A case could be made that the resistance we have all recounted towards GP from the technical and lay audience has little to do with the technical results so amply demonstrated over the last quarter-century, but rather from discomfort among that audience in GP's particular ``way of working'' on problems.
@@
 
-Andrew Pickering's is the one I'd like to use here.
+\subsection{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}
+
+Andrew Pickering's \emph{Mangle of Practice} is the one I'd like to use here.
+This approach is noteworthy because it is so close to our actual experience using GP, of the ``didn't we already know this?'' sort.
+At the risk of eliding a lot of extraordinarily well-considered structure, let me summarize:
+
+The act of ``doing science'' is \emph{at no point whatsoever} a behavior undertaken by an isolated ``researcher'' in an objective field of externalities.
+Rather, the researcher who initiates this process with a vision, intuition, hypothesis or hunch begins always by \emph{making some artifact}: code, an instrument, a proof, a sketch, a maquette or even a thoughtful conversation at a conference.
+
+The thing, inevitably, \emph{resists}.
+That is, Pickering grants it agency, or rather makes it the agent of all those externalities that impinge on the work to make the \emph{vision} differ from \emph{practice}: the facts of the actual world, the cultural assumptions and habits of discipline, the raw materials and toolkit available to the practitioner, and so forth.
+``Resistance'' here is not merely the obvious trouble of a software bug or some missing parts, but includes one's sense \emph{on seeing it} that something's not quite right; the realization that more (or less) is needed.
+The language we use, in the face of this resistance, is that the \emph{thing made} ``feels wrong'' or ``points something out'', that it ``wants to do X instead of Y'', or that ``it's doing something too complicated for me to understand right now''.
+
+And---assuming science is indeed what's being done---the researcher \emph{changes in response to this resistance}.
+That vision changes, the plan adapts, or in some other way the \emph{thing made} causes a response in the state of the researcher herself.
+Pickering's Mangle is this emergent dance of inanimate agency: the researcher starting to follow a vision by making (or altering) a \emph{thing}, and the \emph{thing made} in turn acting as a channel for the world itself to steer the researcher in another direction.
+
+The ``mangling'' of Pickering's metaphoric name refers not to wounding but to the mechanical laundry apparatus, the antique wringer through which sheets and linens are run to drain the water, and which as a \emph{side-effect} impose a novel higher-dimensional structure and juxtapose unexpected components with one another.
+Here, the response of the researcher to resistance she experiences in the course of her project is no less a part of the dynamics of ``science'' than the act of writing and running code, the authority warranted by particular statistical practices in her discipline, or the raw pressure of physical laws.
+
+\subsection{GP as ``mangle-ish'' practice}\hypertarget{gp-as-mangle-ish-practice}{}\label{gp-as-mangle-ish-practice}
+
+The broader field of machine learning leans seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions.
+The result of training a neural network or random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the sufficiently robust result of applying numerical optimization to a particular mathematical programming model.
+Whether one describes the process as minimizing out-of-sample error or maximizing information gain, the focus of the discipline is on contingent reliability rather than exploratory modeling.
+
+GP has the capacity to \emph{tell us stories}, even in the ``simple'' domain of symbolic regression problems.
+The space under consideration is not merely a vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, and of functions over inputs, and of higher-order functions over those.
+While GP can be used to explore arbitrarily close to a paradigmatic model, we argue for it most when its application can produce unexpected insights.
+A number of us treat it as the best candidate for ``real'' artificial intelligence, and rightly so.
+
+We can do that because GP surfaces Pickering's Mangle.
+When it ``works'', it does so by offering \emph{helpful resistance} in our engagement with it, whether in the form of surprising answers, validation of our suspicions, or suggestions of ways to make subsequent moves.
+It \emph{dances} with us, in a way that the other mere tools of machine learning do not.
+
+Now a great deal of the last quarter-decade of work in the field, especially in the early days, seems to treat GP as a close relative of other machine learning techniques---as a methodology for producing \emph{unsurprising} models from data, and within the traditional model of scientific work.
+That is, an implicit and idealized user is expected to proceed something like the users of any other machine learning framework:
+
+\begin{enumerate}
+\item frame your problem in the framework-specific language
+\item ``get'' a GP system
+\item ``run GP on the data''
+\item ???
+\item you have solved your problem
+\end{enumerate}
+
+There's strong pressure from our actual user community to enforce this stance, not least because it is exactly the stance of planning and public policy, and of the mythic science or programming project manager.
+That is, it frames GP as a \emph{tool} to be invoked in a known and well-described planning situation.
+
+The resulting resistance from early-adopting ``users'' who ``get'' an off-the-shelf GP system and try to apply it to their problem---at least within this narrative of work---should not be unexpected.
+\emph{Being surprising} is perhaps the worst choice for any traditional project plan.
+It's no wonder that so much of GP research is focused on particular tweaks and staged horse-races among the unnumbered techniques: as long as there is a sentiment that GP might be ``tamed'' so that step (4) above \emph{never happens}, so that the framework might join its more traditional and popular relatives, there will be strong pressure to calculate statistics on, for example ``proportion of successes in 30 replicates''.
+
+But what does a ``replicate'' mean for somebody using GP for a real project, whether theoretical or practical, where \emph{GP itself} is not the focus of the work?
+Projects which authentically ``use'' GP must necessarily be searching for noteworthy answers, which is to say \emph{surprising} and \emph{interesting} answers, that they cannot otherwise obtain.
+To do otherwise is to somehow expect GP to be a form of ``research automation'', a premature expectation in light of our experience.
+
+Therefore, any researcher who is using GP \emph{realistically} is one who is watching, and adjusting, and engaging and interacting with the process of search itself.
+Seen as working within a traditional (and deeply misleading) narrative framework, we watch as she runs a population of 100 individuals for 100 generations, peers at the results in a CSV file and finds them wanting, and then adjusts the GP parameters to run another 100 generations\ldots{} and repeating as necessary.
+But there is \emph{no discernible difference} when we frame the same process as one of participating in the single application of an interactive algorithm that (remarkably, and unaccountably when you think about it) \emph{throws away all intermediate results every 100 generations}.
+
+Further, let's explore that traditional narrative which sees her workflow as a series of repeated attempts to use GP as a tool in a situation for which it is poorly suited, or maybe ``badly tuned''.
+She is carefully ``not interfering'' with any given run of 100 generations, observing our tradition in this field, and can only peer into the results file after the fact.
+During the course of any 100 generations, all sorts of dynamics have happened: crossover, mutation, selection, all the many random choices.
+Imagine for a moment we were given perfect access to the entire dynamical pedigree of the unsatisfying results she receives at the end, and were able to backtrack to any point in the run and change a single decision.
+Before that point, it's unclear how badly things will actually turn out at the 100-generation mark; at some point after that juncture, it's obvious to anybody watching that the whole thing's a mess.
+
+If such miraculous insights were available, then surely the correct approach would be to intervene and adjust the situation when the crucial point was reached\ldots{} and then continue.
+Lacking (as we do) this miraculous insight, \emph{why then does it seem reasonable to stop any run arbitrarily at a pre-ordained time point and begin again from scratch?}
+
+It is, I think, because the myths of artificial intelligence and vision-driven science are deeply intertwined.
+It is somehow ``cheating'' to admit in a scientific paper, even if no mistakes were made, that the original vision and plan changed over the course of the project; rather we describe research \emph{results} as inevitable outcomes of an ahistorical process that erases the work actually done by human beings.
+Similarly, it is somehow ``cheating'' to admit in a GP project, even if every parameter was set correctly, that the original vision and plan gave way to the inevitable surprises thrown up by GP's tremendous potential to surprise.
+
+But insofar as GP \emph{surprises us}---and that is its sole strength over more predictable and manageable frameworks---we will inevitably see a good fraction of the surprises as \emph{disappointments} rather than opportunities to adapt our selves.
 

### 2015-03-25 markdown/manuscript.tex
exposition exposition exposition

@@
 
-\subsection{GP as ``mangle-ish'' practice}\hypertarget{gp-as-mangle-ish-practice}{}\label{gp-as-mangle-ish-practice}
+\subsection{GP as ``mangle-ish practice''}\hypertarget{gp-as-mangle-ish-practice}{}\label{gp-as-mangle-ish-practice}
 
@@
 
-\section{What it can mean to ``use'' genetic programming}\hypertarget{what-it-can-mean-to-use-genetic-programming}{}\label{what-it-can-mean-to-use-genetic-programming}
+[more here]
 
-\section{Cases for and against ``usable GP'': cost, features, reliability, simplicity}\hypertarget{cases-for-and-against-usable-gp-cost-features-reliability-simplicity}{}\label{cases-for-and-against-usable-gp-cost-features-reliability-simplicity}
+\section{As if we \emph{really} meant it}\hypertarget{as-if-we-really-meant-it}{}\label{as-if-we-really-meant-it}
 
-\section{``Artificial Intelligence investigated by Special Crimes Unit''; or, Even Athena needed the head of Zeus}\hypertarget{artificial-intelligence-investigated-by-special-crimes-unit-or-even-athena-needed-the-head-of-zeus}{}\label{artificial-intelligence-investigated-by-special-crimes-unit-or-even-athena-needed-the-head-of-zeus}
+In the same way that Braithwaite's onerous coding exercise drives the attention of its participants toward test-driven design with the obligation to write ``real'' code \emph{only as a refactoring}, I will here demand a \emph{warrant} for every step that moves the algorithm away from random guessing.
+Not only do traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent, data-driven argument in favor of starting them \emph{as part of an ongoing search process}.
+Similarly, the initial selection criteria will be limited to a minimal subset of the training data, and expansion (and other alterations) of the training set will have to be made in light of measured progress, not assumptions that ``more is better'' in every case.
 
-\section{The mangle of practice and the mangle in practice}\hypertarget{the-mangle-of-practice-and-the-mangle-in-practice}{}\label{the-mangle-of-practice-and-the-mangle-in-practice}
+The result will be an incremental process of refinement of an ongoing search, carried out not at the level of externally-assigned parameter ``tweaks'' but rather by \emph{opening} the black boxes we typically demand and demanding we do surgery to correct their ``pathologies'' (and understand their mysteries) without killing them outright.
+It is not intended as an ``algorithm'' to supplant those used today, but rather as a forced re-description of what we actually already do.
 
-\section{The yardstick and the bamboo hand}\hypertarget{the-yardstick-and-the-bamboo-hand}{}\label{the-yardstick-and-the-bamboo-hand}
+\subsection{A tableau representation for GP systems}\hypertarget{a-tableau-representation-for-gp-systems}{}\label{a-tableau-representation-for-gp-systems}
+
+It will help if I present a bit of formalism.
+
+Operators
+Individuals
+Rubrics
+
+\subsubsection{``traditional GP tableau''}\hypertarget{traditional-gp-tableau}{}\label{traditional-gp-tableau}
+
+The analogy to SQL.
+
+\subsubsection{``lexicase selection tableau''}\hypertarget{lexicase-selection-tableau}{}\label{lexicase-selection-tableau}
+
+The analogy to SQL.
+
+\subsubsection{``lazy lexicase selection tableau''}\hypertarget{lazy-lexicase-selection-tableau}{}\label{lazy-lexicase-selection-tableau}
+
+The analogy to NoSQL.
+
+\subsubsection{``mutually lazy tableau''}\hypertarget{mutually-lazy-tableau}{}\label{mutually-lazy-tableau}
+
+``Send me a message.''
+
+\subsubsection{``as if we really meant it''}\hypertarget{as-if-we-really-meant-it-1}{}\label{as-if-we-really-meant-it-1}
+
+I need your help.
 
@@
 
-\section{``Batch'' and ``interactive'' styles}\hypertarget{batch-and-interactive-styles}{}\label{batch-and-interactive-styles}
+\subsection{dots and lines, in pushforth}\hypertarget{dots-and-lines-in-pushforth}{}\label{dots-and-lines-in-pushforth}
 
-\section{Expressing ``resistance'': the problem of Bertrand's Paradox}\hypertarget{expressing-resistance-the-problem-of-bertrands-paradox}{}\label{expressing-resistance-the-problem-of-bertrands-paradox}
+\subsection{the problem of the unknown language}\hypertarget{the-problem-of-the-unknown-language}{}\label{the-problem-of-the-unknown-language}
+
+\subsection{the problem of the objective(s)}\hypertarget{the-problem-of-the-objectives}{}\label{the-problem-of-the-objectives}
+
+\subsection{expanding the goals when it's boring}\hypertarget{expanding-the-goals-when-its-boring}{}\label{expanding-the-goals-when-its-boring}
 
@@
 
-\section{The deeper question: What should it mean to \emph{act intelligently}?}\hypertarget{the-deeper-question-what-should-it-mean-to-act-intelligently}{}\label{the-deeper-question-what-should-it-mean-to-act-intelligently}
+\section{Final thoughts: What should it mean to \emph{act intelligently}?}\hypertarget{final-thoughts-what-should-it-mean-to-act-intelligently}{}\label{final-thoughts-what-should-it-mean-to-act-intelligently}
+
+(It would mean the sort of self-awareness it takes to notice that something is wrong, and to ask for help.)

### 2015-03-27 markdown/manuscript.tex
second pass, first part

@@
+% kramdown -o latex markdown/manuscript.md > markdown/manuscript.tex
+
 \section{Why}\hypertarget{why}{}\label{why}
@@
 As we were chatting one day, Bill asked Rick what he'd most like to see as part of the research program of GP ``moving forward''.
-Rick's answer informs this contribution, as well as much of my work since.
+Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.
 
-As I recall, he said he'd like to better understand the ``symptoms'' we often see when we run an evolutionary search process: premature convergence, failure to improve, the sense when we look at results or ongoing runs that the many parameters \emph{just aren't quite right}.
-Years later, I take this not only to mean that a benchmarking catalog listing conditions in which search operator $X$ under contingency $Y$ tends to produce outcome $Z$ might be helpful, but also that \emph{the symptoms themselves} are poorly understood.
+As I recall, Rick answered that he'd like to better understand the ``symptoms'' we often see when we run an evolutionary search process: premature convergence, failure to improve, catastrophic lack of diversity, and the sense when we look at results or ongoing runs that some particular choices we've made setting the many parameters \emph{just aren't quite right}.
+There are numerous well-written papers suggesting ways around local minima, proposing reasonable and exotic search operators, and running ``horse races'' between variant algorithms on benchmark problems.
+But years later, I take Rick's challenge not only to mean that it might be useful to have a benchmarking catalog which lists conditions in which search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood, to the point that we think of them as ``subjective''.
 
-The ``field'' of GP has grown quite a bit in the intervening 15 years.
-All of us who follow it closely can see the expanding front of new methodologies and techniques---the ones we discuss in our little workshop, and also the newsworthy and admirable public successes.
-But those advances bring an accompanying dilution of the scope of the theoretical warrants we use to justify them.
-Each example of ``practice'' seems to pose unique domain-specific quirks, each implementation makes numerous contingent (and often arbitrary) design and architectural choices, and even the ontological foundations of ``individual'', ``fitness'', ``behavior'' and ``population'' start to get sloppy around the edges whenever we actually look at What People Actually Do.
+Somewhere during that same project, I remember a Project Manager telling the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were confused.
+``We don't want a choice, we want \emph{the best}.'' This came after months of analysis, interviews, and a collective agreement that the problem was fundamentally multi-objective.
+But when the decision-makers were faced with the unquestionably successful results they had collectively specified, \emph{those results weren't right enough}.
 
-The lack of interest in GP among statisticians, planners, designers and mathematical programming aficionados is not just a matter of disciplinary boundary-setting.
-I imagine much of the skepticism arises from our own inability to say what's happening in a GP run (\emph{even when we look}), let alone why, or what to do in response.
-We have learned that \emph{more CPU time} can get us past the stage of ``not working'', but I argue that we have not been able to satisfactorily explain \emph{why} for any reasonable and interesting example.
+In this contribution we must face the fact that \emph{this response is inevitable}.
+Not just from silly lay ``customers'' in an application project, but from ourselves.
+Whether the project is ``theoretical'' or ``practical''.
+Whether it is ``small'' or ``large''.
 
-I also argue that this is not the \emph{fault} of GP.
-Rather it's the source of GP's long-term survivability as an approach to scientific and engineering projects.
-In making this argument, I'm forced to stray into philosophical territory before coming back to build a concrete example.
-But along that wandering path there are several points where we can productively address Rick's concerns.
+The ``field'' of GP has grown quite a bit in the intervening 15 years since these two anecdotes.
+All of us who follow it closely have played a role in the expanding front of new methodologies and techniques---the extraordinary ones we describe in our little workshop, and also the newsworthy and admirable public successes.
+
+But this acceleration of practical successes has been accompanied by a dilution of the scope of the theoretical warrants we use to justify them: Most examples of ``practical success'' involve unique domain-specific quirks, each implementation makes numerous contingent (and sometimes arbitrary) design and architectural choices, and even the ontological foundations of ``individual'', ``fitness'', ``behavior'' and ``population'' start to get sloppy around the edges whenever we actually \emph{look} at What People Actually Do to Solve Problems With GP.
+
+While this can be disappointing for those of us who follow the progress of theoretical work closely, it has more serious consequences.
+The lack of interest in GP among statisticians, planners, designers and mathematical programming aficionados is, I argue, not just a matter of disciplinary border wars.
+I suspect rather that skepticism arises from our own inability to say what's \emph{happening} in any particular GP run---\emph{even when we look}.
+Nor for that matter can we be very helpful in explaining \emph{why}, or what to do in response to any given contingency.
+We have learned that \emph{more CPU time} can sometimes get us past a tricky stage of ``not working'', but any project that lives very far from the bounds of the last century's version of ``genetic programming'' is unexplored territory.
+
+This is not the fault of GP as a field, nor of our theoreticians and engineers.
+Rather I'll argue that a deep philosophical misconception presides over the work, both within and outside our field.
+This misconception undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.
+
+I will make my case here in the form of an exercise, or \emph{kata}.
+This is not intended to be a ``thought-experiment'', nor is it a suggestion for a new way of working.
+Rather it is a formal exercise to be undertaken by those of us already working closely with GP systems.
+In the immediate case, my intent is to surface the three problems I've identified above:
+
+\begin{enumerate}
+\item What GP \emph{does}
+\item How GP \emph{unfolds}
+\item What GP is \emph{for}
+\end{enumerate}
+
+The rules for this exercise, ``Doing GP as if you meant it'', will feel the most artificial and restrictive to those of us with the most experience.
+Like the martial arts exercises from which it is ultimately inspired, it isn't intended to be simple or even pleasant for the participants.
+But because it emphasizes the interaction between the GP system and the user, it not only surfaces the philosophical mistake I've mentioned, but immediately suggests tools with which we can address Rick's decade-old wish.
 
@@
 
-Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'' and also abbreviate as ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---though least of all---optimization.
+Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'' and will also abbreviate as ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---though least of all---optimization.
 A case could be made that the resistance we have all recounted towards GP from the technical and lay audience has little to do with the technical results so amply demonstrated over the last quarter-century, but rather from discomfort among that audience in GP's particular ``way of working'' on problems.
@@
 
-In the same way that Braithwaite's onerous coding exercise drives the attention of its participants toward test-driven design with the obligation to write ``real'' code \emph{only as a refactoring}, I will here demand a \emph{warrant} for every step that moves the algorithm away from random guessing.
+In the same way that Braithwaite's onerous coding exercise is intended to drive the attention of its participants toward test-driven design with its obligation to write ``real'' code \emph{only as a refactoring}, I'd like to be able to demand a \emph{warrant} for every step that moves our changing genetic programming setup away from just plain random guessing.
+Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed}.\footnote{In much the same way that Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong'', I imagine many GP users might say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''.
+I offer no particular justification for either anecdotal stance here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}
+
 Not only do traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent, data-driven argument in favor of starting them \emph{as part of an ongoing search process}.
@@
 
-\subsection{A tableau representation for GP systems}\hypertarget{a-tableau-representation-for-gp-systems}{}\label{a-tableau-representation-for-gp-systems}
+\subsection{The tableau representation of GP systems}\hypertarget{the-tableau-representation-of-gp-systems}{}\label{the-tableau-representation-of-gp-systems}
 
-It will help if I present a bit of formalism.
+I find it helps to present a simplified but formalized description of GP systems, and one which highlights the particular features I'm considering.
 
-Operators
-Individuals
-Rubrics
+At the highest level of abstraction, we will treat GP as a collection of particular \emph{decisions} made by the user, plus an otherwise autonomous stochastic process executed by software and hardware, which can be ``started'', ``paused'' and ``resumed'' but which cannot be \emph{restarted}.
 
-\subsubsection{``traditional GP tableau''}\hypertarget{traditional-gp-tableau}{}\label{traditional-gp-tableau}
+The decisions available to the user apply to three core components of the stochastic process: {\tt operators}, {\tt answers}, and {\tt rubrics}.
+
+\subsubsection{Operators}\hypertarget{operators}{}\label{operators}
+
+An {\tt operator} is any function which takes as argument a (possibly empty) collection of {\tt answers} and produces a new collection of {\tt answers}.
+Operators thus include ``random guessing'', which in GP systems is often used to build an initial population, any ``crossover'' and ``mutation'' functions, but also any arbitrarily complex function which might be used to \emph{create a new answer}.
+So for example ``particle swarm on an abstract expression tree'' would be an {\tt operator} working on a single {\tt answer} and producing a very large collection of results.
+
+In the exercise (but not in the examples that follow) \emph{every intermediate result} which is an {\tt answer} is considered to be part of the return value.
+That is, {\tt operators} are obliged to return all intermediate {\tt answers} they produce while building their ``actual'' results: a function that implements ``crossover-and-mutation'' will is obliged to return \emph{all} crossover products it builds, in addition to the results of a subsequent internal mutation.
+
+Note that no process is provided for {\tt answers} to be \emph{removed} from a tableau. This framework is purely cumulative.
+
+\subsubsection{Answers}\hypertarget{answers}{}\label{answers}
+
+An {\tt answer} is what might traditionally be called an ``individual'' in GP literature, though there are subtle differences.
+We can model them programmatically as key--value hash, typically beginning with a ``genome'' or ``script'' field set to an abstracted representation for a solution to the problem at hand.
+As an {\tt answer} ``matures'' in the unfolding tableau, various  other attributes will be appended and set by other algorithmic processes, such as fitness scores or measured attributes like ``age'' or ``alive?'' states.
+
+In the tableau layout, we will represent the unfolding collection of {\tt answers} as the \emph{rows} of a spreadsheet-like table, and their attributes and scores as the \emph{columns}.
+
+\subsubsection{Rubrics}\hypertarget{rubrics}{}\label{rubrics}
+
+A {\tt rubric} is any function which returns a scalar numerical value, given arguments of a collection of one or more {\tt answers} (and possibly additional arguments), and assigned to a particular {\tt answer}.
+
+Insofar as any given {\tt rubric} is associated with an objective of search, it should be framed as a \emph{minimization} form.
+For errors, this should be obvious; for {\tt rubrics} used in ALPS-like systems, realize that the desired {\tt rubric} to select \emph{younger} {\tt answers} is not ``age'' but ``youth''.
+
+In our final exercise (though not in the examples to follow) there will also be a strict requirement that any attribute associated with an {\tt answer} is also available as an \emph{implicit} {\tt rubric}: the script, the creation time (if recorded), or any other {\tt rubric}.
+We will not permit {\tt rubrics} to store intermediate values, \emph{except in other rubrics}: thus a {\tt rubric} which specifies ``sum squared error over a training set'' cannot be applied without also \emph{first} creating {\tt rubrics} for ``measured error when provided input $i$'' every element $i$ of the training set.
+If the training set has 100 elements, then the SSE {\tt rubric} implicitly represents a suite of 101.
+
+However, in the examples to follow that sketch ``traditional GP'', we can relax this restriction.
+
+\subsubsection{Search process}\hypertarget{search-process}{}\label{search-process}
+
+\subsubsection{``traditional GP'' tableau}\hypertarget{traditional-gp-tableau}{}\label{traditional-gp-tableau}
+
+It's tempting to be glib about defining ``traditional GP'', and for the sake of brevity let me succumb to that temptation: Say it is a fixed-size population of 100 {\tt answers} which are created initially at random and subsequently by crossover and mutation, a single {\tt rubric} which returns a single SSE score calculated over input--output pairs measured over a static collection of training cases.
+
+\begin{longtable}{|l|l|l|l|}
+\hline
+individual & genome & SSE & (notes)\\
+0.1 & \ldots{} & 0.2 & \\
+0.2 & \ldots{} & 0.8 & \\
+\ldots{} &  &  & \\
+0.N & \ldots{} & 0.1 & \\
+\hline
+\end{longtable}
 

### 2015-03-27 tex/Book/gpbook-Tozier-2015-03-23/chapter.tex
a bit more intro
second pass, first part

@@
 \author{William A. Tozier}
-\authorrunning{Bill Tozier}
+\authorrunning{Tozier}
 
-\institute{William A. Tozier}
+%\institute{William A. Tozier}
 
@@
 
-\abstract{Each chapter should be preceded by an abstract (10--15 lines long) that summarizes the content.
-The abstract will appear \textit{online} at \url{www.SpringerLink.com} and be available with unrestricted access.
-This allows unregistered users to read the abstract as a teaser for the complete chapter.
-As a general rule the abstracts will not appear in the printed version of your book unless it is the style of your particular book or that of the series to which your book belongs.}
+\abstract{(to be written last)}
 
@@
 
-%\section{Section Heading}
-%\label{sec:2}
-%% Always give a unique label
-%% and use \ref{<label>} for cross-references
-%% and \cite{<label>} for bibliographic references
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references and citations.
-%
-%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-%Use the standard \verb|equation| environment to typeset your equations, e.g.
-%%
-%\begin{equation}
-%a \times b = c\;,
-%\end{equation}
-%%
-%however, for multiline equations we recommend to use the \verb|eqnarray| environment.
-%\begin{eqnarray}
-%a \times b = c \nonumber\\
-%\vec{a} \cdot \vec{b}=\vec{c}
-%\label{eq:01}
-%\end{eqnarray}
-%
-%\subsection{Subsection Heading}
-%\label{subsec:2}
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references\index{cross-references} and citations\index{citations}
-%as has already been described in Sect.~\ref{sec:2}.
-%
-%\begin{quotation}
-%Please do not use quotation marks when quoting texts! Simply use the \verb|quotation| environment -- it will automatically render Springer's preferred layout.
-%\end{quotation}
-%
-%
-%\subsubsection{Subsubsection Heading}
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references and citations as has already been described in
-%Sect.~\ref{subsec:2}, see also Fig.~\ref{fig:1}\footnote{Footnotes are easily added with this simple command.}
-%
-%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-%% For figures use
-%%
-%\begin{figure}[b] %[b] sets the image at the bottom of the page; t = top, b = bottom, h = here%
-%\sidecaption
-%% Use the relevant command for your figure-insertion program
-%% to insert the figure file.
-%% For example, with the graphicx style use
-%\includegraphics[scale=.65]{figure}
-%%
-%% If no graphics program available, insert a blank space i.e. use
-%%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
-%%
-%\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page.
-If the figure is positioned at the top of the page, align the sidecaption with the top of the figure -- to achieve this you simply need to use the optional argument \texttt{[t]} with the \texttt{sidecaption} command}
-%\label{fig:1}       % Give a unique label
-%\end{figure}
-%
-%
-%\paragraph{Paragraph Heading} %
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references and citations as has already been described in
-%Sect.~\ref{sec:2}.
-%
-%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-%For typesetting numbered lists we recommend to use the \verb|enumerate| environment -- it will automatically render Springer's preferred layout.
-%
-%\begin{enumerate}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\begin{enumerate}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\end{enumerate}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\end{enumerate}
-%
-%
-%\subparagraph{Subparagraph Heading} In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text.
-Use the \LaTeX\ automatism for all your cross-references and citations as has already been described in Sect.~\ref{sec:2}, see also Fig.~\ref{fig:2}.
-%
-%For unnumbered list we recommend to use the \verb|itemize| environment -- it will automatically render Springer's preferred layout.
-%
-%\begin{itemize}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development, cf. Table~\ref{tab:1}.}
-%\begin{itemize}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\end{itemize}
-%\item{Livelihood and survival mobility are oftentimes coutcomes of uneven socioeconomic development.}
-%\end{itemize}
-%
-%\begin{figure}[t] %[t] sets the image at the top of the page; t = top, b = bottom, h = here%
-%\sidecaption[t]
-%% Use the relevant command for your figure-insertion program
-%% to insert the figure file.
-%% For example, with the option graphics use
-%\includegraphics[scale=.65]{figure}
-%%
-%% If no graphics program available, insert a blank space i.e. use
-%%\picplace{5cm}{2cm} % Give the correct figure height and width in cm
-%%
-%%\caption{Please write your figure caption here}
-%\caption{If the width of the figure is less than 7.8 cm use the \texttt{sidecapion} command to flush the caption on the left side of the page.
-If the figure is positioned at the top of the page, align the sidecaption with the top of the figure -- to achieve this you simply need to use the optional argument \texttt{[t]} with the \texttt{sidecaption} command}
-%\label{fig:2}       % Give a unique label
-%\end{figure}
-%
-%% Use the \index{} command to code your index words
-%% Make sure to inlcude the indexed word inside and outside of the brackets if you want the text to show up in your paragraph:
-%% e.g. This book is about \index{genetic programming}genetic programming. 
-%% If the text is not entered outside the brackets it will appear as: This book is about .
-%%
-%% For tables use
-%%
-%\begin{table}
-%\caption{Please write your table caption here}
-%\label{tab:1}       % Give a unique label
-%%
-%% Follow this input for your own table layout
-%%
-%\begin{tabular}{p{2cm}p{2.4cm}p{2cm}p{4.9cm}}
-%\hline\noalign{\smallskip}
-%Classes & Subclass & Length & Action Mechanism  \\
-%\noalign{\smallskip}\svhline\noalign{\smallskip}
-%Translation & mRNA$^a$  & 22 (19--25) & Translation repression, mRNA cleavage\\
-%Translation & mRNA cleavage & 21 & mRNA cleavage\\
-%Translation & mRNA  & 21--22 & mRNA cleavage\\
-%Translation & mRNA  & 24--26 & Histone and DNA Modification\\
-%\noalign{\smallskip}\hline\noalign{\smallskip}
-%\end{tabular}
-%$^a$ Table foot note (with superscript)
-%\end{table}
-%%
-%\section{Section Heading}
-%\label{sec:3}
-%% Always give a unique label
-%% and use \ref{<label>} for cross-references
-%% and \cite{<label>} for bibliographic references
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references and citations as has already been described in
-%Sect.~\ref{sec:2}.
-%
-%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-%If you want to list definitions or the like we recommend to use the Springer-enhanced \verb|description| environment -- it will automatically render Springer's preferred layout.
-%
-%\begin{description}[Type 1]
-%\item[Type 1]{That addresses central themes pertainng to migration, health, and disease.
-In Sect.~\ref{sec:1}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
-%\item[Type 2]{That addresses central themes pertainng to migration, health, and disease.
-In Sect.~\ref{subsec:2}, Wilson discusses the role of human migration in infectious disease distributions and patterns.}
-%\end{description}
-%
-%\subsection{Subsection Heading} %
-%In order to avoid simply listing headings of different levels we recommend to let every heading be followed by at least a short passage of text.
-Use the \LaTeX\ automatism for all your cross-references and citations citations as has already been described in Sect.~\ref{sec:2}.
-%
-%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-%\begin{svgraybox}
-%If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option \verb|graybox| and the newly defined environment \verb|svgraybox|.
-This will produce a 15 percent screened box 'behind' your text.
-%
-%If you want to emphasize complete paragraphs of texts we recommend to use the newly defined Springer class option and environment \verb|svgraybox|.
-This will produce a 15 percent screened box 'behind' your text.
-%\end{svgraybox}
-%
-%
-%\subsubsection{Subsubsection Heading}
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references and citations as has already been described in
-%Sect.~\ref{sec:2}.
-%
-%Please note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%
-%\begin{theorem}
-%Theorem text goes here.
-%\end{theorem}
-%%
-%% or
-%%
-%\begin{definition}
-%Definition text goes here.
-%\end{definition}
-%
-%\begin{proof}
-%%\smartqed
-%Proof text goes here.
-%\qed
-%\end{proof}
-%
-%\paragraph{Paragraph Heading} %
-%Instead of simply listing headings of different levels we recommend to
-%let every heading be followed by at least a short passage of text.
-%Further on please use the \LaTeX\ automatism for all your
-%cross-references and citations as has already been described in
-%Sect.~\ref{sec:2}.
-%
-%Note that the first line of text that follows a heading is not indented, whereas the first lines of all subsequent paragraphs are.
-%%
-%% For built-in environments use
-%%
-%\begin{theorem}
-%Theorem text goes here.
-%\end{theorem}
-%%
-%\begin{definition}
-%Definition text goes here.
-%\end{definition}
-%%
-%\begin{proof}
-%\smartqed
-%Proof text goes here.
-%\qed
-%\end{proof}
-%%
-%\begin{acknowledgement}
-%If you want to include acknowledgments of assistance and the like at the end of an individual chapter please use the \verb|acknowledgement| environment -- it will automatically render Springer's preferred layout.
-%\end{acknowledgement}
-%%
-%\paragraph{Referencing}
-%This is how to reference. \cite{Kadie}
-
 \bibliographystyle{spbasic}

### 2015-03-27 markdown/manuscript.tex
forgot to render

@@
 
-In this contribution we must face the fact that \emph{this response is inevitable}.
-Not just from silly lay ``customers'' in an application project, but from ourselves.
-Whether the project is ``theoretical'' or ``practical''.
-Whether it is ``small'' or ``large''.
+In this contribution we must face the fact that such a response is inevitable: not just from silly lay ``customers'' in an application project, but from ourselves; whether the project is ``theoretical'' or ``practical''; whether it is ``small'' or ``large''.
+Projects resist our initial plans.
 

### 2015-03-27 markdown/manuscript.tex
second pass, part 2

@@
 
-\section{How we treat GP (and how it treats us in turn)}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}
+\section{How we treat GP, and how it treats us in turn}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}
 
-Genetic Programming (and the broader body of work to which it belongs, which I prefer to call ``generative processing'' and will also abbreviate as ``GP'') embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and---though least of all---optimization.
-A case could be made that the resistance we have all recounted towards GP from the technical and lay audience has little to do with the technical results so amply demonstrated over the last quarter-century, but rather from discomfort among that audience in GP's particular ``way of working'' on problems.
-There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are ``rigorous'', or even successful, only when they proceed through ordered phases of
+Genetic Programming\footnote{And not just Genetic Programming as such, but also the broader discipline to which I claim it belongs and which is not obliged to be either ``genetic'' or ``programming''.
+I prefer to call this looser collection of practices ``generative processing'', and will also abbreviate it ``GP''; assume I mean the latter in every case.} embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and optimization.
+I increasingly suspect the resistance we've all recounted towards GP from our prospective technical and lay audience has little to do with our technical results, but rather arises from uncertainty among that audience with GP's very particular ``way of working''.
 
+There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are only ``rigorous'' when they proceed through a sequence of ordered phases from planning to implementation.
+The ``scientific method'' is most often represented something along the lines of
+
+\begin{quote}
 \begin{enumerate}
-\item conceptualization, the ``vision thing''
-\item planning
-\item design
-\item architecture
-\item implementation
-\item testing
-\item debugging
+\item conceptualization; 2. planning; 3. design; 4. architecture; 5. implementation; 6. testing; 7. debugging
 \end{enumerate}
+\end{quote}
 
-Very few scientists or engineers would admit that any \emph{real} project has followed this narrative structure in practice, but in every phase of work life from fund-raising to published reports the story is framed in that way, especially in scientific work: Because of work gone before, an idea or insight was had.
-The idea was framed as a formal hypothesis.
-The hypothesis (and current Best Statistical Practices) suggested an experimental design obvious to any in the discipline.
-The experimental design was undertaken, the data collected, the hypothesis tested, and we can be confident of its veracity because, well, you just heard me say ``Best Statistical Practices'', right?
+I am sure very few scientists or engineers of my acquaintance would admit any \emph{real} project (in history) has ever followed this narrative arc in a literal sense.
+But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports:
 
-Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe any institutional project management and public policy planning process as well.
+\begin{quote}
+Because of the body of published work, an insight was had.
+The insight was framed as a formal hypothesis.
+The hypothesis (and current Best Statistical Practices) suggested an experimental design, one familiar and obvious to any in Our Discipline.
+The experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say ``Best Statistical Practices'', right?
+\end{quote}
+
+Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe almost any institutional project management and public policy planning process as well.
 The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.
@@
 Of course, nobody ``really believes'' this narrative who has ever done any of the work.
-It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and more today.
+It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and many more to be found in the Table of Contents in any Philosophy of Science text.
+
+One in particular is my focus today: Andrew Pickering.
 
@@
 
-Andrew Pickering's \emph{Mangle of Practice} is the one I'd like to use here.
-This approach is noteworthy because it is so close to our actual experience using GP, of the ``didn't we already know this?'' sort.
-At the risk of eliding a lot of extraordinarily well-considered structure, let me summarize:
+Andrew Pickering's \emph{Mangle of Practice} is a decade old, but surprisingly little-known outside the field of Science Studies.
+His approach is especially noteworthy here because I find it so close to our actual experience using GP.
+Indeed, most colleagues who hear it for the first time utter an inevitable ``didn't we already know this?'' I think the distinction Pickering's approach is able to highlight is exactly the problematic one in our work: between the illusory (but publishable) narrative of ``scientific method'', and the realized experience we have of \emph{performing science}.
+At the cost of glossing a lot of his well-considered structure, let me summarize.
 
-The act of ``doing science'' is \emph{at no point whatsoever} a behavior undertaken by an isolated ``researcher'' in an objective field of externalities.
-Rather, the researcher who initiates this process with a vision, intuition, hypothesis or hunch begins always by \emph{making some artifact}: code, an instrument, a proof, a sketch, a maquette or even a thoughtful conversation at a conference.
+First, the act of ``doing science'' is \emph{at no point whatsoever} to be understood as one in which an isolated ``researcher'' works in an objective and static field of ``externalities'' and ``facts''.
+Rather, research begins only when the researcher undertakes to \emph{makes some artifact}: a block of code, a technical instrument, an equation, a sketch, a maquette or even a thoughtful conversation at a conference.
+Everything before the researcher begins to make these things is a matter of building ``vision'' (my term); research proper only occurs when work is done in the real world.
+I'll call this work product \emph{the thing made}, and keep in mind that it is a proxy for the vague collection we might otherwise call ``the project''.
 
-The thing, inevitably, \emph{resists}.
-That is, Pickering grants it agency, or rather makes it the agent of all those externalities that impinge on the work to make the \emph{vision} differ from \emph{practice}: the facts of the actual world, the cultural assumptions and habits of discipline, the raw materials and toolkit available to the practitioner, and so forth.
-``Resistance'' here is not merely the obvious trouble of a software bug or some missing parts, but includes one's sense \emph{on seeing it} that something's not quite right; the realization that more (or less) is needed.
-The language we use, in the face of this resistance, is that the \emph{thing made} ``feels wrong'' or ``points something out'', that it ``wants to do X instead of Y'', or that ``it's doing something too complicated for me to understand right now''.
+We see Pickering's model moving quickly away from more traditional views of science when we realize he has given the inanimate \emph{thing made} an agency of its own.
+In any real project, we perceive the thing made \emph{resisting}---whether we imagine that it resists ``in itself'', or as an agent of externalities that impinge on the work in progress to make our driving vision differ from reality.
+The thing made comes to represent, in other words, the facts of the actual world, the cultural assumptions and norms of one's discipline, the raw materials and toolkit available to a practitioner, and so forth.
+``Resistance'' here is not merely a problem with a software bug or lost raw materials, but includes one's sense \emph{on seeing it} that something's not quite right.
+The thing made's resistance, as perceived by the researcher, leads her to react in turn.
 
-And---assuming science is indeed what's being done---the researcher \emph{changes in response to this resistance}.
+Consider the language we use in the face of this resistance: it ``feels wrong''; it ``points something out''; it ``wants to do X instead of Y''; it's ``doing something too complicated for me to understand right now''.
+
+And---assuming science, engineering, art or any other creative process is indeed what's being done---the researcher \emph{changes in response to this resistance}.
 That vision changes, the plan adapts, or in some other way the \emph{thing made} causes a response in the state of the researcher herself.
@@
 
-The ``mangling'' of Pickering's metaphoric name refers not to wounding but to the mechanical laundry apparatus, the antique wringer through which sheets and linens are run to drain the water, and which as a \emph{side-effect} impose a novel higher-dimensional structure and juxtapose unexpected components with one another.
-Here, the response of the researcher to resistance she experiences in the course of her project is no less a part of the dynamics of ``science'' than the act of writing and running code, the authority warranted by particular statistical practices in her discipline, or the raw pressure of physical laws.
+Pickering's ``mangle''\footnote{The word ``mangle'' he has chosen is itself interesting and insightful:
+
+\begin{quote}
+\ldots{}I find ``mangle'' a convenient and suggestive shorthand for the dialectic because, for me, it conjures up the image of the unpredictable transformations worked upon whatever gets fed into the old-fashioned device of the same name used to squeeze the water out of the washing.
+It draws attention to the emergently intertwined delineation and reconfiguration of machinic captures and human intentions, practices, and so on.
+The word ``mangle'' can also be used appropriately in other ways, for instance as a verb.
+Thus I say that the contours of material and social agency are mangled in practice, meaning emergently transformed and delineated in the dialectic of resistance and accommodation\ldots{}.
+\end{quote}} is this emergent dance of agency, undertaken between the researcher and the project: the researcher makes, the thing made resists, the researcher is influenced and redirected, \emph{accommodating} the resistance.
+It may seem glib to say that ``no plan survives contact with the enemy,'' but Pickering's narrative of creative work emphasizes the fact that no \emph{revised} plan survives unchanged, either.
+In the traditional narrative, we elide the work as it unfolded and re-frame it as a sort of idealized, apersonal Platonic truth: we use the passive voice, we hide the missteps and confusion, we paint a story moving from vision to plan to success.
+
+In a GP setting, the many modes of resistance we perceive are the very ``symptoms'' Rick mused about.
+Even though we as researchers know we've written all the code and set all the parameters, we're nonetheless willing to speak of a GP run ``doing'' things, as opposed to merely unfolding according to our plan.
+A GP system does not ``resist'' by, for example, ``having the wrong population size''.
+Rather it resists by \emph{causing concern or dissatisfaction in the user}, which in turn sparks in that user a practical (if only explanatory) response, which leads them to change the \emph{thing made}.
 
@@
 
-The broader field of machine learning leans seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions.
+The broader field of machine learning seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions.
 The result of training a neural network or random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the sufficiently robust result of applying numerical optimization to a particular mathematical programming model.

### 2015-03-28 markdown/manuscript.tex
second pass, third section

@@
 We see Pickering's model moving quickly away from more traditional views of science when we realize he has given the inanimate \emph{thing made} an agency of its own.
-In any real project, we perceive the thing made \emph{resisting}---whether we imagine that it resists ``in itself'', or as an agent of externalities that impinge on the work in progress to make our driving vision differ from reality.
-The thing made comes to represent, in other words, the facts of the actual world, the cultural assumptions and norms of one's discipline, the raw materials and toolkit available to a practitioner, and so forth.
+In any real project, we perceive the \emph{thing made} resisting---whether we imagine that it resists ``in itself'', or as an agent of externalities that impinge on the work in progress to make our driving vision differ from reality.
+The \emph{thing made} comes to represent, in other words, the facts of the actual world, the cultural assumptions and norms of one's discipline, the raw materials and toolkit available to a practitioner, and so forth.
 ``Resistance'' here is not merely a problem with a software bug or lost raw materials, but includes one's sense \emph{on seeing it} that something's not quite right.
-The thing made's resistance, as perceived by the researcher, leads her to react in turn.
+The \emph{thing made}'s resistance, as perceived by the researcher, leads her to react in turn.
 
@@
 Thus I say that the contours of material and social agency are mangled in practice, meaning emergently transformed and delineated in the dialectic of resistance and accommodation\ldots{}.
-\end{quote}} is this emergent dance of agency, undertaken between the researcher and the project: the researcher makes, the thing made resists, the researcher is influenced and redirected, \emph{accommodating} the resistance.
+\end{quote}} is this emergent dance of agency, undertaken between the researcher and the project: the researcher makes, the \emph{thing made} resists, the researcher is influenced and redirected, \emph{accommodating} the resistance.
 It may seem glib to say that ``no plan survives contact with the enemy,'' but Pickering's narrative of creative work emphasizes the fact that no \emph{revised} plan survives unchanged, either.
@@
 The broader field of machine learning seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions.
-The result of training a neural network or random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the sufficiently robust result of applying numerical optimization to a particular mathematical programming model.
-Whether one describes the process as minimizing out-of-sample error or maximizing information gain, the focus of the discipline is on contingent reliability rather than exploratory modeling.
+The result of training a neural network or even a random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the reliable and robust end-product of applying numerical optimization to a well-specified mathematical programming problem.
+Whether one describes these machine learning processes as ``minimizing out-of-sample error'' or ``maximizing information gain'', the supposed strength of most machine learning approaches is the \emph{unsurprising} nature of their use cases and outputs.
 
-GP has the capacity to \emph{tell us stories}, even in the ``simple'' domain of symbolic regression problems.
-The space under consideration is not merely a vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, and of functions over inputs, and of higher-order functions over those.
-While GP can be used to explore arbitrarily close to a paradigmatic model, we argue for it most when its application can produce unexpected insights.
-A number of us treat it as the best candidate for ``real'' artificial intelligence, and rightly so.
+On the other hand, GP has the capacity to \emph{tell us stories}---even in the relatively ``simple'' domain of symbolic regression.
+The space under consideration by GP is not merely a vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, functions over inputs, and higher-order functions over those.
+We who work in the field can be glib about the ``open-endedness'' of GP systems, but that open-endedness puts them at odds with their supposed relatives in machine learning.
+While GP \emph{can} be used to explore arbitrarily close to some paradigmatic model, its more typical use case leads to the production of \emph{unexpected} insights---to the degree that a number of us feel justified in treating it as a strong candidate for ``real'' artificial intelligence.
 
-We can do that because GP surfaces Pickering's Mangle.
-When it ``works'', it does so by offering \emph{helpful resistance} in our engagement with it, whether in the form of surprising answers, validation of our suspicions, or suggestions of ways to make subsequent moves.
-It \emph{dances} with us, in a way that the other mere tools of machine learning do not.
+I argue we have that leeway because of the way GP surfaces Pickering's Mangle.
+When the methodology ``works'', it does so by offering \emph{helpful resistance} in our engagement with the problem at hand, whether in the form of surprising answers, or validation of our suspicions, or simply legible suggestions of ways to make subsequent moves.
+GP \emph{dances} with us, while most other machine learning methods are ``mere tools''.
 
-Now a great deal of the last quarter-decade of work in the field, especially in the early days, seems to treat GP as a close relative of other machine learning techniques---as a methodology for producing \emph{unsurprising} models from data, and within the traditional model of scientific work.
-That is, an implicit and idealized user is expected to proceed something like the users of any other machine learning framework:
+\subsection{Against replication}\hypertarget{against-replication}{}\label{against-replication}
+
+Nonetheless, there seems to be a widespread desire, inside and outside our field, to frame GP as a methodology for producing \emph{unsurprising} models from data, more in keeping with the traditional linear of scientific work.
+That is, an idealized user is expected to proceed something like the users of any other machine learning or mathematical programming framework:
 
 \begin{enumerate}
-\item frame your problem in the framework-specific language
+\item frame your problem in the correct formal language
 \item ``get'' a GP system
-\item ``run GP on the data''
-\item ???
+\item run GP ``on your data''
+\item (whatever this is, it's not our problem)
 \item you have solved your problem
@@
 
-There's strong pressure from our actual user community to enforce this stance, not least because it is exactly the stance of planning and public policy, and of the mythic science or programming project manager.
-That is, it frames GP as a \emph{tool} to be invoked in a known and well-described planning situation.
+This is not original with the GP community; there is strong pressure from our peers in other disciplines and our users to promote this use case, not least because it is \emph{exactly} the stance expected in any planning or public policy setting, or in any scientific or programming project management setting.
+That is, we are under tremendous social pressure to treat GP as a \emph{tool} to be invoked in a known, predictable and well-described planning situation.
 
-The resulting resistance from early-adopting ``users'' who ``get'' an off-the-shelf GP system and try to apply it to their problem---at least within this narrative of work---should not be unexpected.
-\emph{Being surprising} is perhaps the worst choice for any traditional project plan.
-It's no wonder that so much of GP research is focused on particular tweaks and staged horse-races among the unnumbered techniques: as long as there is a sentiment that GP might be ``tamed'' so that step (4) above \emph{never happens}, so that the framework might join its more traditional and popular relatives, there will be strong pressure to calculate statistics on, for example ``proportion of successes in 30 replicates''.
+The resulting resistance from early-adopting users shouldn't be unexpected.
+\emph{Being surprising} may well be the worst conceivable behavior for any tool to be used in a traditional project setting.
+And given that pressure, it's no wonder that so much of GP research is focused on constraining tweaks to bring GP ``into line''.
+If only GP could be ``tamed'' or made ``adaptive'' so that step (4) above \emph{never happens}.
 
-But what does a ``replicate'' mean for somebody using GP for a real project, whether theoretical or practical, where \emph{GP itself} is not the focus of the work?
-Projects which authentically ``use'' GP must necessarily be searching for noteworthy answers, which is to say \emph{surprising} and \emph{interesting} answers, that they cannot otherwise obtain.
-To do otherwise is to somehow expect GP to be a form of ``research automation'', a premature expectation in light of our experience.
+I imagine this is why so many GP research projects strive for rigor in the form of counting replicates which ``find the solution'': they aim not to convince users of the known strengths of GP, but rather demonstrate to critical peers that GP can be ``tamed'' into a mere tool.
 
-Therefore, any researcher who is using GP \emph{realistically} is one who is watching, and adjusting, and engaging and interacting with the process of search itself.
-Seen as working within a traditional (and deeply misleading) narrative framework, we watch as she runs a population of 100 individuals for 100 generations, peers at the results in a CSV file and finds them wanting, and then adjusts the GP parameters to run another 100 generations\ldots{} and repeating as necessary.
-But there is \emph{no discernible difference} when we frame the same process as one of participating in the single application of an interactive algorithm that (remarkably, and unaccountably when you think about it) \emph{throws away all intermediate results every 100 generations}.
+What would a ``replicate'' stand for, to a user who sought to exploit GP's strength in a project (whether theoretical or practical) where \emph{search} rather than the algorithm itself is the primary focus?
+Projects which authentically ``use'' GP \emph{must necessarily} be those searching for noteworthy answers---which is to say \emph{surprising} and \emph{interesting} answers---that they could not otherwise obtain.
+Thus, we should better think of a ``replicate'' as a sort of proxy for user frustration in step (4) above: that is, it represents a project in which search begins, stalls, and for which the user cannot see a way to move search forward towards more interesting and useful answers.
 
-Further, let's explore that traditional narrative which sees her workflow as a series of repeated attempts to use GP as a tool in a situation for which it is poorly suited, or maybe ``badly tuned''.
-She is carefully ``not interfering'' with any given run of 100 generations, observing our tradition in this field, and can only peer into the results file after the fact.
+But I think we would agree: it is a poor researcher who, when faced with a stumbling block in the form of a black box's misbehavior, doesn't attempt to work around that obstacle.
+Not to \emph{begin the project again from scratch}, but to make changes and continue.
+Any researcher who is using GP \emph{realistically} will be watching, and adjusting, and engaging and interacting with the process of search itself.
+
+When a GP user is viewed as working within the traditional \emph{linear} narrative of research, we see her run a population of 100 individuals for 100 generations, peer at the results that have been dumped into a CSV file, and find them wanting.
+Desiring an actual \emph{answer}, she adjusts the GP parameters and begins another run of 100 generations\ldots{} and repeats as necessary.
+
+But there is \emph{no discernible difference} when we frame this same process of ``many runs'' as a single project involving initial researcher moves, resistances thrown up by the \emph{thing made}, and subsequent accommodations made by the researcher to the new perceived truths.
+Except, that is, in the researcher's view of her own response to resistance: if she comes to the project with \emph{plans} for running ``ten replicates'', we can only assume she has learned from someone that a search algorithm is \emph{supposed} to forget everything it has learned every 100 generations\ldots{}.
+
+I cannot help but be reminded of the fallacy, surprisingly common both in and outside of our field, that ``artificial intelligence'' must somehow be a self-contained and non-interactive process.
+That is, that an ``AI candidate'' loses all authenticity as soon as it is ``tweaked'' or ``adjusted'' in the course of operation.
+It is as if every new-born ``AI'' must be jammed into an air-tight computational container and isolated \emph{until it learns to reason}, and that without exceeding a finite computational budget.
+If humans creating \emph{real} intelligences treated them anything like the way computer scientists insist we treat nascent \emph{artificial} ones, murder charges would be forthcoming.
+
+There are several practical reasons for us to try something different.
+
+Consider our hypothetical GP user.
+In keeping with the Behaviorist standards of GP, she is carefully ``not interfering'' with any given run of 100 generations; she can only peer at a results file after the fact.
 During the course of any 100 generations, all sorts of dynamics have happened: crossover, mutation, selection, all the many random choices.
@@
 
-It is, I think, because the myths of artificial intelligence and vision-driven science are deeply intertwined.
-It is somehow ``cheating'' to admit in a scientific paper, even if no mistakes were made, that the original vision and plan changed over the course of the project; rather we describe research \emph{results} as inevitable outcomes of an ahistorical process that erases the work actually done by human beings.
-Similarly, it is somehow ``cheating'' to admit in a GP project, even if every parameter was set correctly, that the original vision and plan gave way to the inevitable surprises thrown up by GP's tremendous potential to surprise.
+It is, I think, because the myths of artificial intelligence and the linear narrative of science are so deeply intertwined.
+It is frowned upon to admit in a scientific paper, even when no mistakes were made, that the original vision and plan changed over the course of the project; rather we are expected to describe research \emph{results} as the inevitable outcomes of an ahistorical process, and erase all resistance and accommodation actually done by human beings in the context of their projects.
+Similarly, it feels somehow wrong to admit in a GP project, even if every parameter was set correctly, that the original vision and plan gave way to the inevitable surprises thrown up by GP's inherent tendencies to do just that.
 
-But insofar as GP \emph{surprises us}---and that is its sole strength over more predictable and manageable frameworks---we will inevitably see a good fraction of the surprises as \emph{disappointments} rather than opportunities to adapt our selves.
+But insofar as GP \emph{surprises us}---and that is its sole strength over more predictable and manageable frameworks---we will inevitably see a good fraction of those surprises as \emph{disappointments} rather than opportunities to adapt ourselves.
+I'd like to suggest we learn new ways to \emph{accommodate} those disappointments, and stop trying to make them go away.
 

### 2015-03-28 markdown/manuscript.tex
fixed a typo

@@
 
-But insofar as GP \emph{surprises us}---and that is its sole strength over more predictable and manageable frameworks---we will inevitably see a good fraction of those surprises as \emph{disappointments} rather than opportunities to adapt ourselves.
-I'd like to suggest we learn new ways to \emph{accommodate} those disappointments, and stop trying to make them go away.
+But insofar as GP \emph{surprises us}, and since that is its sole strength over more predictable and manageable frameworks, we must inevitably see a good fraction of those surprises at least in part as \emph{disappointments} rather than encouraging opportunities to change our plans.
+Let's learn new ways to accommodate those disappointments, and stop trying to make them go away.
 
@@
 \begin{enumerate}
-\item Add a little test
-\item Run all tests and fail
-\item Make a little change
-\item Run the tests and succeed
-\item Refactor to remove duplication
+\item Add a little (failing) test which exercises the next behavior you want to build into your codebase
+\item Run all tests, expecting only the newest to fail
+\item Make the minimal change to your codebase that permits the new test to pass
+\item Run all tests, expecting them all to succeed
+\item Refactor codebase to remove duplication
 \end{enumerate}
 
-Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the overarching flow of implementation (or ``design'') that it imposes: Each cycle, and also the overarching design process, begins with the choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced over all iterations of this cycle.
+Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the iterative flow of implementation (or ``design'') that it imposes: Each cycle begins with a choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced so far.
 The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working through them as if they could be automated easily.
@@
 Rather than producing a suite of tests and a self-contained codebase, they are forced to use \emph{only} refactoring of code added to tests to produce their eventual ``codebase''.
-In other words, no code can be ``produced'' until \emph{duplication in code added multiple passing tests} provides a warrant for refactoring it out.
+In other words, no code can be ``produced'' until the ``smell'' of duplication in the code added to multiple passing tests \emph{provides a warrant} for refactoring it out.
+
 Further, a facilitator patrols ongoing work and deletes \emph{any and all code not called for by a pre-existing failing test}.
-
-Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens\ldots{} as one might imagine.
-But as Gojko Adzik emphasizes in his descriptions of workshops he's run, the results of ``design'' of even simple algorithms in this artificial amplified setting seems much more \emph{open-ended} than it would if the software were built within the typical framing of habits and assumptions that an experienced programmer carries along with her to any project.
+Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine.
+But as Gojko Adzik emphasizes in his descriptions of workshops, the resulting designs for even simple algorithms in this artificially amplified setting seems much more \emph{open-ended} than it would if the software were built under the typical norms and habits an experienced programmer uses in normal conditions.
 
@@
 
-\section{As if we \emph{really} meant it}\hypertarget{as-if-we-really-meant-it}{}\label{as-if-we-really-meant-it}
+\section{GP as if we meant it}\hypertarget{gp-as-if-we-meant-it}{}\label{gp-as-if-we-meant-it}
 
@@
 
-\subsubsection{``as if we really meant it''}\hypertarget{as-if-we-really-meant-it-1}{}\label{as-if-we-really-meant-it-1}
+\subsubsection{``as if we really meant it''}\hypertarget{as-if-we-really-meant-it}{}\label{as-if-we-really-meant-it}
 

### 2015-03-28 markdown/manuscript.tex
session.md appendix using definition list (becomes description list in LaTeX)

@@
 In the same way that Braithwaite's onerous coding exercise is intended to drive the attention of its participants toward test-driven design with its obligation to write ``real'' code \emph{only as a refactoring}, I'd like to be able to demand a \emph{warrant} for every step that moves our changing genetic programming setup away from just plain random guessing.
-Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed}.\footnote{In much the same way that Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong'', I imagine many GP users might say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''.
-I offer no particular justification for either anecdotal stance here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}
+Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed}.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''.
+I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''.
+I offer no particular justification for either anecdote here, but the curious reader is encouraged to poll a sample of participants at any conference (agile or GP).}
 
@@
 
-\subsection{The tableau representation of GP systems}\hypertarget{the-tableau-representation-of-gp-systems}{}\label{the-tableau-representation-of-gp-systems}
+\subsection{A tableau representation}\hypertarget{a-tableau-representation}{}\label{a-tableau-representation}
 
-I find it helps to present a simplified but formalized description of GP systems, and one which highlights the particular features I'm considering.
+The exercise proceeds as a sort of game, in which the User and the System take alternating turns.
+During the User's turn, she can see state of the system so far and make any of several \emph{moves} from a limited set, each of which involve changing the settings of a \emph{tableau}, which completely describes the state of the search system.
+During the System's turn, it will execute a finite number of steps in which it creates new individuals.
+At the end of the System's turn, control reverts to the User, and vice versa.
 
-At the highest level of abstraction, we will treat GP as a collection of particular \emph{decisions} made by the user, plus an otherwise autonomous stochastic process executed by software and hardware, which can be ``started'', ``paused'' and ``resumed'' but which cannot be \emph{restarted}.
-
-The decisions available to the user apply to three core components of the stochastic process: {\tt operators}, {\tt answers}, and {\tt rubrics}.
+The tableau, and the suite of moves available to the User, affect three core aspects of search: {\tt operators}, {\tt answers}, and {\tt rubrics}.
 
@@
 An {\tt operator} is any function which takes as argument a (possibly empty) collection of {\tt answers} and produces a new collection of {\tt answers}.
-Operators thus include ``random guessing'', which in GP systems is often used to build an initial population, any ``crossover'' and ``mutation'' functions, but also any arbitrarily complex function which might be used to \emph{create a new answer}.
-So for example ``particle swarm on an abstract expression tree'' would be an {\tt operator} working on a single {\tt answer} and producing a very large collection of results.
+Operators thus include ``random guessing'', which in GP systems is often used to build an initial population, any ``crossover'' and ``mutation'' functions.
+No {\tt operator} can refer to any {\tt answer} not part of its argument set, not can any unset values be set within an {\tt operator}.
 
-In the exercise (but not in the examples that follow) \emph{every intermediate result} which is an {\tt answer} is considered to be part of the return value.
-That is, {\tt operators} are obliged to return all intermediate {\tt answers} they produce while building their ``actual'' results: a function that implements ``crossover-and-mutation'' will is obliged to return \emph{all} crossover products it builds, in addition to the results of a subsequent internal mutation.
+So for example, if the User wished an {\tt operator} could be constructed which took 30 {\tt answers} as inputs, read their attributes (including {\tt script} and that subset of their {\tt rubric} values which were already set), and produce a single {\tt answer} in its return set.
+It could \emph{not} take an {\tt answer}, change it, and keep the new one ``if it was better'', since the new {\tt answer} would have no scores at all.
 
-Note that no process is provided for {\tt answers} to be \emph{removed} from a tableau. This framework is purely cumulative.
+Note that no process is provided for {\tt answers} to be \emph{removed} from a tableau. The framework is purely cumulative.
+
+All ``parents'' and other {\tt answer} inputs required by an {\tt operator} are chosen by \emph{lexicase selection} automatically, using the complete suite of {\tt rubrics} in play when invoked.
+Lexicase selection samples each {\tt rubric} with equal probability.
 
@@
 
-An {\tt answer} is what might traditionally be called an ``individual'' in GP literature, though there are subtle differences.
-We can model them programmatically as key--value hash, typically beginning with a ``genome'' or ``script'' field set to an abstracted representation for a solution to the problem at hand.
-As an {\tt answer} ``matures'' in the unfolding tableau, various  other attributes will be appended and set by other algorithmic processes, such as fitness scores or measured attributes like ``age'' or ``alive?'' states.
+An {\tt answer} might traditionally be called an ``individual'' in GP literature.
+We can model them programmatically as key--value hash.
+All new {\tt answers} are born with only a unique {\tt id} and {\tt script} field set.
+As an {\tt answer} ``matures'' in the unfolding tableau, various other attributes will be set by the System player through the application of {\tt rubrics}.
 
-In the tableau layout, we will represent the unfolding collection of {\tt answers} as the \emph{rows} of a spreadsheet-like table, and their attributes and scores as the \emph{columns}.
+In the tableau visualization, we represent the unfolding collection of {\tt answers} as the \emph{rows} of a spreadsheet-like table, and their attributes and scores as the \emph{columns}.
 
@@
 
-A {\tt rubric} is any function which returns a scalar numerical value, given arguments of a collection of one or more {\tt answers} (and possibly additional arguments), and assigned to a particular {\tt answer}.
+A {\tt rubric} is any function which returns a scalar numerical value, given the instantaneous state of the tableau itself.
+For the most part these values can be understood as ``scores'' for various search objectives, though it is likely that a number of ``implicit objectives'' will also create new {\tt rubrics}.
 
-Insofar as any given {\tt rubric} is associated with an objective of search, it should be framed as a \emph{minimization} form.
-For errors, this should be obvious; for {\tt rubrics} used in ALPS-like systems, realize that the desired {\tt rubric} to select \emph{younger} {\tt answers} is not ``age'' but ``youth''.
+A {\tt rubric} cannot store intermediate values, but can refer to the values of other {\tt rubric} columns.
+Thus on specifying a {\tt rubric} like ``sum squared error over a training set'', one must also create  {\tt rubrics} for ``measured error when provided input $i$'' for every element $i$ of the training set.
+If the training set has 100 elements, then the single SSE {\tt rubric} \emph{implicitly} represents a suite of 101 new columns added to the tableau.
 
-In our final exercise (though not in the examples to follow) there will also be a strict requirement that any attribute associated with an {\tt answer} is also available as an \emph{implicit} {\tt rubric}: the script, the creation time (if recorded), or any other {\tt rubric}.
-We will not permit {\tt rubrics} to store intermediate values, \emph{except in other rubrics}: thus a {\tt rubric} which specifies ``sum squared error over a training set'' cannot be applied without also \emph{first} creating {\tt rubrics} for ``measured error when provided input $i$'' every element $i$ of the training set.
-If the training set has 100 elements, then the SSE {\tt rubric} implicitly represents a suite of 101.
+A {\tt rubric} function does however have access to the state of the entire tableau as needed, including ``forcing'' other {\tt rubric} values to be calculated.
+So it is entirely possible to specify {\tt rubric} functions which score:
 
-However, in the examples to follow that sketch ``traditional GP'', we can relax this restriction.
+\begin{itemize}
+\item number of characters in the {\tt answer}'s script
+\item maximum error measured in any of 35 other {\tt rubrics}
+\item number of {\tt div0} errors produced when running with a particular set of inputs
+\item number of stochastic instructions appearing in the {\tt answer}'s script, compared to \emph{all other {\tt answers}}
+\end{itemize}
 
-\subsubsection{Search process}\hypertarget{search-process}{}\label{search-process}
+\subsubsection{User moves}\hypertarget{user-moves}{}\label{user-moves}
 
-\subsubsection{``traditional GP'' tableau}\hypertarget{traditional-gp-tableau}{}\label{traditional-gp-tableau}
+\begin{itemize}
+\item add one {\tt operator}
+\item add one {\tt rubric}, which may create other \emph{entailed} {\tt rubrics} as described above
+\end{itemize}
 
-It's tempting to be glib about defining ``traditional GP'', and for the sake of brevity let me succumb to that temptation: Say it is a fixed-size population of 100 {\tt answers} which are created initially at random and subsequently by crossover and mutation, a single {\tt rubric} which returns a single SSE score calculated over input--output pairs measured over a static collection of training cases.
+\subsubsection{System turn}\hypertarget{system-turn}{}\label{system-turn}
 
-\begin{longtable}{|l|l|l|l|}
-\hline
-individual & genome & SSE & (notes)\\
-0.1 & \ldots{} & 0.2 & \\
-0.2 & \ldots{} & 0.8 & \\
-\ldots{} &  &  & \\
-0.N & \ldots{} & 0.1 & \\
-\hline
-\end{longtable}
+During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau. It follows a single rote cycle to do so:
 
-The analogy to SQL.
+\begin{enumerate}
+\item select an {\tt operator} from those in play, with equal probability
+\item apply \emph{lexicase selection} to the tableau to select the required number of input {\tt answers}, filling in missing {\tt rubric} scores as needed
+\item apply that one {\tt operator} to the inputs to produce new {\tt answers}, and append those new {\tt answers} immediately to the tableau
+\item {\tt HALT} if the number of new {\tt answers} added in this turn meets or exceeds the number present when the turn started (in other words, if the number of {\tt answers} has doubled); otherwise, go to step (1)
+\end{enumerate}
 
-\subsubsection{``lexicase selection tableau''}\hypertarget{lexicase-selection-tableau}{}\label{lexicase-selection-tableau}
+\subsubsection{Initial setup and restrictions}\hypertarget{initial-setup-and-restrictions}{}\label{initial-setup-and-restrictions}
 
-The analogy to SQL.
+\begin{itemize}
+\item the only {\tt operator} is ``random guess'', which creates a new {\tt answer} with an arbitrary script
+\item no {\tt rubrics} exist (only the {\tt script} and {\tt id} attributes of the {\tt answers})
+\item 50 new {\tt answers} will be produced in the first System turn
+\item No {\tt operator} can \emph{evaluate} any {\tt answer} except in a self-contained local namespace which lacks access to any tableau values or state information
+\item There is no mechanism for \emph{removing} {\tt answers} from the tableau
+\item The System player \emph{always} uses lexicase selection, and \emph{always} uses all {\tt rubrics} as the selection features with equal probability.
+\item A {\tt rubric} can only \emph{run} a single {\tt answer} once; stochastic scripts will only be sampled one time, and no {\tt rubric} score is ever recalculated after the first time
+\end{itemize}
 
-\subsubsection{``lazy lexicase selection tableau''}\hypertarget{lazy-lexicase-selection-tableau}{}\label{lazy-lexicase-selection-tableau}
+\subsubsection{Interface, the \emph{Goal} of the Exercise}\hypertarget{interface-the-goal-of-the-exercise}{}\label{interface-the-goal-of-the-exercise}
 
-The analogy to NoSQL.
+During the User's turn, they can of course interrogate the tableau in any way they want, without changing it.
+The point of the exercise is to drive the User to explore and create analytics and visualizations which can better inform their decisions over the course of the game.
 
-\subsubsection{``mutually lazy tableau''}\hypertarget{mutually-lazy-tableau}{}\label{mutually-lazy-tableau}
-
-``Send me a message.''
-
-\subsubsection{``as if we really meant it''}\hypertarget{as-if-we-really-meant-it}{}\label{as-if-we-really-meant-it}
-
-I need your help.
-
-\section{Leveraging ``resistance'': the problem of the dots and lines}\hypertarget{leveraging-resistance-the-problem-of-the-dots-and-lines}{}\label{leveraging-resistance-the-problem-of-the-dots-and-lines}
-
-\subsection{dots and lines, in pushforth}\hypertarget{dots-and-lines-in-pushforth}{}\label{dots-and-lines-in-pushforth}
-
-\subsection{the problem of the unknown language}\hypertarget{the-problem-of-the-unknown-language}{}\label{the-problem-of-the-unknown-language}
-
-\subsection{the problem of the objective(s)}\hypertarget{the-problem-of-the-objectives}{}\label{the-problem-of-the-objectives}
-
-\subsection{expanding the goals when it's boring}\hypertarget{expanding-the-goals-when-its-boring}{}\label{expanding-the-goals-when-its-boring}
+\section{An example session}\hypertarget{an-example-session}{}\label{an-example-session}
 
@@
 
-(It would mean the sort of self-awareness it takes to notice that something is wrong, and to ask for help.)
+The idea of this exercise came from observing the differences between experienced and ``novice'' (though with strong technical skills) GP users as they explored new problems in a GP setting.

### 2015-03-28 markdown/session.tex
session.md appendix using definition list (becomes description list in LaTeX)

\section{An example session}\hypertarget{an-example-session}{}\label{an-example-session}

(test placeholder)

\begin{description}
\item[{\tt me}] I do and say a thing here

This is a continuation of the first item, and should include a few extra fillips, like $x=y$ and a little

\begin{itemize}
\item list
\item of
\item items
\end{itemize}

And also it includes a

\begin{quote}
blockquote right here
\end{quote}

\item[{\tt system}] it does and says something back to me

\item[{\tt me}] he look I can have a little dialog sortof

\item[{\tt system}] I can fight you every step of the way, though!
\end{description}

### 2015-03-28 markdown/manuscript.tex
scenario placeholder'

@@
+% kramdown -o latex markdown/session.md > markdown/session.tex
 % kramdown -o latex markdown/manuscript.md > markdown/manuscript.tex
@@
 
+If we think of the tableau as a ``spreadsheet'' with a small list of {\tt operators} and a large sheet of {\tt answers} as rows and {\tt rubrics} as columns, the User can add one new {\tt operator} or one new column.
+
 \begin{itemize}
-\item add one {\tt operator}
-\item add one {\tt rubric}, which may create other \emph{entailed} {\tt rubrics} as described above
+\item add (or activate from a pre-existing list) exactly one {\tt operator}
+\item add exactly one {\tt rubric}, which may in turn create other \emph{entailed} {\tt rubrics} as described above
 \end{itemize}
@@
 
-During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau. It follows a single rote cycle to do so:
+During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau.
+If we think of the tableau as a ``spreadsheet'' the System can only add new {\tt answers}.
+It follows a single rote cycle to do so:
 
@@
 
-\subsubsection{Interface, the \emph{Goal} of the Exercise}\hypertarget{interface-the-goal-of-the-exercise}{}\label{interface-the-goal-of-the-exercise}
+\subsubsection{Interface: the Goal of the Exercise}\hypertarget{interface-the-goal-of-the-exercise}{}\label{interface-the-goal-of-the-exercise}
 

### 2015-03-28 markdown/session.tex
scenario placeholder'

@@
 
-(test placeholder)
+\textbf{Placeholder}: I am converting the transcript of the session from an awful multipage LaTeX {\tt \textbackslash{}longtable} into a much more manageable {\tt \textbackslash{}enum} list like this.
+It will take a while, since the original {\tt \textbackslash{}longtable} was more than nine pages long and full of table markup.
+
+\subsection{The stage is set}\hypertarget{the-stage-is-set}{}\label{the-stage-is-set}
 
 \begin{description}
-\item[{\tt me}] I do and say a thing here
-
-This is a continuation of the first item, and should include a few extra fillips, like $x=y$ and a little
-
-\begin{itemize}
-\item list
-\item of
-\item items
-\end{itemize}
-
-And also it includes a
-
-\begin{quote}
-blockquote right here
-\end{quote}
+\item[{\tt me}] well what have we here?
 
@@
 
-\item[{\tt me}] he look I can have a little dialog sortof
+\item[{\tt me}] hey look I can have a little dialog sortof
 
@@
 \end{description}
+
+\subsection{Progress and Stagnation}\hypertarget{progress-and-stagnation}{}\label{progress-and-stagnation}
+
+\begin{description}
+\item[{\tt me}] Keep going
+
+\item[{\tt system}] OK!
+\end{description}
+
+\subsection{Innovation}\hypertarget{innovation}{}\label{innovation}
+
+\begin{description}
+\item[{\tt me}] Hey!
+
+\item[{\tt system}] What?
+\end{description}
+
+\subsection{Frustrations}\hypertarget{frustrations}{}\label{frustrations}
+
+(\&c)

### 2015-03-29 markdown/manuscript.tex
third pass, intro

@@
 
-More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved evolutionary algorithms and genetic programming.
-As we were chatting one day, Bill asked Rick what he'd most like to see as part of the research program of GP ``moving forward''.
+More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved genetic programming.
+As we chatted one day, Rick was asked what he'd most like to see as part of the research program of GP ``moving forward''.
 Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.
 
-As I recall, Rick answered that he'd like to better understand the ``symptoms'' we often see when we run an evolutionary search process: premature convergence, failure to improve, catastrophic lack of diversity, and the sense when we look at results or ongoing runs that some particular choices we've made setting the many parameters \emph{just aren't quite right}.
-There are numerous well-written papers suggesting ways around local minima, proposing reasonable and exotic search operators, and running ``horse races'' between variant algorithms on benchmark problems.
-But years later, I take Rick's challenge not only to mean that it might be useful to have a benchmarking catalog which lists conditions in which search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood, to the point that we think of them as ``subjective''.
+As I recall, Rick said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, failure to improve, catastrophic lack of diversity, mysterious things like that sense we all get when we look at results that suggests \emph{we made bad choices} of some sort.
 
-Somewhere during that same project, I remember a Project Manager telling the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were confused.
-``We don't want a choice, we want \emph{the best}.'' This came after months of analysis, interviews, and a collective agreement that the problem was fundamentally multi-objective.
-But when the decision-makers were faced with the unquestionably successful results they had collectively specified, \emph{those results weren't right enough}.
+The literature abounds with well-written papers describing tips for avoiding local minima, improving on common search operators, and describing ``horse races'' between Bad Old and Better New search methodologies applied to benchmark problems.
+But I take Rick's challenge not only to mean that it would be useful to have a catalog which lists situations where search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood.
 
-In this contribution we must face the fact that such a response is inevitable: not just from silly lay ``customers'' in an application project, but from ourselves; whether the project is ``theoretical'' or ``practical''; whether it is ``small'' or ``large''.
-Projects resist our initial plans.
+This contribution aims squarely at that sensibility which sees the ``symptoms'' and ``pathologies'' as something we as a research community should be trying to stamp out \emph{before} letting users run their own GP systems.
+Instead, I'll argue that because GP differs qualitatively and philosophically from most other machine learning approaches, and also from the broader lay understanding of the ``automated design'' and exploratory research, much of the value GP offers is lost or ignored when we approach it as something to be ``cleaned up''.
 
-The ``field'' of GP has grown quite a bit in the intervening 15 years since these two anecdotes.
-All of us who follow it closely have played a role in the expanding front of new methodologies and techniques---the extraordinary ones we describe in our little workshop, and also the newsworthy and admirable public successes.
+\subsection{Challenges}\hypertarget{challenges}{}\label{challenges}
 
-But this acceleration of practical successes has been accompanied by a dilution of the scope of the theoretical warrants we use to justify them: Most examples of ``practical success'' involve unique domain-specific quirks, each implementation makes numerous contingent (and sometimes arbitrary) design and architectural choices, and even the ontological foundations of ``individual'', ``fitness'', ``behavior'' and ``population'' start to get sloppy around the edges whenever we actually \emph{look} at What People Actually Do to Solve Problems With GP.
+Somewhere during that same project years ago, I remember a Project Manager recounting the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were upset and confused: ``We don't want a choice, we want \emph{the best}.'' Of course there could be no \emph{best}, and they supposedly knew that; this response came after \emph{months} of analysis, interviews, technical planning and a collective agreement that the problem was fundamentally multi-objective.
+Even so, when the decision-makers were faced with the \emph{unquestionably} successful results they had collectively specified and specifically asked for\ldots{} those results weren't right enough.
 
-While this can be disappointing for those of us who follow the progress of theoretical work closely, it has more serious consequences.
-The lack of interest in GP among statisticians, planners, designers and mathematical programming aficionados is, I argue, not just a matter of disciplinary border wars.
-I suspect rather that skepticism arises from our own inability to say what's \emph{happening} in any particular GP run---\emph{even when we look}.
-Nor for that matter can we be very helpful in explaining \emph{why}, or what to do in response to any given contingency.
-We have learned that \emph{more CPU time} can sometimes get us past a tricky stage of ``not working'', but any project that lives very far from the bounds of the last century's version of ``genetic programming'' is unexplored territory.
+I would much rather we faced the fact that this shocked response to \emph{seeing what's happening} is inevitable---not just in the form of unwitting surprise from silly lay ``customers'' in an application project, but also from ourselves when we undertake any \emph{interesting} GP project, whether the project is ``theoretical'' or ``practical'', ``small'' or ``large''.
 
-This is not the fault of GP as a field, nor of our theoreticians and engineers.
-Rather I'll argue that a deep philosophical misconception presides over the work, both within and outside our field.
-This misconception undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.
+Any \emph{interesting} project will resist our best expectations. Now and then we should question whether our habitual response to that inevitable resistance is in any sense effective.
 
-I will make my case here in the form of an exercise, or \emph{kata}.
-This is not intended to be a ``thought-experiment'', nor is it a suggestion for a new way of working.
-Rather it is a formal exercise to be undertaken by those of us already working closely with GP systems.
-In the immediate case, my intent is to surface the three problems I've identified above:
+The field of GP has grown quite a bit in the 15 years since my two anecdotes.
+We who follow it closely have watched as frameworks, techniques and methodologies have diversified.
+There are the extraordinary ones we describe to one another in our little workshops, but also an increasing number of newsworthy and admirable public successes.
+
+But this acceleration of innovation brings with it a dilution of the theoretical warrants we use to justify results.
+Most practical successes today highlight unique domain-specific quirks.
+New language and framework implementations makes numerous contingent (and often arbitrary) design and architectural choices.
+Even the fundamental ontological concepts of ``individual'', ``fitness'', ``behavior'', ``generation'' and ``population'' start to get troubling when try to fit many modern algorithms into their categories.
+
+This growing divergence between current theory and practice has consequences both outside and within our field.
+The continued lack of interest in GP among statisticians, planners, designers and fans of traditional mathematical programming models is not merely a matter of disciplinary border wars.
+As we increasingly lose the ability to say \emph{why} something is happening in a given GP project, we are faced with an audience who stopped paying attention to progress in GP last century.
+As a result, students and researchers aiming to convince peers outside the field are forced to undertake mismatched experiments and apply unconvincing statistical tests.
+In response to these constant distractions, much time and attention goes towards unhelpful work guaranteeing that which cannot be guaranteed: running ``replicates'' of a process that is \emph{designed} to provide novel answers, measuring ``reliability'' of a process that \emph{intentionally} skirts dynamical chaos, providing ``summary charts'' of a process that strives to be as complex as the evolution of living beings, and (possibly worst) setting strict deadlines for an open-ended process to ``finish'' or ``succeed''.
+
+I hasten to say this is not a fault in our field, but rather a broader philosophical and cultural problem.
+But it is nonetheless a misconception that undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.
+
+\subsection{A top-down approach}\hypertarget{a-top-down-approach}{}\label{a-top-down-approach}
+
+I will make my case here in the form of an exercise for advanced GP users, or \emph{kata}.
+The habit of pursuing \emph{kata}, ``code retreats'', ``hackathons'' and other skill-honing practices is popular among software developers, and especially among the more advanced.
+Indeed, the title of my exercise (``GP as if you meant it'') is taken from a particularly influential exercise designed by Keith Braithwaite, ``TDD as if you meant it''.
+
+Braithwaite's exercise feels subjectively \emph{harder} for more advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices.
+In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP.
+But like the martial arts exercises from software \emph{kata} were inspired, it isn't intended to be simple or even pleasant for the participants.
+
+I intend it to expose habits I'll tactlessly call ``psueudo-GP'' among those of us who have learned through the years to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project.
+As an important side-effect, it helps surface that philosophical problem I alluded to above.
+But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's decade-old wish.
+It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.
+
+Note, what I'll describe is a ``thought-experiment'', nor is it a serious suggestion for a new way of working on ``real problems'', nor as ``training'' for newcomers to the field.
+Rather it is designed as a rigorous and formal exercise to be undertaken by those of us already working closely with GP systems.
+Its intent is to surface the three shortcomings I've identified above:
 
 \begin{enumerate}
-\item What GP \emph{does}
-\item How GP \emph{unfolds}
-\item What GP is \emph{for}
+\item what GP \emph{does}
+\item how GP \emph{unfolds}
+\item what GP is \emph{for}
 \end{enumerate}
-
-The rules for this exercise, ``Doing GP as if you meant it'', will feel the most artificial and restrictive to those of us with the most experience.
-Like the martial arts exercises from which it is ultimately inspired, it isn't intended to be simple or even pleasant for the participants.
-But because it emphasizes the interaction between the GP system and the user, it not only surfaces the philosophical mistake I've mentioned, but immediately suggests tools with which we can address Rick's decade-old wish.
 

### 2015-03-29 markdown/manuscript.tex
first draft:

@@
 Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.
-
-As I recall, Rick said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, failure to improve, catastrophic lack of diversity, mysterious things like that sense we all get when we look at results that suggests \emph{we made bad choices} of some sort.
+As I recall, he said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, slow and spotty improvement, the catastrophic lack of diversity, and that ineffable feeling all GP professionals experience when they see results and \emph{know} that somewhere, somebody has Chosen Unwisely.
 
@@
 
-This contribution aims squarely at that sensibility which sees the ``symptoms'' and ``pathologies'' as something we as a research community should be trying to stamp out \emph{before} letting users run their own GP systems.
-Instead, I'll argue that because GP differs qualitatively and philosophically from most other machine learning approaches, and also from the broader lay understanding of the ``automated design'' and exploratory research, much of the value GP offers is lost or ignored when we approach it as something to be ``cleaned up''.
+This contribution aims squarely at an unfortunate sensibility I've detected in our field, which sees the many ``symptoms'' and ``pathologies'' as bugs to be eradicated \emph{before} we let ``real users'' run their own GP systems.
+Instead I'll argue (belatedly following Rick's lead) that these ``surprises'' and ``disappointments'' are not only inevitable but should be the main focus of our theoretical and practical work.
+Indeed, much of the value GP can offer can be lost if we approach these phenomena as something to be ``cleaned up'' rather than being \emph{accommodated}.
 
@@
 
-I would much rather we faced the fact that this shocked response to \emph{seeing what's happening} is inevitable---not just in the form of unwitting surprise from silly lay ``customers'' in an application project, but also from ourselves when we undertake any \emph{interesting} GP project, whether the project is ``theoretical'' or ``practical'', ``small'' or ``large''.
+In hindsight, I've come to think this shocked response was inevitable.
+It wasn't an unwitting surprise from some silly lay ``customers'' in an application project.
+Indeed, it's analogous to the way many in our field treat \emph{interesting} GP projects, whether they're ``theoretical'' or ``practical'', ``small'' or ``large''.
+\emph{Any interesting project will resist our initial plans and expectations.} When we're faced with even a little resistance, our first reaction is to imagine something has ``gone wrong in the setup'', shut the thing off and start again with ``better'' parameters.
 
-Any \emph{interesting} project will resist our best expectations. Now and then we should question whether our habitual response to that inevitable resistance is in any sense effective.
+It's unusual for anybody to consider interrogating the system \emph{in itself} when it resists, and adapt or accommodate that resistance.
+Within our field, but also in a surprisingly broad layer of technical society.
 
@@
 In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP.
-But like the martial arts exercises from software \emph{kata} were inspired, it isn't intended to be simple or even pleasant for the participants.
+But like the martial arts exercises from which software \emph{kata} first were inspired, it isn't intended to be simple or even pleasant for the participants.
 
-I intend it to expose habits I'll tactlessly call ``psueudo-GP'' among those of us who have learned through the years to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project.
+I intend it to expose habits I'll tactlessly call ``pseudo-GP'' among those of us who have come to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project.
 As an important side-effect, it helps surface that philosophical problem I alluded to above.
-But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's decade-old wish.
+But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's years-ago wish.
 It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.
@@
 
-\section{How we treat GP, and how it treats us in turn}\hypertarget{how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{how-we-treat-gp-and-how-it-treats-us-in-turn}
-
-Genetic Programming\footnote{And not just Genetic Programming as such, but also the broader discipline to which I claim it belongs and which is not obliged to be either ``genetic'' or ``programming''.
-I prefer to call this looser collection of practices ``generative processing'', and will also abbreviate it ``GP''; assume I mean the latter in every case.} embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and optimization.
-I increasingly suspect the resistance we've all recounted towards GP from our prospective technical and lay audience has little to do with our technical results, but rather arises from uncertainty among that audience with GP's very particular ``way of working''.
-
-There is a tacit assumption, even among GP theorists and practitioners, that science and engineering are only ``rigorous'' when they proceed through a sequence of ordered phases from planning to implementation.
-The ``scientific method'' is most often represented something along the lines of
-
-\begin{quote}
-\begin{enumerate}
-\item conceptualization; 2. planning; 3. design; 4. architecture; 5. implementation; 6. testing; 7. debugging
-\end{enumerate}
-\end{quote}
-
-I am sure very few scientists or engineers of my acquaintance would admit any \emph{real} project (in history) has ever followed this narrative arc in a literal sense.
-But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports:
-
-\begin{quote}
-Because of the body of published work, an insight was had.
-The insight was framed as a formal hypothesis.
-The hypothesis (and current Best Statistical Practices) suggested an experimental design, one familiar and obvious to any in Our Discipline.
-The experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say ``Best Statistical Practices'', right?
-\end{quote}
-
-Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe almost any institutional project management and public policy planning process as well.
-The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.
-
-Of course, nobody ``really believes'' this narrative who has ever done any of the work.
-It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, though.} There have been numerous philosophical challenges to this artificial narrative of course, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos in the 1970s, and many more to be found in the Table of Contents in any Philosophy of Science text.
-
-One in particular is my focus today: Andrew Pickering.
-
-\subsection{Pickering's ``Mangle of Practice''}\hypertarget{pickerings-mangle-of-practice}{}\label{pickerings-mangle-of-practice}
-
-Andrew Pickering's \emph{Mangle of Practice} is a decade old, but surprisingly little-known outside the field of Science Studies.
-His approach is especially noteworthy here because I find it so close to our actual experience using GP.
-Indeed, most colleagues who hear it for the first time utter an inevitable ``didn't we already know this?'' I think the distinction Pickering's approach is able to highlight is exactly the problematic one in our work: between the illusory (but publishable) narrative of ``scientific method'', and the realized experience we have of \emph{performing science}.
-At the cost of glossing a lot of his well-considered structure, let me summarize.
-
-First, the act of ``doing science'' is \emph{at no point whatsoever} to be understood as one in which an isolated ``researcher'' works in an objective and static field of ``externalities'' and ``facts''.
-Rather, research begins only when the researcher undertakes to \emph{makes some artifact}: a block of code, a technical instrument, an equation, a sketch, a maquette or even a thoughtful conversation at a conference.
-Everything before the researcher begins to make these things is a matter of building ``vision'' (my term); research proper only occurs when work is done in the real world.
-I'll call this work product \emph{the thing made}, and keep in mind that it is a proxy for the vague collection we might otherwise call ``the project''.
-
-We see Pickering's model moving quickly away from more traditional views of science when we realize he has given the inanimate \emph{thing made} an agency of its own.
-In any real project, we perceive the \emph{thing made} resisting---whether we imagine that it resists ``in itself'', or as an agent of externalities that impinge on the work in progress to make our driving vision differ from reality.
-The \emph{thing made} comes to represent, in other words, the facts of the actual world, the cultural assumptions and norms of one's discipline, the raw materials and toolkit available to a practitioner, and so forth.
-``Resistance'' here is not merely a problem with a software bug or lost raw materials, but includes one's sense \emph{on seeing it} that something's not quite right.
-The \emph{thing made}'s resistance, as perceived by the researcher, leads her to react in turn.
-
-Consider the language we use in the face of this resistance: it ``feels wrong''; it ``points something out''; it ``wants to do X instead of Y''; it's ``doing something too complicated for me to understand right now''.
-
-And---assuming science, engineering, art or any other creative process is indeed what's being done---the researcher \emph{changes in response to this resistance}.
-That vision changes, the plan adapts, or in some other way the \emph{thing made} causes a response in the state of the researcher herself.
-Pickering's Mangle is this emergent dance of inanimate agency: the researcher starting to follow a vision by making (or altering) a \emph{thing}, and the \emph{thing made} in turn acting as a channel for the world itself to steer the researcher in another direction.
-
-Pickering's ``mangle''\footnote{The word ``mangle'' he has chosen is itself interesting and insightful:
-
-\begin{quote}
-\ldots{}I find ``mangle'' a convenient and suggestive shorthand for the dialectic because, for me, it conjures up the image of the unpredictable transformations worked upon whatever gets fed into the old-fashioned device of the same name used to squeeze the water out of the washing.
-It draws attention to the emergently intertwined delineation and reconfiguration of machinic captures and human intentions, practices, and so on.
-The word ``mangle'' can also be used appropriately in other ways, for instance as a verb.
-Thus I say that the contours of material and social agency are mangled in practice, meaning emergently transformed and delineated in the dialectic of resistance and accommodation\ldots{}.
-\end{quote}} is this emergent dance of agency, undertaken between the researcher and the project: the researcher makes, the \emph{thing made} resists, the researcher is influenced and redirected, \emph{accommodating} the resistance.
-It may seem glib to say that ``no plan survives contact with the enemy,'' but Pickering's narrative of creative work emphasizes the fact that no \emph{revised} plan survives unchanged, either.
-In the traditional narrative, we elide the work as it unfolded and re-frame it as a sort of idealized, apersonal Platonic truth: we use the passive voice, we hide the missteps and confusion, we paint a story moving from vision to plan to success.
-
-In a GP setting, the many modes of resistance we perceive are the very ``symptoms'' Rick mused about.
-Even though we as researchers know we've written all the code and set all the parameters, we're nonetheless willing to speak of a GP run ``doing'' things, as opposed to merely unfolding according to our plan.
-A GP system does not ``resist'' by, for example, ``having the wrong population size''.
-Rather it resists by \emph{causing concern or dissatisfaction in the user}, which in turn sparks in that user a practical (if only explanatory) response, which leads them to change the \emph{thing made}.
-
-\subsection{GP as ``mangle-ish practice''}\hypertarget{gp-as-mangle-ish-practice}{}\label{gp-as-mangle-ish-practice}
-
-The broader field of machine learning seems to take a much more traditional stance towards its subject matter: machine learning frameworks (excepting GP) are each discrete tools aimed at producing standardized and reproducible results to particular statistical questions.
-The result of training a neural network or even a random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the reliable and robust end-product of applying numerical optimization to a well-specified mathematical programming problem.
-Whether one describes these machine learning processes as ``minimizing out-of-sample error'' or ``maximizing information gain'', the supposed strength of most machine learning approaches is the \emph{unsurprising} nature of their use cases and outputs.
-
-On the other hand, GP has the capacity to \emph{tell us stories}---even in the relatively ``simple'' domain of symbolic regression.
-The space under consideration by GP is not merely a vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, functions over inputs, and higher-order functions over those.
-We who work in the field can be glib about the ``open-endedness'' of GP systems, but that open-endedness puts them at odds with their supposed relatives in machine learning.
-While GP \emph{can} be used to explore arbitrarily close to some paradigmatic model, its more typical use case leads to the production of \emph{unexpected} insights---to the degree that a number of us feel justified in treating it as a strong candidate for ``real'' artificial intelligence.
-
-I argue we have that leeway because of the way GP surfaces Pickering's Mangle.
-When the methodology ``works'', it does so by offering \emph{helpful resistance} in our engagement with the problem at hand, whether in the form of surprising answers, or validation of our suspicions, or simply legible suggestions of ways to make subsequent moves.
-GP \emph{dances} with us, while most other machine learning methods are ``mere tools''.
-
-\subsection{Against replication}\hypertarget{against-replication}{}\label{against-replication}
-
-Nonetheless, there seems to be a widespread desire, inside and outside our field, to frame GP as a methodology for producing \emph{unsurprising} models from data, more in keeping with the traditional linear of scientific work.
-That is, an idealized user is expected to proceed something like the users of any other machine learning or mathematical programming framework:
-
-\begin{enumerate}
-\item frame your problem in the correct formal language
-\item ``get'' a GP system
-\item run GP ``on your data''
-\item (whatever this is, it's not our problem)
-\item you have solved your problem
-\end{enumerate}
-
-This is not original with the GP community; there is strong pressure from our peers in other disciplines and our users to promote this use case, not least because it is \emph{exactly} the stance expected in any planning or public policy setting, or in any scientific or programming project management setting.
-That is, we are under tremendous social pressure to treat GP as a \emph{tool} to be invoked in a known, predictable and well-described planning situation.
-
-The resulting resistance from early-adopting users shouldn't be unexpected.
-\emph{Being surprising} may well be the worst conceivable behavior for any tool to be used in a traditional project setting.
-And given that pressure, it's no wonder that so much of GP research is focused on constraining tweaks to bring GP ``into line''.
-If only GP could be ``tamed'' or made ``adaptive'' so that step (4) above \emph{never happens}.
-
-I imagine this is why so many GP research projects strive for rigor in the form of counting replicates which ``find the solution'': they aim not to convince users of the known strengths of GP, but rather demonstrate to critical peers that GP can be ``tamed'' into a mere tool.
-
-What would a ``replicate'' stand for, to a user who sought to exploit GP's strength in a project (whether theoretical or practical) where \emph{search} rather than the algorithm itself is the primary focus?
-Projects which authentically ``use'' GP \emph{must necessarily} be those searching for noteworthy answers---which is to say \emph{surprising} and \emph{interesting} answers---that they could not otherwise obtain.
-Thus, we should better think of a ``replicate'' as a sort of proxy for user frustration in step (4) above: that is, it represents a project in which search begins, stalls, and for which the user cannot see a way to move search forward towards more interesting and useful answers.
-
-But I think we would agree: it is a poor researcher who, when faced with a stumbling block in the form of a black box's misbehavior, doesn't attempt to work around that obstacle.
-Not to \emph{begin the project again from scratch}, but to make changes and continue.
-Any researcher who is using GP \emph{realistically} will be watching, and adjusting, and engaging and interacting with the process of search itself.
-
-When a GP user is viewed as working within the traditional \emph{linear} narrative of research, we see her run a population of 100 individuals for 100 generations, peer at the results that have been dumped into a CSV file, and find them wanting.
-Desiring an actual \emph{answer}, she adjusts the GP parameters and begins another run of 100 generations\ldots{} and repeats as necessary.
-
-But there is \emph{no discernible difference} when we frame this same process of ``many runs'' as a single project involving initial researcher moves, resistances thrown up by the \emph{thing made}, and subsequent accommodations made by the researcher to the new perceived truths.
-Except, that is, in the researcher's view of her own response to resistance: if she comes to the project with \emph{plans} for running ``ten replicates'', we can only assume she has learned from someone that a search algorithm is \emph{supposed} to forget everything it has learned every 100 generations\ldots{}.
-
-I cannot help but be reminded of the fallacy, surprisingly common both in and outside of our field, that ``artificial intelligence'' must somehow be a self-contained and non-interactive process.
-That is, that an ``AI candidate'' loses all authenticity as soon as it is ``tweaked'' or ``adjusted'' in the course of operation.
-It is as if every new-born ``AI'' must be jammed into an air-tight computational container and isolated \emph{until it learns to reason}, and that without exceeding a finite computational budget.
-If humans creating \emph{real} intelligences treated them anything like the way computer scientists insist we treat nascent \emph{artificial} ones, murder charges would be forthcoming.
-
-There are several practical reasons for us to try something different.
-
-Consider our hypothetical GP user.
-In keeping with the Behaviorist standards of GP, she is carefully ``not interfering'' with any given run of 100 generations; she can only peer at a results file after the fact.
-During the course of any 100 generations, all sorts of dynamics have happened: crossover, mutation, selection, all the many random choices.
-Imagine for a moment we were given perfect access to the entire dynamical pedigree of the unsatisfying results she receives at the end, and were able to backtrack to any point in the run and change a single decision.
-Before that point, it's unclear how badly things will actually turn out at the 100-generation mark; at some point after that juncture, it's obvious to anybody watching that the whole thing's a mess.
-
-If such miraculous insights were available, then surely the correct approach would be to intervene and adjust the situation when the crucial point was reached\ldots{} and then continue.
-Lacking (as we do) this miraculous insight, \emph{why then does it seem reasonable to stop any run arbitrarily at a pre-ordained time point and begin again from scratch?}
-
-It is, I think, because the myths of artificial intelligence and the linear narrative of science are so deeply intertwined.
-It is frowned upon to admit in a scientific paper, even when no mistakes were made, that the original vision and plan changed over the course of the project; rather we are expected to describe research \emph{results} as the inevitable outcomes of an ahistorical process, and erase all resistance and accommodation actually done by human beings in the context of their projects.
-Similarly, it feels somehow wrong to admit in a GP project, even if every parameter was set correctly, that the original vision and plan gave way to the inevitable surprises thrown up by GP's inherent tendencies to do just that.
-
-But insofar as GP \emph{surprises us}, and since that is its sole strength over more predictable and manageable frameworks, we must inevitably see a good fraction of those surprises at least in part as \emph{disappointments} rather than encouraging opportunities to change our plans.
-Let's learn new ways to accommodate those disappointments, and stop trying to make them go away.
-
 \section{``TDD as if you meant it''}\hypertarget{tdd-as-if-you-meant-it}{}\label{tdd-as-if-you-meant-it}
@@
 As far as I can tell, Keith Braithwaite first described his \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{training exercise} for software developers in 2009.
-The target of the exercise is ``Pseudo-TDD'': the noted habit among software developers who claim to ``know'' and ``do'' test-driven development as part of their daily work towards a sort of thoughtless approximation of the technique.
+His target was a sense that software developers who thought they were using test-driven development practices were in fact doing something more like ``Pseudo-TDD'', a sort of slapdash and habitual approximation lacking many of the benefits of mindful practice.
 
-I should note that a number of agile software development practices share informative relations to genetic programming's dynamics\footnote{I imagine there is an Engineering Studies thesis in this for some aspiring graduate student: Genetic programming and agile development practices arose in the same period and more or less the same culture, and both informed by the same currents in complex systems and emergent approaches to problem-solving.}, but in this work I'll focus on those of TDD.
-In particular, test-driven development (or more accurately ``test-driven design'') \emph{when done correctly} can break down the complex design space of a software project into a value-ordered set of incremental test cases, focus the developers' attention on those cases alone, inhibit unnecessary ``code bloat'' and feature creep, and produce low-complexity understandable and maintainable software.
+While I've noted elsewhere that several agile software development practices share useful overlaps with the problems of GP science and engineering\footnote{I imagine there is an Engineering Studies thesis in this for some aspiring graduate student: Genetic programming and agile development practices arose in the same period and more or less the same culture, and both informed by the same currents in complex systems and emergent approaches to problem-solving.}, in this work I'll focus on those of TDD.
+In particular, the observation that test-driven development (or more accurately ``test-driven design'') \emph{when done correctly} can break down the complex design space of a software project into a value-ordered set of incremental test cases, focus developers' attention on those cases alone, inhibit unnecessary ``code bloat'' and feature creep, and produce low-complexity understandable and maintainable software.
 
-TDD \emph{as such} is a rigorous process, to the point where it can be described as ``painful'' (though also ``useful'') by experienced programmers.
-The steps are deceptively easy to trivialize and misunderstand, especially for those whose habits of thinking about code are ingrained:
+TDD \emph{as such} is a very constraining and rigorous process---to the point where it can easily be described as ``painful'' (though also ``useful'') by experienced programmers.
+The steps are deceptively easy to gloss, misunderstand or miscommunicate, especially for those whose coding habits are ingrained.
+To paraphrase Beck and Braithwaite:
 
@@
 \item Add a little (failing) test which exercises the next behavior you want to build into your codebase
-\item Run all tests, expecting only the newest to fail
+\item Run all tests, expecting \emph{only the newest} to fail
 \item Make the minimal change to your codebase that permits the new test to pass
-\item Run all tests, expecting them all to succeed
-\item Refactor codebase to remove duplication
+\item Run all tests, expecting them \emph{all} to succeed
+\item Refactor the codebase to remove duplication
 \end{enumerate}
 
-Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the iterative flow of implementation (or ``design'') that it imposes: Each cycle begins with a choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced so far.
-The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working through them as if they could be automated easily.
+Even though a single cycle through this iterative process can take less than a minute, each step can throw itself up as a stumbling block for an experienced programmer.
+But the most salient for us here is the iterative flow of implementation (or ``design'') that the cycle imposes: it begins with a choice of \emph{which little test should next be added}, and ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced so far.
+The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working as though they could be automated easily.
 The \emph{mindfulness} of the process lives in the choice of next steps and (though somewhat less so) of standard refactoring operations.
 
-Braithwaite's exercise does an interesting thing to surface the formal rigor of this approach.
-In it, the participants (willing, of course, because the exercise is a \emph{kata} or ``refresher'' for experienced software developers to hone their skills) are asked to implement a nominally simple project like the game of Tic-Tac-Toe, given an \emph{ordered} list of features to implement and the artificial restriction that they must go farther than normal TDD practice asks.
-Rather than producing a suite of tests and a self-contained codebase, they are forced to use \emph{only} refactoring of code added to tests to produce their eventual ``codebase''.
-In other words, no code can be ``produced'' until the ``smell'' of duplication in the code added to multiple passing tests \emph{provides a warrant} for refactoring it out.
+Braithwaite's exercise does an interesting thing to surface the formal rigor of those decisions, by making them \emph{harder} rather than easier.
+In ``TDD as if you meant it'', the participants (willing, of course) are asked to implement a nominally simple project like the game of Tic-Tac-Toe using TDD, and are given a list of requisite features and an extra constraint.
+Rather than using ``normal'' TDD and producing a suite of tests to exercise a separate and self-contained codebase, they are forced to add code \emph{only} to the tests themselves (to make them pass), and can only produce a separate ``codebase'' when duplication or other ``code smells'' drive them to refactor the code already added to tests.
+In other words, the \emph{demand for a warrant} for writing code is much more stringent.
 
-Further, a facilitator patrols ongoing work and deletes \emph{any and all code not called for by a pre-existing failing test}.
+Throughout the exercise, a facilitator patrols teams of participants and deletes \emph{any and all code not called for by a pre-existing failing test}.
 Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine.
-But as Gojko Adzik emphasizes in his descriptions of workshops, the resulting designs for even simple algorithms in this artificially amplified setting seems much more \emph{open-ended} than it would if the software were built under the typical norms and habits an experienced programmer uses in normal conditions.
+But as Gojko Adzik emphasizes in descriptions of workshops he's facilitated, the resulting designs even for well-known algorithms in this artificially amplified setting become much more ``open-ended'' than would be expected if the code were written under the offhand attention of an experienced programmer without painful constraints.
 
-A number of contextually positive benefits are attributed to agile software development practices, and to TDD within that suite of practices.
-But the one that brings us here today is that aspect surfaced particularly in Adzik's account of Tic-tac-toe:
+Adzik mentions something in his account of a Tic-tac-toe exercise that's especially interesting:
 
@@
 In other words: innovative approaches to the problem at hand began to arise, though there wasn't enough time to finish them in the time alloted for the exercise.
-The analogy to our collective experience to date with genetic programming should start to peek through at this point---though the thoughtful reader will hopefully wonder what utility there is in an analogy drawn between two similarly unsatisfying outcomes.
-
-[more here]
 
@@
 
-In the same way that Braithwaite's onerous coding exercise is intended to drive the attention of its participants toward test-driven design with its obligation to write ``real'' code \emph{only as a refactoring}, I'd like to be able to demand a \emph{warrant} for every step that moves our changing genetic programming setup away from just plain random guessing.
-Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed}.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''.
+In the same way that Braithwaite's coding exercise uses an onerous extra constraint\footnote{In this the sensibility reminds me of the constraint-driven art collective \href{http://en.wikipedia.org/wiki/Oulipo}{Oulipo}, who are perhaps most famous for the \emph{lipogram}, a literary work which cannot use a particular letter of the alphabet.} to drive participants towards more mindful and insightful decision-making, in this exercise I will demand a \emph{warrant} for each implementation decision that moves a running GP setup away from random guessing.
+Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function and \emph{post hoc} analysis is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed} in the course of the run.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''.
 I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''.
@@
 
-Not only do traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent, data-driven argument in favor of starting them \emph{as part of an ongoing search process}.
-Similarly, the initial selection criteria will be limited to a minimal subset of the training data, and expansion (and other alterations) of the training set will have to be made in light of measured progress, not assumptions that ``more is better'' in every case.
+Not only will traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent argument in favor of starting them \emph{as part of an ongoing search process}.
+Similarly, the initial selection criteria will be limited to a minimal single element of the training data, and expansion of the training set in use will have to be made in light of measured progress, not the assumption that ``more is better''.
 
-The result will be an incremental process of refinement of an ongoing search, carried out not at the level of externally-assigned parameter ``tweaks'' but rather by \emph{opening} the black boxes we typically demand and demanding we do surgery to correct their ``pathologies'' (and understand their mysteries) without killing them outright.
-It is not intended as an ``algorithm'' to supplant those used today, but rather as a forced re-description of what we actually already do.
+The result is a painfully incremental process of refinement of an unstoppable search which has \emph{intentionally} ``started wrong'', which must be carried out not by stopping the whole thing and launching it again with parameter ``tweaks'' but rather by \emph{doing surgery on the living patient} to correct any perceived ``pathologies''.
+Along the way, a fraction of the mysteries of ``pathology'' become much clearer and well-defined, even when they cannot be solved easily.
 
-\subsection{A tableau representation}\hypertarget{a-tableau-representation}{}\label{a-tableau-representation}
+Again, let me emphasize that this is not intended as an ``algorithm'' in its own right, but rather as a suite of constraints imposed on what we do without paying attention already.
 
-The exercise proceeds as a sort of game, in which the User and the System take alternating turns.
-During the User's turn, she can see state of the system so far and make any of several \emph{moves} from a limited set, each of which involve changing the settings of a \emph{tableau}, which completely describes the state of the search system.
-During the System's turn, it will execute a finite number of steps in which it creates new individuals.
+\subsection{The game tableau}\hypertarget{the-game-tableau}{}\label{the-game-tableau}
+
+The exercise is structured as a sort of game with two players, one of which is a running GP System, and one which is a human User (or team of Users collaborating on making a single move at a time).
+The two players make alternating turns.
+
+The game ``setup'' can be imagined as a \emph{tableau} with two components: A single list of search {\tt operators}, and a large (and growing) two-dimensional spreadsheet-like table with {\tt answers} as its row labels and {\tt rubrics} as its column labels.
+
+During the User's turn, she can examine the state of the tableau, including all of its history, and can do any amount of data analysis or statistical consideration she wants.
+She is however permitted no more than \emph{two} moves: in each turn she can code and launch one new {\tt operator}, and can code and launch one new {\tt rubric}.
+
+Finally, the User is obliged to provide a convincing \emph{warrant} for every move she make, to be reviewed by an external Facilitator.
+In the event this warrant is seen by the Facilitator to justify the particular move, he has the right to delete all {\tt operators} and {\tt rubrics} added by the User from that point forward \emph{without resetting the System's moves}.
+
+Facilitator buy-in for every move is therefore \emph{strongly} encouraged.
+
+During the System's turn, it will execute a finite number of steps in which it creates new individuals using the specified search {\tt operators} and {\tt rubrics}.
 At the end of the System's turn, control reverts to the User, and vice versa.
 
-The tableau, and the suite of moves available to the User, affect three core aspects of search: {\tt operators}, {\tt answers}, and {\tt rubrics}.
+\subsubsection{Answers}\hypertarget{answers}{}\label{answers}
+
+An {\tt answer} might traditionally be called an ``individual'' in the GP literature.
+It is modeled here as key--value hash (or a simple prototypal object).
+Every new {\tt answer} is born with only a unique {\tt id} and its {\tt script} field set.
+As they ``mature'' during the System's turns, various other attributes (``scores'') will be set by {\tt rubrics}.
+
+Answers never ``die'', and cannot be removed from the tableau by either player as search progresses.
+The User can only reduce or increase the chance that certain types of {\tt answer} contribute to later generations by constructing and launching {\tt rubrics} to adjust selective forces.
 
@@
 
-An {\tt operator} is any function which takes as argument a (possibly empty) collection of {\tt answers} and produces a new collection of {\tt answers}.
-Operators thus include ``random guessing'', which in GP systems is often used to build an initial population, any ``crossover'' and ``mutation'' functions.
-No {\tt operator} can refer to any {\tt answer} not part of its argument set, not can any unset values be set within an {\tt operator}.
+An {\tt operator} is a coded function which takes as its argument an unordered  collection of {\tt answers}, and which produces a new collection of {\tt answers} as output.
 
-So for example, if the User wished an {\tt operator} could be constructed which took 30 {\tt answers} as inputs, read their attributes (including {\tt script} and that subset of their {\tt rubric} values which were already set), and produce a single {\tt answer} in its return set.
-It could \emph{not} take an {\tt answer}, change it, and keep the new one ``if it was better'', since the new {\tt answer} would have no scores at all.
+The initial tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm.
+On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like ``crossover'' or ``mutation'' functions.
+A wide variety of {\tt operator} structures are possible in the domain-specific language provided.
 
-Note that no process is provided for {\tt answers} to be \emph{removed} from a tableau. The framework is purely cumulative.
+The specific set of {\tt answers} to which an {\tt operator} is applied cannot be chosen directly by the User.
+All ``parents'' defined for each {\tt operator} are chosen by \emph{lexicase selection} automatically during the System's turns, using the complete suite of {\tt rubrics} in play when invoked.
+This selection process samples each {\tt rubric} in the tableau with equal probability.
 
-All ``parents'' and other {\tt answer} inputs required by an {\tt operator} are chosen by \emph{lexicase selection} automatically, using the complete suite of {\tt rubrics} in play when invoked.
-Lexicase selection samples each {\tt rubric} with equal probability.
+It is feasible for the User to design a new ``selection'' {\tt operator}, but it must rely on the immediate state of the {\tt answers} it is given; in other words, no {\tt operator} can explicitly invoke a particular {\tt rubric} (or vice versa).
 
-\subsubsection{Answers}\hypertarget{answers}{}\label{answers}
-
-An {\tt answer} might traditionally be called an ``individual'' in GP literature.
-We can model them programmatically as key--value hash.
-All new {\tt answers} are born with only a unique {\tt id} and {\tt script} field set.
-As an {\tt answer} ``matures'' in the unfolding tableau, various other attributes will be set by the System player through the application of {\tt rubrics}.
-
-In the tableau visualization, we represent the unfolding collection of {\tt answers} as the \emph{rows} of a spreadsheet-like table, and their attributes and scores as the \emph{columns}.
+Note again that no mechanism exists which removes {\tt answers} from the tableau.
 
@@
 
-A {\tt rubric} is any function which returns a scalar numerical value, given the instantaneous state of the tableau itself.
-For the most part these values can be understood as ``scores'' for various search objectives, though it is likely that a number of ``implicit objectives'' will also create new {\tt rubrics}.
+A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for a given {\tt answer}, conditioned as needed on the instantaneous tableau state.
+When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') in that {\tt answer}.
 
-A {\tt rubric} cannot store intermediate values, but can refer to the values of other {\tt rubric} columns.
-Thus on specifying a {\tt rubric} like ``sum squared error over a training set'', one must also create  {\tt rubrics} for ``measured error when provided input $i$'' for every element $i$ of the training set.
-If the training set has 100 elements, then the single SSE {\tt rubric} \emph{implicitly} represents a suite of 101 new columns added to the tableau.
+No ``score'' is ever updated in an {\tt anwer} once set by a {\tt rubric}.
+However, if multiple {\tt answers} exist which have the same {\tt script}, each may be scored at different times or in different contexts, and the resulting values may differ.
 
-A {\tt rubric} function does however have access to the state of the entire tableau as needed, including ``forcing'' other {\tt rubric} values to be calculated.
-So it is entirely possible to specify {\tt rubric} functions which score:
+An aggregate or \emph{higher-order} {\tt rubric} can be created, but as a consequence it ``entails'' all of its component {\tt rubrics}.
+So if the User constructs a higher-order {\tt rubric} like ``max error over 30 training cases'', when she launches that {\tt rubric} 31 in total will be launched, with the first 30 being in the form ``absolute error for training case $i$''.
+
+If a User launches such a higher-order {\tt rubric} on her turn, all its entailed {\tt rubrics} will also be added before the System takes its turn.
+
+When one of these ``input'' {\tt rubrics} is applied during the System's turns, it does \emph{not} automatically cause the higher-order aggregate to be applied.
+However, when the higher-order {\tt rubric} is applied, it will force all of its ``input'' {\tt rubrics} to be calculated and applied first.
+
+In addition to ``entailed'' {\tt rubrics} for aggregate scoring, the User can use any of the other extant {\tt rubrics} present in the tableau in her definition, or the state of the interpreter before and after the {\tt script} is run.
+It is therefore entirely possible to specify functions which score:
 
 \begin{itemize}
-\item number of characters in the {\tt answer}'s script
+\item number of characters in the {\tt answer}'s script (since the {\tt script} is an attribute of every {\tt answer} with its own column in the tableau, and therefore its value is exactly equivalent to any other {\tt rubric}-created score)
 \item maximum error measured in any of 35 other {\tt rubrics}
@@
 
-\subsubsection{User moves}\hypertarget{user-moves}{}\label{user-moves}
+Various problem-specific aspects are glossed here. Suffice to say, the construction of reasonable {\tt rubrics} in response the System's moves is the core of the User player's game strategy.
 
-If we think of the tableau as a ``spreadsheet'' with a small list of {\tt operators} and a large sheet of {\tt answers} as rows and {\tt rubrics} as columns, the User can add one new {\tt operator} or one new column.
+\subsubsection{User turn}\hypertarget{user-turn}{}\label{user-turn}
 
-\begin{itemize}
-\item add (or activate from a pre-existing list) exactly one {\tt operator}
-\item add exactly one {\tt rubric}, which may in turn create other \emph{entailed} {\tt rubrics} as described above
-\end{itemize}
+In her turn, the User can do either or both of the following:
+
+\begin{enumerate}
+\item create and launch one {\tt operator}
+\item create and launch one {\tt rubric} (which can be ``higher-order'' and entail others)
+\end{enumerate}
+
+To support these decisions, the User can examine the tableau state and history in detail.
+Before the decisions are implemented, though, a \emph{warrant} must be written for each one.
+Whether the Facilitator is asked to review this is entirely up to the User player, subject to the warnings mentioned above.
 
@@
 
-During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau.
-If we think of the tableau as a ``spreadsheet'' the System can only add new {\tt answers}.
-It follows a single rote cycle to do so:
+During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau, one at a time, using a simple form of lexicase selection.
+Until it reaches its halting state it iterates this cycle:
 
 \begin{enumerate}
-\item select an {\tt operator} from those in play, with equal probability
+\item select one {\tt operator} from those in the tableau, with equal probability
 \item apply \emph{lexicase selection} to the tableau to select the required number of input {\tt answers}, filling in missing {\tt rubric} scores as needed
-\item apply that one {\tt operator} to the inputs to produce new {\tt answers}, and append those new {\tt answers} immediately to the tableau
-\item {\tt HALT} if the number of new {\tt answers} added in this turn meets or exceeds the number present when the turn started (in other words, if the number of {\tt answers} has doubled); otherwise, go to step (1)
+\item apply the {\tt operator} to the inputs to produce one or more new {\tt answers}, and append those new {\tt answers} immediately to the tableau
+\item {\tt HALT} if the number of new {\tt answers} meets or exceeds the limit; otherwise, go to step (1)
 \end{enumerate}
+
+\subsubsection{Choosing a language}\hypertarget{choosing-a-language}{}\label{choosing-a-language}
+
+No particular constraints apply to the language or representation chosen for the exercise, except that it should be sophisticated enough to support \emph{redundant capacity}.
+That is, for any given algorithmic goal (or stage), there should be multiple paths to success.
+So for example if the problem could reasonably be expected to involve ordered lists, then it would be wise if the language had at least two different ways to ``deal with lists'': iterators, recursion, a comprehensive set of second-order functional operators, an explicit {\tt List} type with associated methods, that sort of thing.
+
+If the language is sophisticated enough to have ``libraries'' of instructions and types intended for specialized domains, I'd strongly encourage that \emph{all of these should be used}.
+That is, the facilitator should err on the side of ``winnowing complexity'', rather than forcing the User and System to \emph{invent basic data structures} or the idea of floating-point numbers at the same time they're trying to solve the ``real'' problem under consideration.
+
+Finally, if there is a choice between a familiar language and an odd one, then the odd one should be chosen, all other things being equal.
+
+\subsubsection{Problems}\hypertarget{problems}{}\label{problems}
+
+In keeping with my notes on languages above, in the choice of problems the planner should aim for something that would strike any experienced GP person as ``ambitious''.
+Which is to say: a reasonably good programmer would be able to hand-code the answer with a day's thought and work, but only in a familiar language; in the oddball GP language, it should feel ``practically'' impossible to hand-code.
+That said, it should \emph{also} be clear to any programmer familiar with the language specification that the needed \emph{components} are all there, and that there are enough ``parts'' to approach any sub-task that crops up along the way from more than one angle.
+
+Colleagues have pointed out that even if the chosen problem turns out to be ``too simple'', in the sense that the System player solves it quickly without much input from the User, then the ``problem'' addressed in the exercise immediately becomes one of driving the System to find a second \emph{dissimilar} solution\ldots{} and without restarting the search process, of course.
+
+\subsubsection{The Facilitator}\hypertarget{the-facilitator}{}\label{the-facilitator}
+
+There need not be a formal ``Facilitator'' if the User is willing and able to honor the spirit of the exercise.
+But especially in more formal and public situations, it makes perfect sense for the Facilitator to be the one to choose the problem and language, and also to provide the necessary technical infrastructure on which the computational part of the exercise will be run.
+
+Any warrant that relies on ``best practices'', habit, tradition, or anything considered or described as ``obvious'' deserves to be invalidated as soon as it's noticed.
+Warrants should rather be judged valid only when the state of the tableau (or its history) is invoked to make a cogent argument for the submitted moves.
+
+That said, a warrant does \emph{not} need to be ``technical'' or even ``rigorous'', but merely robust enough to be convincing.
+For example, ``I made crossover because I think it needs to \emph{search in between the parents},'' sounds to me like a shoddy excuse that invokes received wisdom.
+``I made crossover to drop the variance on this rubric and foster inbreeding of \emph{these} solutions \emph{here},'' given a glance at the tableau and charts on hand, sounds good to me.
 
@@
 \begin{itemize}
-\item the only {\tt operator} is ``random guess'', which creates a new {\tt answer} with an arbitrary script
-\item no {\tt rubrics} exist (only the {\tt script} and {\tt id} attributes of the {\tt answers})
-\item 50 new {\tt answers} will be produced in the first System turn
-\item No {\tt operator} can \emph{evaluate} any {\tt answer} except in a self-contained local namespace which lacks access to any tableau values or state information
-\item There is no mechanism for \emph{removing} {\tt answers} from the tableau
-\item The System player \emph{always} uses lexicase selection, and \emph{always} uses all {\tt rubrics} as the selection features with equal probability.
-\item A {\tt rubric} can only \emph{run} a single {\tt answer} once; stochastic scripts will only be sampled one time, and no {\tt rubric} score is ever recalculated after the first time
+\item the only {\tt operator} is ``random guess'', which creates one new {\tt answer} with an arbitrary script
+\item the only {\tt rubrics} present are the {\tt script} and unique {\tt id} fields of the {\tt answers}
+\item the System player will add 250 new {\tt answers} (or slightly more, depending on the {\tt operators} chosen) in each of its turns
+\item the System player moves first
+\item there is no mechanism for \emph{removing} {\tt answers} from the tableau
+\item the System player \emph{always} uses lexicase selection, \emph{always} choses {\tt operators} with equal probability from the current list, and \emph{always} uses all {\tt rubrics} as the selection features with equal probability
+\item a {\tt rubric} can only \emph{run} a single {\tt answer} once; stochastic scripts will only be sampled one time, and no {\tt rubric} score is ever recalculated after the first time, though multiple copies of the same stochatic {\tt answer} will probably end up with different scores in the same field
 \end{itemize}
 
-\subsubsection{Interface: the Goal of the Exercise}\hypertarget{interface-the-goal-of-the-exercise}{}\label{interface-the-goal-of-the-exercise}
+\subsection{Interface}\hypertarget{interface}{}\label{interface}
 
-During the User's turn, they can of course interrogate the tableau in any way they want, without changing it.
-The point of the exercise is to drive the User to explore and create analytics and visualizations which can better inform their decisions over the course of the game.
+[this entire section is still being transcribed from an ill-advised experiment with LaTeX table typesetting; it'll be inserted as soon as possible]
+
+\subsubsection{The {\tt operator} language}\hypertarget{the-operator-language}{}\label{the-operator-language}
+
+\subsubsection{The {\tt rubric} language}\hypertarget{the-rubric-language}{}\label{the-rubric-language}
+
+\subsubsection{Visualization}\hypertarget{visualization}{}\label{visualization}
 
@@
 
-\section{Exploration and exploitation interfaces and affordances}\hypertarget{exploration-and-exploitation-interfaces-and-affordances}{}\label{exploration-and-exploitation-interfaces-and-affordances}
+[this entire section is still being transcribed from an ill-advised experiment with LaTeX table typesetting; it'll be inserted as soon as possible]
 
-\section{Final thoughts: What should it mean to \emph{act intelligently}?}\hypertarget{final-thoughts-what-should-it-mean-to-act-intelligently}{}\label{final-thoughts-what-should-it-mean-to-act-intelligently}
+\section{Summing up: How we treat GP, and how it treats us in turn}\hypertarget{summing-up-how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{summing-up-how-we-treat-gp-and-how-it-treats-us-in-turn}
 
-The idea of this exercise came from observing the differences between experienced and ``novice'' (though with strong technical skills) GP users as they explored new problems in a GP setting.
+Genetic Programming\footnote{And not just Genetic Programming as such, but also the broader discipline to which I claim it belongs and which is not obliged to be either ``genetic'' or ``programming''.
+I prefer to call this looser collection of practices ``generative processing'', and will also abbreviate it ``GP''; assume I mean the latter in every case.} embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and optimization.
+I increasingly suspect the resistance from colleagues we've all encountered towards GP has little to do with our technical results as such.
+Rather it arises from unfamiliarity with GP's very particular ``way of working''.
+We are used to it---perhaps to the point of taking it for granted---but the mismatch and resulting response from those less familiar with the system has I think led us to \emph{mask} our sense of it in an effort to fit in.
+
+Briefly, the systemic fault lies in the awful ``scientific method'' that permeates our cultural dialog about the practice of science and engineering.
+You know the one, which is often something along the lines of:
+
+\begin{quote}
+vision $\rightarrow$ planning $\rightarrow$ design $\rightarrow$ architecture $\rightarrow$ implementation $\rightarrow$ testing $\rightarrow$ debugging
+\end{quote}
+
+I'm sure very few scientists or engineers of my acquaintance would admit any \emph{real} project has \emph{ever} followed this narrative in a literal sense.
+But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports:
+
+\begin{quote}
+Based on the body of published work, an insight was had.
+The insight was framed as a formal hypothesis.
+The hypothesis (shaped by current Best Statistical Practices) immediately suggested an experimental design, which design is obvious to anyone familiar with Our Discipline.
+That experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say ``Best  Practices'', right?
+\end{quote}
+
+``Nothing surprising happened while we were working on this project,'' in other words.
+
+Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe almost any institutional project management or public policy planning process as well.
+The flow in every case is essentially from \emph{vision} to \emph{plan}, \emph{plan} to \emph{implementation}, \emph{implementation} to \emph{verification}, and \emph{verification} to \emph{validation}.
+
+Of course, nobody ``really believes'' this narrative who has ever done the work.
+It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, I suspect.} And I am not the first to point it out; the history of Philosophy of Science is built primarily from the numerous philosophical challenges to this artificial narrative, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos and Feyerabend in the 1970s, and with many more to be found in the Table of Contents in any Philosophy of Science text.
+
+Andrew Pickering is the inspiration for this project, though.
+
+\subsection{On the Mangle of Practice}\hypertarget{on-the-mangle-of-practice}{}\label{on-the-mangle-of-practice}
+
+Andrew Pickering's monograph \emph{The Mangle of Practice} is a decade old, but surprisingly little-known outside his discipline of Science Studies.
+His approach is especially useful here, because I find it captures a surprising amount of our \emph{actual experience} of building and using GP systems.
+Indeed, most colleagues who hear it for the first time utter an inevitable ``didn't we already know this?''
+
+Pickering's approach focuses on that problematic division I've sketched above, between the illusory (but publishable) linear narrative of the ``scientific method'', and the realized experience we all have had of \emph{performing science} (or Mathematics, or Engineering, or for that matter Art).
+At the cost of glossing too much of his well-considered structure, let me summarize.
+
+First, I should remind us all that the \emph{performance of science} is just that: not an isolated but perceptive mind standing apart from the world, working in an objective and static field of ``externalities'' and ``facts'', but a \emph{performance} done by a human being present in that world.
+In Pickering's  framework, we can say that research proper begins only when the researcher makes some artifact or formal ``machine'' in the world: writes a block of code, designs a technical instrument, considers a particular equation, draws a pencil sketch, or even a has a thoughtful conversation at a conference.
+Let me call this artifact \emph{the thing made}.
+This is not the scientific paper that results in the end of the project, but rather the sum of all the sketchy notes, the cloud of more-or-less coherent ideas, the code and instruments, the collected observations, the plan and the community of colleagues helping with that plan: everything done in the world, mentally or physically, towards the goals of the project.
+
+Pickering's model jumps quickly away from more traditional ``scientific methods'' when he treats this mechanism as capable of \emph{agency in its own right}, and is willing to say that it can and does \emph{resist} our intentions.
+In the context of the Mangle, the \emph{thing made} is the conduit of the facts of the actual world to the researcher (and also of the cultural assumptions and norms of one's discipline, of the inherent tendencies of the raw materials and the practitioner's toolkit).
+``Resistance'' here is not merely a reference to a software bug, a mathematical mistake or a shortage of crucial raw materials, but specifically denote one's sense \emph{on seeing it} that ``something's not quite right''.
+In other words, it is the \emph{thing made}'s resistance ``on behalf of'' the real world which forces the researcher to reconsider, change or adapt her plans, or otherwise \emph{accommodate} that resistance.
+
+The inspiration for granting agency to machinic abstractions (or even concrete dynamics) is obvious whenever we hear the language that crops out in the course of our work: the code ``feels wrong''; the mathematics is ``pointing something out''; the machine ``wants to do X instead of Y''.
+Projects in science, engineering and the arts do not proceed from a stage of planning to a stage of implementation, except in the ahistorical mythology of our published papers.
+Pickering's Mangle\footnote{The word ``mangle'' he has chosen is itself interesting and insightful:
+
+\begin{quote}
+\ldots{}I find ``mangle'' a convenient and suggestive shorthand for the dialectic because, for me, it conjures up the image of the unpredictable transformations worked upon whatever gets fed into the old-fashioned device of the same name used to squeeze the water out of the washing.
+It draws attention to the emergently intertwined delineation and reconfiguration of machinic captures and human intentions, practices, and so on.
+The word ``mangle'' can also be used appropriately in other ways, for instance as a verb.
+Thus I say that the contours of material and social agency are mangled in practice, meaning emergently transformed and delineated in the dialectic of resistance and accommodation\ldots{}.
+\end{quote}} does much better at capturing our first-hand experience of the work as an emergent dance of human and machinic agency \emph{with one another}.
+The researcher starts to follow her vision by making (and altering) some artificial thing, that \emph{thing made} acts as a channel for the world itself to resist, and as a result the researcher \emph{accommodates} that resistance by moving in some different direction.
+In the traditional linear narrative, we elide the work as it unfolded and re-frame it as a sort of idealized, apersonal Platonic truth: we use the passive voice, we hide the missteps and confusion, after the fact paint a story which flows from vision to plan to success.
+But within the dance of Pickering's Mangle, the degree to which we as researchers can successfully \emph{accommodate} the inevitable resistance
+
+In a GP setting, the notion of ``machinic agency'' seems much closer to our experience; after all, we are obliged not only a pick or write a specialized formal language to represent the space of solutions, but in every project we must also cobble together some framework of search operators, fitness operators, algorithms and instrumentation.
+But even when we've written all the code and set all the parameters personally, we're \emph{still} willing to speak of a GP run ``doing'' things, as opposed to merely unfolding according to our plan.
+Indeed, if it happens by chance that GP ``runs according to our plan'' then arguably the problem was too \emph{boring} to be worth mentioning\ldots{}.
+
+I will argue below that GP's power (and difference from other machine learning approaches) lies in the very particular form of resistance it can offer us as its users.
+This is not merely ``resistance'' of the frustrating kind: we use GP most effectively when we want it to surprise us.
+The ``surprise'' is certainly something we are forced to accommodate with just as much attention and concentration as any more annoying resistance which might be thrown up, for example when we are forced to piece together ``how the hell this thing actually works''.
+
+It is worth saying explicitly now, and again later, that by granting the \emph{thing made} a machinic agency of its own, we can frame the problem of ``pathology'' and ``symptoms'' in GP more constructively.
+A GP system does not resist by ``having the wrong population size'' or ``having too high a mutation rate''; those are not \emph{behaviors}, but tiny facets of a complex plan instantiated (to some extent) in a complex dynamical system that is the GP system.
+Rather I would say a GP system resists by \emph{raising concern or causing dissatisfaction in a human user}.
+It is inevitably that \emph{human} observer who produces a productive and accommodating response.
+
+\subsection{GP as ``mangle-ish practice''}\hypertarget{gp-as-mangle-ish-practice}{}\label{gp-as-mangle-ish-practice}
+
+The broader field of machine learning seems to take a much more traditional stance towards its subject matter.
+The result of training a neural network or even a random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the reliable and robust end-product of applying numerical optimization to a well-specified mathematical programming problem.
+Indeed, the supposed strength of most machine learning approaches is the very \emph{unsurprising} nature of their use cases and outputs.
+
+On the other hand, we all know that GP embodies a capacity to \emph{tell us stories}, even in the relatively ``simple'' domain of symbolic regression.
+The space under consideration by GP is not some vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, functions over inputs, and higher-order functions over those.
+We who work in the field can be glib about the ``open-endedness'' of GP systems, but that open-endedness puts GP at odds with its supposed machine learning cousins.
+While GP can be used to explore arbitrarily close to some parametric model, its more common use case is exactly the production of \emph{unexpected} insights.
+
+When the GP approach ``works'', it does so by offering \emph{helpful resistance} in our engagement with the problem at hand, whether in the form of surprising answers, validation of our suspicions, or simply as a set of legible suggestions of ways to make subsequent moves.
+GP \emph{dances} with us, while most other machine learning methods are exactly the ``mere tools'' they have been designed to be.
+
+\subsection{Against replication}\hypertarget{against-replication}{}\label{against-replication}
+
+Nonetheless, there seems to be a widespread desire inside and outside our field to frame GP as a way of exploring \emph{unsurprising} models from data.
+As with neural networks or decision trees, the machine learning tool-user is expected to proceed something like this:
+
+\begin{enumerate}
+\item frame your problem in the correct formal language
+\item ``get'' a GP system
+\item run GP ``on your data''
+\item (whatever happens here, it's not our problem)
+\item you have solved your problem
+\end{enumerate}
+
+This does not arise from within the GP community; it is \emph{exactly} the stance expected in any planning or public policy setting, or in any waterfall management or programming project management setting.
+Twenty years and more of skepticism even from early-adopting users shouldn't be unexpected.
+\emph{Being surprising} might be the worst conceivable outcome for any tool used in a traditional project management setting.
+
+Given that pressure, it's no wonder that so much of GP research is focused on constraining tweaks to bring GP ``into line''.
+If only GP could be ``tamed'' or made ``adaptive'' so that step (4) above \emph{never happens}\ldots{}.
+I imagine this is why so many GP research projects strive for rigor in the form of counting replicates which ``find a solution'': they aim not to convince \emph{users}, but rather to demonstrate to critical peers that GP can be ``tamed'' into another mere tool.
+
+Think about ``replicates'' for a moment.
+What might a ``replicate'' be for a user who wants to exploit GP's strength of discovering new solutions?
+If one is searching for noteworthy answers---which is to say \emph{surprising} and \emph{interesting} answers---then a ``replicate'' must be some sort of proxy for user frustration in step (4) above.
+That is, a ``replicate'' stands in for a project in which search begins, stalls, and where the user cannot see a way to accommodate the resistance in context\ldots{} and just gives up trying.
+
+I cannot help but be reminded of the fallacy, surprisingly common both in and outside of our field, that ``artificial intelligence'' must somehow be a self-contained and non-interactive process.
+That is, that an ``AI candidate'' loses  authenticity as soon as it's ``tweaked'' or ``adjusted'' in the course of operation.
+It is as if every new-born ``AI'' must be quickly jammed into an air-tight computational container and isolated \emph{until it learns to reason by itself}, and for that matter without exceeding a finite computational budget.\footnote{If humans creating real intelligences treated them anything like the way computer scientists insist we treat nascent artificial ones, murder charges would be forthcoming.}
+
+Maybe the responsibility for this failure lies neither with the stupid user or the shoddily-made GP system, but in our unwillingness to let them give more useful feedback to one another.
+
+Consider our typical GP user who is ``not interfering'' with her running GP system; she can only peer at a results file after the fact, and can't fiddle with the ``settings'' while the thing is actually working.
+But of course during any given run of 100 generations, \emph{all sorts} of dynamics have happened: crossover, mutation, selection, all the many random choices.
+
+Imagine for a moment we were given perfect access to the entire dynamical pedigree of the unsatisfying results she receives at the end, and were able to backtrack to any point in the run and change a single decision.
+Before that point, it's unclear how badly things will actually turn out at the 100-generation mark; at some point after that juncture, it's obvious to anybody watching that the whole thing's a mess.
+
+If such miraculous insights were available, then surely the correct approach would be to intervene and adjust the situation when the crucial point was reached\ldots{} and then continue.
+Lacking (as we do) this miraculous insight, \emph{why then does it seem reasonable to stop any run arbitrarily at a pre-ordained time point and begin again from scratch?}
+
+Insofar as GP \emph{surprises us}, and since that is its sole strength over more predictable and manageable frameworks, we must inevitably see a good fraction of those surprises at least in part as disappointments rather than encouraging opportunities to change our plans.
+Let's learn new ways to accommodate those disappointments, and stop trying to make them ``go away''.

### 2015-03-29 tex/Book/gpbook-Tozier-2015-03-23/tozier.tex
fixed LaTeX setup per instructions
session.md appendix using definition list (becomes description list in LaTeX)
first draft:

--- 2015-03-27 tex/Book/gpbook-Tozier-2015-03-23/chapter.tex
@@
 
-\abstract{(to be written last)}
+\abstract{In this contribution I describe a \emph{kata}, or exercise intended for advanced GP users, called ``GP as if you meant it''.
+In the exercise, the human participant(s) are charged with trying to ``rescue'' an ineffectual but unstoppable GP system which is set up initially to only use ``random guessing''.
+The exercise is in the form of a game of alternating turns, in which the human ``User'' player is provided complete information but only a very limited toolkit, and the (automated) GP System ``player'' can only be modified and ``adjusted'' by adding new search operators and objectives.
+The crux of the exercise is the human players' development of cogent \emph{warrants} for every design decision they make when building and using GP systems---decisions which otherwise can be habitual or arbitrary.
+By explicitly prohibiting the most common response of  ``stop it and restart with different parameters'', the participants are mindful of ways to \emph{accommodate} the noted ``pathologies'' and ``symptoms''.
+The computational participant tends to just get warmer.}
 
@@
 
+[version of \today\ at \currenttime]
+
 \subimport{../../../markdown/}{manuscript}
@@
 \bibliographystyle{spbasic}
-\bibliography{gp-bibliography,chapter}
+\bibliography{gp-bibliography,tozier}

### 2015-06-29 markdown/manuscript.tex
final revisions

@@
 
-\section{Why}\hypertarget{why}{}\label{why}
+\section{Why: An Excuse}\hypertarget{why-an-excuse}{}\label{why-an-excuse}
 
-More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved genetic programming.
-As we chatted one day, Rick was asked what he'd most like to see as part of the research program of GP ``moving forward''.
-Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.
-As I recall, he said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, slow and spotty improvement, the catastrophic lack of diversity, and that ineffable feeling all GP professionals experience when they see results and \emph{know} that somewhere, somebody has Chosen Unwisely.
+More than a decade ago, Rick Riolo, Bill Worzel and I worked together on a genetic programming consulting project.
+As we chatted one day, one of them---I don't recall which, and accounts vary---was asked what he'd most like to see as GP ``moved forward'', and said he'd want the field to think more about the ``symptoms'' we so often see when we apply evolutionary search processes in complex settings.
+That is: premature convergence, slow and spotty improvement, a catastrophic lack of diversity, and more generally that ineffable feeling all GP professionals experience when we look at results and \emph{know we have chosen unwisely}.
 
-The literature abounds with well-written papers describing tips for avoiding local minima, improving on common search operators, and describing ``horse races'' between Bad Old and Better New search methodologies applied to benchmark problems.
-But I take Rick's challenge not only to mean that it would be useful to have a catalog which lists situations where search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood.
+Now the reader will point out that the literature in our field overflows with well-written papers describing tips for avoiding local minima, improving on common search operators, and running ``horse races'' between Bad Old and Better New search methodologies applied to benchmark problems.
+But while as a rule these are presented as a sort of \emph{general principle}, most in practice are \emph{case studies} in which it is shown, for example, that search operator $X$ acts under contingency $Y$ and sometimes produces outcome $Z$.
 
-This contribution aims squarely at an unfortunate sensibility I've detected in our field, which sees the many ``symptoms'' and ``pathologies'' as bugs to be eradicated \emph{before} we let ``real users'' run their own GP systems.
-Instead I'll argue (belatedly following Rick's lead) that these ``surprises'' and ``disappointments'' are not only inevitable but should be the main focus of our theoretical and practical work.
-Indeed, much of the value GP can offer can be lost if we approach these phenomena as something to be ``cleaned up'' rather than being \emph{accommodated}.
+I am not left with a sense that this catalog addresses my colleague's stated wish.
+While it is surely necessary to compile such a list of individual observations under particular suites of experimental treatments and benchmarks, it is insufficient until we can achieve the skill of recognizing \emph{when} something noteworthy is happening.
+In particular, I want to focus on those frequent but contingent situations in which we are willing to say our GP system \emph{resists our expectations}.
 
-\subsection{Challenges}\hypertarget{challenges}{}\label{challenges}
+A good deal of this chapter will be spent explaining just what I mean by this  particular usage of ``resistance'', but for the moment the idea will be clear enough from this simple mental exercise: What are things GP has ``done''---in the context of a particular project---that have left you feeling unsatisfied, confused, or frustrated?
+Not because you've made a simple mistake setting a parameter, or introduced a bug in a codebase, but because (at that moment) you have no idea why the GP process is doing \emph{that thing} under those particular circumstances?
+Maybe it's not converging when you know it should; maybe it's exploring unexpectedly complex algorithms rather than obvious simpler ones, despite your reasonable parameter settings; maybe it's failing to solve simple problems but easily solving hard ones.
+In any case, the system ``resists'' your careful \emph{and knowledgeable} expectations by inexplicably refusing to follow your plan.
 
-Somewhere during that same project years ago, I remember a Project Manager recounting the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were upset and confused: ``We don't want a choice, we want \emph{the best}.'' Of course there could be no \emph{best}, and they supposedly knew that; this response came after \emph{months} of analysis, interviews, technical planning and a collective agreement that the problem was fundamentally multi-objective.
-Even so, when the decision-makers were faced with the \emph{unquestionably} successful results they had collectively specified and specifically asked for\ldots{} those results weren't right enough.
+My general point, and the particular point of this chapter and the exercise it describes, is that we should then ask: \emph{When we feel GP is resisting us, what should we do---and why?} Further, after more than twenty years of work in the field, I think we have learned enough for this to be reasonable and productive research program.
 
-In hindsight, I've come to think this shocked response was inevitable.
-It wasn't an unwitting surprise from some silly lay ``customers'' in an application project.
-Indeed, it's analogous to the way many in our field treat \emph{interesting} GP projects, whether they're ``theoretical'' or ``practical'', ``small'' or ``large''.
-\emph{Any interesting project will resist our initial plans and expectations.} When we're faced with even a little resistance, our first reaction is to imagine something has ``gone wrong in the setup'', shut the thing off and start again with ``better'' parameters.
+I'm sure every reader has at least one story they can tell, in which GP has ``resisted'' in this sense.
+I want to draw our attention to these incidents from a conviction that it is exactly the steps we take to \emph{accommodate} GP's resistance which provoke our sudden insights into the structure of our problem, drive us to build a ``workaround'' that becomes a novel selection algorithm or architecture, and lead us to successfully re-frame our project to use a completely new approach.
+Almost any chapter in this series of GPTP Workshop Proceedings will contain a story just like this: We create an age-layered population because working without age-layering didn't perform as expected \citep{Hornby:2009:GPTP}; we invent a new selection mechanism because traditional algorithms worked as described, but failed to capture crucial details of our problems \citep{Spector:2012:GECCOcompA}; we create an entirely new representation and suite of search operators because our goals aren't met by snipping up and recombining in the traditional way \citep{ryan:2003:GPTP}; and so on.
 
-It's unusual for anybody to consider interrogating the system \emph{in itself} when it resists, and adapt or accommodate that resistance.
-Within our field, but also in a surprisingly broad layer of technical society.
+The point in each case is: one gains little \emph{insight} into a problem when GP quickly pops out the ``right answer'' without a fight.
+Too often I see papers treating this ubiquitous resistance as something to be eradicated \emph{before} ``real users'' are allowed run their own GP searches.
+Instead I'll argue here that ``surprises'' and ``disappointments'' are not only inevitable but \emph{are the main source of value in many GP projects}, and as such should be the main focus of our theoretical and practical work.
 
-The field of GP has grown quite a bit in the 15 years since my two anecdotes.
-We who follow it closely have watched as frameworks, techniques and methodologies have diversified.
-There are the extraordinary ones we describe to one another in our little workshops, but also an increasing number of newsworthy and admirable public successes.
+Every interesting project will resist our plans and expectations.
+But in our field, whenever we're faced with even a bit of this sort of behavior, our first reaction is generally to act as though something has ``gone wrong in the setup'', then shut the misbehaving run down so we can start again with ``better parameters'' next time.
+I find it unusual for anybody to interrogate the system \emph{in itself} when it resists, or to make an attempt to adapt or accommodate perceived resistance.
+I hasten to say this is not a fault with our field, but rather a symptom of broader philosophical and cultural problems in our approach to programming and computational research projects, and our understanding of computing more generally.
 
-But this acceleration of innovation brings with it a dilution of the theoretical warrants we use to justify results.
-Most practical successes today highlight unique domain-specific quirks.
-New language and framework implementations makes numerous contingent (and often arbitrary) design and architectural choices.
-Even the fundamental ontological concepts of ``individual'', ``fitness'', ``behavior'', ``generation'' and ``population'' start to get troubling when try to fit many modern algorithms into their categories.
+The exercise I describe in this chapter is intended to bring our attention back on GP as a dynamical process \emph{in itself}, as opposed to a tool to be adjusted ``in between'' applications.
+It may be the case that other ``traditional'' machine learning methodologies are built on better-defined information-theoretic foundations, and come with suites of strong statistical tests for overfitting and robustness; as a result it's perfectly reasonable to treat them as \emph{tools}, and as \emph{malfunctioning} or just being \emph{wrong for a problem} when they act in unexpected ways.
+But GP is somehow a different sort of animal: when it ``resists'', we are left with an essentially unlimited choice of how we should \emph{accommodate} that resistance.
 
-This growing divergence between current theory and practice has consequences both outside and within our field.
-The continued lack of interest in GP among statisticians, planners, designers and fans of traditional mathematical programming models is not merely a matter of disciplinary border wars.
-As we increasingly lose the ability to say \emph{why} something is happening in a given GP project, we are faced with an audience who stopped paying attention to progress in GP last century.
-As a result, students and researchers aiming to convince peers outside the field are forced to undertake mismatched experiments and apply unconvincing statistical tests.
-In response to these constant distractions, much time and attention goes towards unhelpful work guaranteeing that which cannot be guaranteed: running ``replicates'' of a process that is \emph{designed} to provide novel answers, measuring ``reliability'' of a process that \emph{intentionally} skirts dynamical chaos, providing ``summary charts'' of a process that strives to be as complex as the evolution of living beings, and (possibly worst) setting strict deadlines for an open-ended process to ``finish'' or ``succeed''.
+\subsection{On mindful exercises}\hypertarget{on-mindful-exercises}{}\label{on-mindful-exercises}
 
-I hasten to say this is not a fault in our field, but rather a broader philosophical and cultural problem.
-But it is nonetheless a misconception that undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.
+The habit of pursuing \emph{kata}, ``code retreats'', ``hackathons'' and other skill-honing practices is popular among software developers, and especially among the more advanced.
+It apparently arose independently among a few groups in the 1980s, but Dave Thomas seems to have first used the Japanese term \citep{Thomas2013}.
 
-\subsection{A top-down approach}\hypertarget{a-top-down-approach}{}\label{a-top-down-approach}
+Indeed, the title of my exercise (``GP As If You Meant It'') is directly inspired by from an exercise designed by Keith Braithwaite, ``TDD as if you meant it'' \citep{Braithwaite2011}.
+One of the interesting aspects of Braithwaite's exercise is that it feels subjectively ``harder'' when attempted by advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices.
 
-I will make my case here in the form of an exercise for advanced GP users, or \emph{kata}.
-The habit of pursuing \emph{kata}, ``code retreats'', ``hackathons'' and other skill-honing practices is popular among software developers, and especially among the more advanced.
-Indeed, the title of my exercise (``GP as if you meant it'') is taken from a particularly influential exercise designed by Keith Braithwaite, ``TDD as if you meant it''.
+In the same way, the exercise I describe will feel \emph{most} artificial and restrictive to those of us with the most experience with GP.
+But like the martial arts exercises by which software \emph{kata} first were inspired, it isn't intended to be simple or even pleasant for the participants.
+Just as Braithwaite's exercise targeted what he calls ``Pseudo-TDD'', I intend mine to surface habits we can think of as ``pseudo-GP'': the sense that it's cheap and painless to \emph{just shut it off and start over} when we start sensing a problem is arising.
 
-Braithwaite's exercise feels subjectively \emph{harder} for more advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices.
-In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP.
-But like the martial arts exercises from which software \emph{kata} first were inspired, it isn't intended to be simple or even pleasant for the participants.
+\subsection{Caveats}\hypertarget{caveats}{}\label{caveats}
 
-I intend it to expose habits I'll tactlessly call ``pseudo-GP'' among those of us who have come to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project.
-As an important side-effect, it helps surface that philosophical problem I alluded to above.
-But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's years-ago wish.
-It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.
+What I describe here should be considered a ``thought-experiment'' backed up by my own experiences; there are only a few sketchy manual implementations to date.
+Some readers have imagined that the ``automated'' System player must somehow be a self-contained and tested process running on the cloud (and that is an eventual goal), but in all my early tests that role has been played by a handful of simple Ruby scripts, edited between ``turns'' by the Facilitator to reflect changes made by the User player.
+In other words, \emph{there need not be an actual autonomous ``computer player'' in this game}, unless you feel compelled to write one up beforehand, and the game can as easily be played by a Facilitator who \emph{codes} the moves of the System and \emph{plays that role}, plus the User as described.
 
-Note, what I'll describe is a ``thought-experiment'', nor is it a serious suggestion for a new way of working on ``real problems'', nor as ``training'' for newcomers to the field.
-Rather it is designed as a rigorous and formal exercise to be undertaken by those of us already working closely with GP systems.
-Its intent is to surface the three shortcomings I've identified above:
+Further, there is no reason the User player need be a single human being.
+Indeed, it's common practice in many software development \emph{katas} to work in teams or even as an aggregated crowd of everybody in the room.
+Many hands make light work, especially on a tight schedule.
+In other words, these roles are intended to be ``notional'' rather than definitional.
 
-\begin{enumerate}
-\item what GP \emph{does}
-\item how GP \emph{unfolds}
-\item what GP is \emph{for}
-\end{enumerate}
+I should also point out that this game is not a serious suggestion for a new way of working on ``real problems'', nor as any sort of ``training'' for newcomers to the field of GP.
+Rather it is designed as a rigorous and formal exercise in mindful practice, to be undertaken by people already working closely enough with GP systems to recognize the problems when they arise.
 
@@
 
-As far as I can tell, Keith Braithwaite first described his \href{http://cumulative-hypotheses.org/2011/08/30/tdd-as-if-you-meant-it/}{training exercise} for software developers in 2009.
-His target was a sense that software developers who thought they were using test-driven development practices were in fact doing something more like ``Pseudo-TDD'', a sort of slapdash and habitual approximation lacking many of the benefits of mindful practice.
+Keith Braithwaite seems to have first described this training exercise for software developers in 2009 \citep{Braithwaite2011}.
+His target was a sense that software developers who thought they were using test-driven development practices \citep{Beck2002} were in fact doing something more like ``Pseudo-TDD'', a sort of slapdash and habitual approximation lacking many of the benefits of mindful practice.
 
@@
 Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine.
-But as Gojko Adzik emphasizes in descriptions of workshops he's facilitated, the resulting designs even for well-known algorithms in this artificially amplified setting become much more ``open-ended'' than would be expected if the code were written under the offhand attention of an experienced programmer without painful constraints.
+But as Gojko Adzic \citep{Adzic2009} has said in descriptions of workshops he's facilitated, the resulting designs even for well-known algorithms in this artificially amplified setting become much more ``open-ended'' than would be expected if the code were written under the offhand attention of an experienced programmer without painful constraints.
 
-Adzik mentions something in his account of a Tic-tac-toe exercise that's especially interesting:
+Adzic passes along some observations in his account  of a Tic-Tac-Toe exercise \citep{Adzic2009} that are especially interesting for me:
 
@@
 
-\section{GP as if we meant it}\hypertarget{gp-as-if-we-meant-it}{}\label{gp-as-if-we-meant-it}
+\section{GP As If You Meant It}\hypertarget{gp-as-if-you-meant-it}{}\label{gp-as-if-you-meant-it}
 
-In the same way that Braithwaite's coding exercise uses an onerous extra constraint\footnote{In this the sensibility reminds me of the constraint-driven art collective \href{http://en.wikipedia.org/wiki/Oulipo}{Oulipo}, who are perhaps most famous for the \emph{lipogram}, a literary work which cannot use a particular letter of the alphabet.} to drive participants towards more mindful and insightful decision-making, in this exercise I will demand a \emph{warrant} for each implementation decision that moves a running GP setup away from random guessing.
-Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function and \emph{post hoc} analysis is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed} in the course of the run.\footnote{Braithwaite's participants often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''.
+In the same way that Braithwaite's coding exercise uses an onerous extra constraint\footnote{In this the sensibility reminds me of the constraint-driven art collective Oulipo \citep{Becker:2012}, who are perhaps most famous for the \emph{lipogram}, a literary work which cannot use a particular letter of the alphabet.} to drive participants towards more mindful and insightful decision-making, in this exercise I will demand a \emph{warrant} for each implementation decision that moves a running GP setup away from random guessing.
+Braithwaite's target of ``Pseudo-TDD'' suggests an analogous ``Pseudo-GP'': one in which the fitness function and \emph{post hoc} analysis is the only ``interface'' with the problem itself, and where the representation language, search operators, search objectives and other algorithmic ``parameters'' are \emph{fixed} in the course of the run.\footnote{Braithwaite's participants \citep{Braithwaite2012} often acknowledge they \emph{know} and \emph{use} TDD as it's formally described, but rarely take the time to do so unless ``something goes wrong''.
 I imagine many GP users will say they \emph{know} and \emph{use} all the innumerable design and setup options of GP, but treat them as adjustments to be invoked only when ``something goes wrong''.
@@
 
-Not only will traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent argument in favor of starting them \emph{as part of an ongoing search process}.
-Similarly, the initial selection criteria will be limited to a minimal single element of the training data, and expansion of the training set in use will have to be made in light of measured progress, not the assumption that ``more is better''.
+Not only will traditional search operators like crossover, mutation and [negative] selection not come ``for free'' in this variant, but in every case we must develop a cogent argument in favor of starting them as part of an ongoing search process.
+Similarly, the initial selection criteria will be limited to a  single training case, and expansion of the active training set will have to be made \emph{in response to particular features} of observed progress, not merely on the basis of the assumption that ``more will be better''.
 
-The result is a painfully incremental process of refinement of an unstoppable search which has \emph{intentionally} ``started wrong'', which must be carried out not by stopping the whole thing and launching it again with parameter ``tweaks'' but rather by \emph{doing surgery on the living patient} to correct any perceived ``pathologies''.
-Along the way, a fraction of the mysteries of ``pathology'' become much clearer and well-defined, even when they cannot be solved easily.
+The result is a painfully incrementalized process, one that focuses on the refinement and eventual correction of an unstoppable search which was \emph{intentionally} ``started wrong'', and which must be carried out by \emph{doing surgery on the living patient} to correct perceived ``pathologies'' and ``resistance''.
+Along the way, a fraction of the mysteries of ``pathology'' has the potential to be much clearer and better-defined.
 
-Again, let me emphasize that this is not intended as an ``algorithm'' in its own right, but rather as a suite of constraints imposed on what we do without paying attention already.
+\section{Overview}\hypertarget{overview}{}\label{overview}
 
-\subsection{The game tableau}\hypertarget{the-game-tableau}{}\label{the-game-tableau}
+The exercise is structured as though it were a game for two players, plus a Facilitator who establishes the ground rules, provides any needed technical infrastructure, and acts as referee.
+One player is (or represents) a running GP System, and one player is (or represents) a human User trying to mindfully drive the System's performance in a desirable direction over a series of turns.
 
-The exercise is structured as a sort of game with two players, one of which is a running GP System, and one which is a human User (or team of Users collaborating on making a single move at a time).
-The two players make alternating turns.
+While I speak below of the System player ``being'' a self-contained software process, it is of course a loose role that might more easily be played by another human, potentially the Facilitator herself, writing and running a simple series of scripts on a laptop.
+Similarly, while I may say that the User player ``is'' a single human being, it could as easily be a room-full of students or workshop participants, or a mailing list voting over many weeks on strategies for each turn.
+Indeed, in working out the exercise as it's described here, I've ``played'' all the roles myself, simultaneously, and still found interesting and unexpected insight.
 
-The game ``setup'' can be imagined as a \emph{tableau} with two components: A single list of search {\tt operators}, and a large (and growing) two-dimensional spreadsheet-like table with {\tt answers} as its row labels and {\tt rubrics} as its column labels.
+In preparing the \emph{kata}, the Facilitator selects a \emph{target problem}, which should be a supervised learning task for which plenty of data is available.
+The target should not be ``toy'' in the sense of having a simple, well-known answer; rather it should be challenging and open-ended enough to warrant a publication if solved (a problem that has \emph{recently} been solved might do in a pinch, though there is no shortage of open ones).
+Any good book of mathematical recreations (for example, \citep{Winkler:2003} or nearly any book by Martin Gardner, Ian Stewart or Ivan Moscovich) will provide numerous abstract problems that have never been attempted with GP.
 
-During the User's turn, she can examine the state of the tableau, including all of its history, and can do any amount of data analysis or statistical consideration she wants.
-She is however permitted no more than \emph{two} moves: in each turn she can code and launch one new {\tt operator}, and can code and launch one new {\tt rubric}.
+The Facilitator will also need to choose a representation language, set up an initial Tableau, and provide enough software infrastructure so that the System's turns can be made easily.
+In each turn of the game, both the System and the User players take actions to modify and extend the initial Tableau: the User's options include adding \emph{rubrics} and \emph{operators} (described below), and in its turns the System acts ``mindlessly'' by \emph{invoking} the specified rubrics and operators to add a fixed number of new individuals to the growing population.
 
-Finally, the User is obliged to provide a convincing \emph{warrant} for every move she make, to be reviewed by an external Facilitator.
-In the event this warrant is seen by the Facilitator to justify the particular move, he has the right to delete all {\tt operators} and {\tt rubrics} added by the User from that point forward \emph{without resetting the System's moves}.
+The User's goal is to drive the System towards producing \emph{sufficiently good solutions} to the target problem, but their decisions are constrained by the obligation to provide a \emph{warrant} for every change made which is convincing to the Facilitator.
+The particular definition of a ``sufficiently good solution'' is left for the players and Facilitator to decide in context: it may be impossible to completely ``solve'' some target problems, given the tools at hand.
 
-Facilitator buy-in for every move is therefore \emph{strongly} encouraged.
+It should always be kept in mind that \emph{the point of the exercise is to evoke interesting and useful warrants}, not merely to drive the System in a desired direction.
+In each turn the User is obliged to produce a convincing \emph{warrant} for every move she makes, which must be reviewed and approved by the Facilitator before play proceeds.
+These warrants need not be \emph{factually correct}, but they must be \emph{convincing} in the context of the game state at the time they're put forth.
 
-During the System's turn, it will execute a finite number of steps in which it creates new individuals using the specified search {\tt operators} and {\tt rubrics}.
-At the end of the System's turn, control reverts to the User, and vice versa.
+\subsection{The Tableau}\hypertarget{the-tableau}{}\label{the-tableau}
+
+The game ``board'' is a \emph{Tableau} with two components: A list of search {\tt operators}, and a two-dimensional spreadsheet-like table which uses {\tt answers} as its row labels and {\tt rubrics} as its column labels.
+Initially the Tableau is empty, except for a single entry on the list of {\tt operators} labeled ``random guess''.
+
+The ``random guess'' operator is set up by the Facilitator before play begins; it produces a single random {\tt answer} for the target problem, with no arguments.
+Think of it as equivalent to the ``initialization'' function used to produce a single ``random'' individual in a traditional GP setting.
+
+During the User's turn, she can examine the state of the Tableau, including any or all of its history, and the algorithms in play, and apply any amount of data analysis or statistical work she wants.
+On her turn she is permitted no more than \emph{two} moves: She can (optionally) code and append one new {\tt operator} to the list of {\tt operators}, and she can (optionally) code and append one new {\tt rubric} column to that table.
+
+During the System's turn, it will produce a fixed number of new {\tt answers}.
+It creates each new {\tt answer} by first picking an entry from the list of {\tt operators} with uniform probability.
+When an {\tt operator} is chosen, it executes the indicated algorithm, selecting parents from the {\tt answers} table as needed.
+For each required selection of a parent, the scores recorded in the table of {\tt answers} vs {\tt rubrics} are used (see below).
+Once parents are selected, the {\tt operator} is applied and a new {\tt answer} is immediately appended to that part of the Tableau.
 
@@
 
-An {\tt answer} might traditionally be called an ``individual'' in the GP literature.
-It is modeled here as key--value hash (or a simple prototypal object).
-Every new {\tt answer} is born with only a unique {\tt id} and its {\tt script} field set.
-As they ``mature'' during the System's turns, various other attributes (``scores'') will be set by {\tt rubrics}.
+What I'm calling an {\tt answer} here would probably be called an ``individual'' in the GP literature.
+In this case, it is a particular \emph{script} or program in the representation language the Facilitator has chosen for the target problem.
 
-Answers never ``die'', and cannot be removed from the tableau by either player as search progresses.
-The User can only reduce or increase the chance that certain types of {\tt answer} contribute to later generations by constructing and launching {\tt rubrics} to adjust selective forces.
+Answers never ``die'', and cannot be removed from the Tableau by either player.
 
@@
 
-An {\tt operator} is a coded function which takes as its argument an unordered  collection of {\tt answers}, and which produces a new collection of {\tt answers} as output.
+Each {\tt operator} is a function which takes as its argument an unordered  collection of zero or more {\tt answers}, and which produces a new collection of one or more {\tt answers} as output.
 
-The initial tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm.
-On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like ``crossover'' or ``mutation'' functions.
-A wide variety of {\tt operator} structures are possible in the domain-specific language provided.
+The initial Tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm.
+On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like crossover or mutation.
+The only permitted argument is a set of zero or more {\tt answers}; no numerical or other parameters are allowed.
+A wide variety of {\tt operator} algorithms are still possible, and the details of the code in which they are implemented is left to the Facilitator to specify as part of the game setup.
 
-The specific set of {\tt answers} to which an {\tt operator} is applied cannot be chosen directly by the User.
-All ``parents'' defined for each {\tt operator} are chosen by \emph{lexicase selection} automatically during the System's turns, using the complete suite of {\tt rubrics} in play when invoked.
-This selection process samples each {\tt rubric} in the tableau with equal probability.
+Except for some unusual edge cases, the \emph{specific} set of {\tt answers} to which an {\tt operator} will be applied cannot be chosen directly by the User.
+All ``parents'' defined for each {\tt operator} are chosen independently (and incrementally) by \emph{lexicase selection}, using the suite of {\tt rubrics} in play when the System takes its turn.
+This lexicase selection process samples every {\tt rubric} in the Tableau with equal probability.
+It may be possible for the User to design a new ``selection'' {\tt operator} which takes as its argument ``all the answers'' and somehow culls that collection down to a subset---but to do so, it must rely on the immediate state of the {\tt answers} it is given; {\tt rubric} scores are ``stored'' only in the Tableau {\tt answer} table itself, not as gettable attributes of the {\tt answers}, making it very difficult for any new {\tt operator} algorithm to use {\tt rubric} scores in its implementation.
 
-It is feasible for the User to design a new ``selection'' {\tt operator}, but it must rely on the immediate state of the {\tt answers} it is given; in other words, no {\tt operator} can explicitly invoke a particular {\tt rubric} (or vice versa).
-
-Note again that no mechanism exists which removes {\tt answers} from the tableau.
+Note again: no mechanism exists which \emph{removes} {\tt answers} from the Tableau.
 
@@
 
-A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for a given {\tt answer}, conditioned as needed on the instantaneous tableau state.
-When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') in that {\tt answer}.
+A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for any given {\tt answer}, conditioned (as needed) on the instantaneous Tableau state.
+When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') for that {\tt answer} in the appropriate cell of the ``spreadsheet'' portion of the Tableau.
 
-No ``score'' is ever updated in an {\tt anwer} once set by a {\tt rubric}.
+No score is ever changed for an {\tt answer}, once it has been set by a {\tt rubric}.
 However, if multiple {\tt answers} exist which have the same {\tt script}, each may be scored at different times or in different contexts, and the resulting values may differ.
 
-An aggregate or \emph{higher-order} {\tt rubric} can be created, but as a consequence it ``entails'' all of its component {\tt rubrics}.
-So if the User constructs a higher-order {\tt rubric} like ``max error over 30 training cases'', when she launches that {\tt rubric} 31 in total will be launched, with the first 30 being in the form ``absolute error for training case $i$''.
+An aggregate or \emph{higher-order} {\tt rubric} can be created, but its existence ``entails'' all of the component {\tt rubrics} from which its score is derived.
+So, for example, if the User constructs a {\tt rubric} like ``maximum absolute error observed over any of these 30 training cases'', on the turn when she adds that {\tt rubric} up to 31 new columns will be added to the spreadsheet portion of the Tableau: one for each of the 30 requisite ``sub-{\tt rubrics}'', which produce as their scores the absolute error for a single training case, and also one \emph{aggregate} {\tt rubric} which calculates and reports ``maximum of those other 30 scores''.
+Whenever the User submits such a higher-order {\tt rubric} on her turn, all of the entailed {\tt rubrics} are added automatically if they do not already exist.
+Note however that there must still be a \emph{warrant} for the {\tt rubric}, which convinces the Facilitator of its potential usefulness.
 
-If a User launches such a higher-order {\tt rubric} on her turn, all its entailed {\tt rubrics} will also be added before the System takes its turn.
-
-When one of these ``input'' {\tt rubrics} is applied during the System's turns, it does \emph{not} automatically cause the higher-order aggregate to be applied.
-However, when the higher-order {\tt rubric} is applied, it will force all of its ``input'' {\tt rubrics} to be calculated and applied first.
-
-In addition to ``entailed'' {\tt rubrics} for aggregate scoring, the User can use any of the other extant {\tt rubrics} present in the tableau in her definition, or the state of the interpreter before and after the {\tt script} is run.
-It is therefore entirely possible to specify functions which score:
+The User can use any information present in the Tableau in building new {\tt rubric} functions (such as row numbers, string values of {\tt answers}, or scores), and can also examine the detailed state of the interpreter before and after the {\tt script} is run.
+It should be clear therefore that the User can specify {\tt rubric} functions which score aspects of the problem such as:
 
 \begin{itemize}
-\item number of characters in the {\tt answer}'s script (since the {\tt script} is an attribute of every {\tt answer} with its own column in the tableau, and therefore its value is exactly equivalent to any other {\tt rubric}-created score)
+\item the absolute error for a single training case
+\item the number of tokens in the {\tt answer}'s script
 \item maximum error measured in any of 35 other {\tt rubrics}
-\item number of {\tt div0} errors produced when running with a particular set of inputs
-\item number of stochastic instructions appearing in the {\tt answer}'s script, compared to \emph{all other {\tt answers}}
+\item number of {\tt div0} errors produced when running a script with a particular set of inputs
+\item number of stochastic instructions appearing in the {\tt answer}'s script
+\item rewrite difference between the scored script the most common {\tt answer} in the population
+\item (and so on)
 \end{itemize}
 
-Various problem-specific aspects are glossed here. Suffice to say, the construction of reasonable {\tt rubrics} in response the System's moves is the core of the User player's game strategy.
+Various problem-specific aspects are glossed here, and it is left to the Facilitator to be reasonable in the context of the target problem and the representation she has chosen.
+Suffice to say, the construction of useful {\tt rubrics} in response the System's moves is the core of the User player's game strategy.
 
-\subsubsection{User turn}\hypertarget{user-turn}{}\label{user-turn}
+\subsection{Lexicase Selection}\hypertarget{lexicase-selection}{}\label{lexicase-selection}
 
-In her turn, the User can do either or both of the following:
+All {\tt operator} inputs are chosen by lexicase selection. Slightly simplifying Spector's original description \citep{Spector:2012:GECCOcompA}, the following algorithm can be used:
+
+\begin{itemize}
+\item (beginning with a ``set under consideration'' which includes all {\tt answers}\ldots{})
+\item for each {\tt rubric} in a random permutation of all the {\tt rubrics} in the Tableau, discard all {\tt answers} from the set under consideration whose score on this {\tt rubric} is sub-optimal (relative to the current set under consideration)
+\item if multiple {\tt answers} remain after all {\tt rubrics} have been applied, return one {\tt answer} picked randomly, with uniform probability, from the remaining set under consideration
+\end{itemize}
+
+At the beginning of the game when no {\tt rubrics} have been added and no {\tt operators} exist which require ``parents'', no selection occurs or needs to occur.
+But note that the algorithm as described would \emph{still} provide a parent, even when no {\tt rubrics} have been specified: an empty set of {\tt rubrics} would be immediately exhausted without eliminating any {\tt answers}, and then a single {\tt answer} would be selected with uniform probability from that complete set.
+
+Lexicase selection has several characteristics which argue for its use here; in particular it is interesting because it seems (anecdotally, but see also Helmuth in this volume) to be a relatively ``slow'' selection method, permitting a diversity of {\tt answers} to coexist in a population at the same time.
+However I admit readily that I've specified it as a core part of this \emph{kata} because it is \emph{unfamiliar to most prospective players}, and therefore more likely to produce unexpected behavior of a useful sort.
+
+\subsection{The User's turn}\hypertarget{the-users-turn}{}\label{the-users-turn}
+
+In her turn, the User can do either (or both) of the following:
 
 \begin{enumerate}
-\item create and launch one {\tt operator}
-\item create and launch one {\tt rubric} (which can be ``higher-order'' and entail others)
+\item add one {\tt operator} to the list in the Tableau
+\item add one {\tt rubric} to the Tableau, which can be ``higher-order'' and therefore entail others
 \end{enumerate}
 
-To support these decisions, the User can examine the tableau state and history in detail.
-Before the decisions are implemented, though, a \emph{warrant} must be written for each one.
-Whether the Facilitator is asked to review this is entirely up to the User player, subject to the warnings mentioned above.
+To support these decisions, the User can examine the Tableau state and history in detail. Before any code is implemented, though, a \emph{warrant} must be written for each one.
 
-\subsubsection{System turn}\hypertarget{system-turn}{}\label{system-turn}
+\subsubsection{Warrants}\hypertarget{warrants}{}\label{warrants}
 
-During its turn, the System player adds a specified minimum number of new {\tt answers} to the tableau, one at a time, using a simple form of lexicase selection.
-Until it reaches its halting state it iterates this cycle:
+A \emph{warrant} is a verbal or written argument which spells out the justification for one of the User's moves \emph{in the context of the game as it stands when the move is made}.
 
-\begin{enumerate}
-\item select one {\tt operator} from those in the tableau, with equal probability
-\item apply \emph{lexicase selection} to the tableau to select the required number of input {\tt answers}, filling in missing {\tt rubric} scores as needed
-\item apply the {\tt operator} to the inputs to produce one or more new {\tt answers}, and append those new {\tt answers} immediately to the tableau
-\item {\tt HALT} if the number of new {\tt answers} meets or exceeds the limit; otherwise, go to step (1)
-\end{enumerate}
+Suppose, for example, that the User has not yet added \emph{any} {\tt rubrics}, and would like to add a new one that scores {\tt answers} for a single training case.
+Since the System is simply making new {\tt answers} with the ``random guess'' {\tt operator}, it seems perfectly reasonable to warrant this new {\tt rubric} simply by pointing out that selection can't drive search towards better answers without at least one training case being used.
 
-\subsubsection{Choosing a language}\hypertarget{choosing-a-language}{}\label{choosing-a-language}
+Suppose further that later in the same game the User has added more {\tt rubrics}, such that more than 100 training cases are being used to select parents for the several {\tt operators} in play.
+In this context, the reasons given for adding a new {\tt rubric} that specifies \emph{just that one particular training case} must surely be very different: perhaps there is a problematic region of the response surface, or a different training case that needs additional ``support'' to differentiate between two close input states (and so forth).
+
+That said, a warrant does \emph{not} need to be ``technical'' or even ``rigorous'', but merely robust enough to be convincing \emph{in the moment}.
+For example, ``I made a crossover {\tt operator} because I think we need to search for new answers `in between' the parents,'' sounds to me like a shoddy excuse that invokes received wisdom.
+On the other hand, ``I made crossover so we can drop the variance on this rubric and foster inbreeding of \emph{these} solutions \emph{here},'' given a glance at the Tableau and charts on hand, should be more convincing.
+
+Note though that decisions and the warrants that supported them may well prove to have been wrong in hindsight.
+This in itself is an interesting and useful outcome in the game: when an earlier decision does not actually produce the expected effect in the system, this is a perfect point for the User to be reminded of the apparent inconsistency.
+Such ``failed'' warrants shouldn't be rescinded, but it may be appropriate to bring them up as ``concerns'' in later moves, especially if they begin to accumulate.
+
+There is one particular warrant that the Facilitator should always permit, as long as it isn't abused by the User: ``Because I have no idea what will happen when I do this, but I suspect it may be useful in order to [X] later on.'' In other words, it is perfectly reasonable for the User to state outright they are ``exploring'' the range of system behaviors.
+More generally, the basis on which the User, and therefore the Facilitator, decides to make or permit a change should always depend on the history of the game so far.
+Indeed,  reasoning will \emph{always} change dramatically over the course of any interesting game, under the constraint that the Facilitator should never approve a change on the argument that ``that's the way we always do it''.
+Every clear and convincing \emph{why} argument will surely be contingent on the immediate state of the system and dependent on all the prior decisions made by the User and Facilitator.
+
+\subsection{The Facilitator}\hypertarget{the-facilitator}{}\label{the-facilitator}
+
+The Facilitator is responsible for picking a representation language and target problem, for providing  technical infrastructure as required for the \emph{kata}, and for approving and implementing any changes to the System that the User player makes.
+As noted several times above, \emph{warrants} are the focus of the exercise, and as such great care should be taken that each decision is well-justified.
+
+\subsubsection{Choosing a representation language}\hypertarget{choosing-a-representation-language}{}\label{choosing-a-representation-language}
 
 No particular constraints apply to the language or representation chosen for the exercise, except that it should be sophisticated enough to support \emph{redundant capacity}.
-That is, for any given algorithmic goal (or stage), there should be multiple paths to success.
-So for example if the problem could reasonably be expected to involve ordered lists, then it would be wise if the language had at least two different ways to ``deal with lists'': iterators, recursion, a comprehensive set of second-order functional operators, an explicit {\tt List} type with associated methods, that sort of thing.
+That is, for any given algorithmic goal, there should be multiple paths to success.
+So for example if the problem's solution could reasonably be expected to involve ordered lists, then it would be best if the language had at least two different ways to ``use lists'', for example with iterators, recursion, a comprehensive set of second-order functional operators, an explicit {\tt List} type with associated methods, and so forth.
 
 If the language is sophisticated enough to have ``libraries'' of instructions and types intended for specialized domains, I'd strongly encourage that \emph{all of these should be used}.
-That is, the facilitator should err on the side of ``winnowing complexity'', rather than forcing the User and System to \emph{invent basic data structures} or the idea of floating-point numbers at the same time they're trying to solve the ``real'' problem under consideration.
+That is, the Facilitator should err on the side of ``winnowing complexity'', rather than forcing the User (and thus the System) to \emph{invent basic data structures} or the idea of floating-point numbers at the same time they're trying to solve the ``real'' problem under consideration.
 
-Finally, if there is a choice between a familiar language and an odd one, then the odd one should be chosen, all other things being equal.
+Finally, if there is a choice between a familiar language and an odd one, then the odd one should be chosen, all other things being equal. Unfamiliarity can be a useful constraint here.
 
-\subsubsection{Problems}\hypertarget{problems}{}\label{problems}
+\subsubsection{Target problems}\hypertarget{target-problems}{}\label{target-problems}
 
-In keeping with my notes on languages above, in the choice of problems the planner should aim for something that would strike any experienced GP person as ``ambitious''.
-Which is to say: a reasonably good programmer would be able to hand-code the answer with a day's thought and work, but only in a familiar language; in the oddball GP language, it should feel ``practically'' impossible to hand-code.
+As with the choice of language, in the choice of problems the planner should aim for something that would strike any experienced GP person as ``ambitious''.
+Which is to say: a reasonably good programmer would be able to hand-code the answer with a day's thought and work, but only in a familiar language; in an oddball GP language, it should feel ``practically impossible'' to hand-code.
 That said, it should \emph{also} be clear to any programmer familiar with the language specification that the needed \emph{components} are all there, and that there are enough ``parts'' to approach any sub-task that crops up along the way from more than one angle.
 
-Colleagues have pointed out that even if the chosen problem turns out to be ``too simple'', in the sense that the System player solves it quickly without much input from the User, then the ``problem'' addressed in the exercise immediately becomes one of driving the System to find a second \emph{dissimilar} solution\ldots{} and without restarting the search process, of course.
-
-\subsubsection{The Facilitator}\hypertarget{the-facilitator}{}\label{the-facilitator}
-
-There need not be a formal ``Facilitator'' if the User is willing and able to honor the spirit of the exercise.
-But especially in more formal and public situations, it makes perfect sense for the Facilitator to be the one to choose the problem and language, and also to provide the necessary technical infrastructure on which the computational part of the exercise will be run.
-
-Any warrant that relies on ``best practices'', habit, tradition, or anything considered or described as ``obvious'' deserves to be invalidated as soon as it's noticed.
-Warrants should rather be judged valid only when the state of the tableau (or its history) is invoked to make a cogent argument for the submitted moves.
-
-That said, a warrant does \emph{not} need to be ``technical'' or even ``rigorous'', but merely robust enough to be convincing.
-For example, ``I made crossover because I think it needs to \emph{search in between the parents},'' sounds to me like a shoddy excuse that invokes received wisdom.
-``I made crossover to drop the variance on this rubric and foster inbreeding of \emph{these} solutions \emph{here},'' given a glance at the tableau and charts on hand, sounds good to me.
+Colleagues have pointed out that even if the chosen problem turns out to be ``too simple'', in the sense that the System solves it quickly without much input from the User, then the ``problem'' addressed in the exercise can be extended by the Facilitator to one of driving the System to find a second \emph{dissimilar} solution\ldots{} still without restarting the search process, of course.
+Keeping in mind that the purpose of the exercise is to provoke insight into the justification for particular decisions, this decision (by the Facilitator) can be justified by being a perfectly reasonable event in a real research project, in which intellectual ``stretch goals'' are commonplace when resources permit.
 
@@
 \item the only {\tt operator} is ``random guess'', which creates one new {\tt answer} with an arbitrary script
-\item the only {\tt rubrics} present are the {\tt script} and unique {\tt id} fields of the {\tt answers}
-\item the System player will add 250 new {\tt answers} (or slightly more, depending on the {\tt operators} chosen) in each of its turns
+\item no {\tt rubrics} are present
 \item the System player moves first
 \item there is no mechanism for \emph{removing} {\tt answers} from the tableau
-\item the System player \emph{always} uses lexicase selection, \emph{always} choses {\tt operators} with equal probability from the current list, and \emph{always} uses all {\tt rubrics} as the selection features with equal probability
-\item a {\tt rubric} can only \emph{run} a single {\tt answer} once; stochastic scripts will only be sampled one time, and no {\tt rubric} score is ever recalculated after the first time, though multiple copies of the same stochatic {\tt answer} will probably end up with different scores in the same field
+\item the System player \emph{always} uses lexicase selection, \emph{always} choses {\tt operators} with equal probability from the current list, and \emph{always} uses all {\tt rubrics} in the Tableau
+\item a {\tt rubric} can only be \emph{run} on a given {\tt answer} once; stochastic scripts will only be sampled one time, and no {\tt rubric} score is ever recalculated after the first time, though multiple copies of the same stochastic {\tt answer} will probably end up with different scores in the same field
 \end{itemize}
 
-\subsection{Interface}\hypertarget{interface}{}\label{interface}
+\subsubsection{The System's turn}\hypertarget{the-systems-turn}{}\label{the-systems-turn}
 
-[this entire section is still being transcribed from an ill-advised experiment with LaTeX table typesetting; it'll be inserted as soon as possible]
+During its turn, the System player will add a specified number of new {\tt answers} to the Tableau, one at a time, using a simple form of lexicase selection.
+Until it reaches its halting state it iterates this cycle:
 
-\subsubsection{The {\tt operator} language}\hypertarget{the-operator-language}{}\label{the-operator-language}
+\begin{enumerate}
+\item select one {\tt operator} from those in the Tableau, with equal probability
+\item apply lexicase selection to select the required number of input {\tt answers}
+\item apply the chosen {\tt operator} to the inputs {\tt answers} to produce one or more new {\tt answers}, and append those new {\tt answers} immediately to the Tableau
+\item {\tt HALT} if the number of new {\tt answers} meets or exceeds the limit, and delete any extra {\tt answers} that exceed the limit; otherwise, go to step (1)
+\end{enumerate}
 
-\subsubsection{The {\tt rubric} language}\hypertarget{the-rubric-language}{}\label{the-rubric-language}
+The number of {\tt answers} created in each turn should be enough to have a chance of providing new and useful information to the User, and not so much that the System state grows out of control.
+I would suggest 100 or 500 {\tt answers} per turn; this is about the size of the typical ``population'' for most GP users, and is on a comfortable scale for them.
 
-\subsubsection{Visualization}\hypertarget{visualization}{}\label{visualization}
-
-\section{An example session}\hypertarget{an-example-session}{}\label{an-example-session}
-
-[this entire section is still being transcribed from an ill-advised experiment with LaTeX table typesetting; it'll be inserted as soon as possible]
-
-\section{Summing up: How we treat GP, and how it treats us in turn}\hypertarget{summing-up-how-we-treat-gp-and-how-it-treats-us-in-turn}{}\label{summing-up-how-we-treat-gp-and-how-it-treats-us-in-turn}
+\section{Why: A Warrant}\hypertarget{why-a-warrant}{}\label{why-a-warrant}
 
@@
 I prefer to call this looser collection of practices ``generative processing'', and will also abbreviate it ``GP''; assume I mean the latter in every case.} embodies a very particular \emph{stance} towards the scientific and engineering work of modeling, design, analysis and optimization.
-I increasingly suspect the resistance from colleagues we've all encountered towards GP has little to do with our technical results as such.
+I increasingly suspect that social resistance to GP has little to do with the quality of our technical results.
 Rather it arises from unfamiliarity with GP's very particular ``way of working''.
-We are used to it---perhaps to the point of taking it for granted---but the mismatch and resulting response from those less familiar with the system has I think led us to \emph{mask} our sense of it in an effort to fit in.
+We in the field have become used to it---perhaps to the point of taking it for granted---but colleagues in other fields have not.
 
 Briefly, the systemic fault lies in the awful ``scientific method'' that permeates our cultural dialog about the practice of science and engineering.
-You know the one, which is often something along the lines of:
+You know the one, which I can pastiche here as something like:
 
@@
 I'm sure very few scientists or engineers of my acquaintance would admit any \emph{real} project has \emph{ever} followed this narrative in a literal sense.
-But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports:
-
-\begin{quote}
-Based on the body of published work, an insight was had.
+But that story nonetheless informs and constrains much of our work lives, from fund-raising to publishing reports, producing narratives of our work that run more or less like this: ``Based on the body of published work, an insight was had.
 The insight was framed as a formal hypothesis.
 The hypothesis (shaped by current Best Statistical Practices) immediately suggested an experimental design, which design is obvious to anyone familiar with Our Discipline.
-That experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say ``Best  Practices'', right?
-\end{quote}
+That experimental design was undertaken, the data were collected, the hypothesis duly tested, and now we can be confident of its veracity because\ldots{} well, you just heard me say `Best  Practices', right?''
 
 ``Nothing surprising happened while we were working on this project,'' in other words.
-
 Under trivial term substitutions---``cost--benefit analysis'' and ``requirements document'' for ``hypotheses'' and ``experimental design'', for example---the same narrative can be used to describe almost any institutional project management or public policy planning process as well.
@@
 Of course, nobody ``really believes'' this narrative who has ever done the work.
-It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} might be an interesting starting point, I suspect.} And I am not the first to point it out; the history of Philosophy of Science is built primarily from the numerous philosophical challenges to this artificial narrative, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos and Feyerabend in the 1970s, and with many more to be found in the Table of Contents in any Philosophy of Science text.
+It is a matter for another day to draw parallels with the social construction of religious belief.\footnote{Paul Veyne's excellent \emph{Did the Greeks Believe in Their Myths?} \citep{Veyne:1988} might be an interesting starting point, I suspect.} And I am not the first to point it out; the history of Philosophy of Science is built primarily from the numerous philosophical challenges to this artificial narrative, from Peirce and Dewey nearly a century ago, to Kuhn and Lakatos and Feyerabend in the 1970s, and with many more to be found in the Table of Contents in any Philosophy of Science text.
 
-Andrew Pickering is the inspiration for this project, though.
+That said, it is Andrew Pickering who has provided my immediate inspiration for this project.
 
@@
 
-Andrew Pickering's monograph \emph{The Mangle of Practice} is a decade old, but surprisingly little-known outside his discipline of Science Studies.
+Andrew Pickering's monograph \emph{The Mangle of Practice} \citep{Pickering1995} is a decade old, but surprisingly little-known outside his discipline of Science Studies.
 His approach is especially useful here, because I find it captures a surprising amount of our \emph{actual experience} of building and using GP systems.
@@
 First, I should remind us all that the \emph{performance of science} is just that: not an isolated but perceptive mind standing apart from the world, working in an objective and static field of ``externalities'' and ``facts'', but a \emph{performance} done by a human being present in that world.
-In Pickering's  framework, we can say that research proper begins only when the researcher makes some artifact or formal ``machine'' in the world: writes a block of code, designs a technical instrument, considers a particular equation, draws a pencil sketch, or even a has a thoughtful conversation at a conference.
+In Pickering's  framework, we can say that research proper begins only when the researcher makes some artifact or formal ``machine'' in the world: writes a block of code, designs a technical instrument, considers a particular equation, draws a pencil sketch, or simply has a thoughtful conversation at a conference.
 Let me call this artifact \emph{the thing made}.
@@
 
-The inspiration for granting agency to machinic abstractions (or even concrete dynamics) is obvious whenever we hear the language that crops out in the course of our work: the code ``feels wrong''; the mathematics is ``pointing something out''; the machine ``wants to do X instead of Y''.
-Projects in science, engineering and the arts do not proceed from a stage of planning to a stage of implementation, except in the ahistorical mythology of our published papers.
-Pickering's Mangle\footnote{The word ``mangle'' he has chosen is itself interesting and insightful:
+The inspiration for granting agency to machinic abstractions (or even concrete dynamics) is obvious whenever we hear the phrases we utter in the course of our work: the system ``is acting up''; the mathematics is ``pointing something out''; the machine ``wants to do X instead of Y''.
+Projects in science, engineering and the arts do not proceed from a stage of planning to a stage of implementation, except in the ahistorical mythology of our published papers (see \citep{Koutalos:2008} for a particularly good assessment of this from a biologist).
+Pickering's Mangle\footnote{Pickering's word ``mangle'' and the way he came to choose it are a recursive example of the framework itself:
 
@@
 In the traditional linear narrative, we elide the work as it unfolded and re-frame it as a sort of idealized, apersonal Platonic truth: we use the passive voice, we hide the missteps and confusion, after the fact paint a story which flows from vision to plan to success.
-But within the dance of Pickering's Mangle, the degree to which we as researchers can successfully \emph{accommodate} the inevitable resistance
+But within the dance of Pickering's Mangle, the degree to which we as researchers can successfully \emph{accommodate} the resistance we encounter is \emph{exactly} the degree to which we can say we have made progress in our projects.
 
 In a GP setting, the notion of ``machinic agency'' seems much closer to our experience; after all, we are obliged not only a pick or write a specialized formal language to represent the space of solutions, but in every project we must also cobble together some framework of search operators, fitness operators, algorithms and instrumentation.
-But even when we've written all the code and set all the parameters personally, we're \emph{still} willing to speak of a GP run ``doing'' things, as opposed to merely unfolding according to our plan.
+But even when we've written all the code and set all the parameters personally---dotted the Ts and crossed the Is, as it were---we're \emph{still} driven to speak of our GP run ``doing'' things, rather than merely unfolding according to our plan.
 Indeed, if it happens by chance that GP ``runs according to our plan'' then arguably the problem was too \emph{boring} to be worth mentioning\ldots{}.
@@
 This is not merely ``resistance'' of the frustrating kind: we use GP most effectively when we want it to surprise us.
-The ``surprise'' is certainly something we are forced to accommodate with just as much attention and concentration as any more annoying resistance which might be thrown up, for example when we are forced to piece together ``how the hell this thing actually works''.
+The ``surprise'' is certainly something we are forced to accommodate with just as much attention and concentration as any more annoying resistance which might be thrown up, for example when we are forced to figure out how the ``winning solution'' GP has disgorged actually works.
 
-It is worth saying explicitly now, and again later, that by granting the \emph{thing made} a machinic agency of its own, we can frame the problem of ``pathology'' and ``symptoms'' in GP more constructively.
-A GP system does not resist by ``having the wrong population size'' or ``having too high a mutation rate''; those are not \emph{behaviors}, but tiny facets of a complex plan instantiated (to some extent) in a complex dynamical system that is the GP system.
-Rather I would say a GP system resists by \emph{raising concern or causing dissatisfaction in a human user}.
-It is inevitably that \emph{human} observer who produces a productive and accommodating response.
+It is worth saying explicitly now (and then again as many times as necessary) that by granting the \emph{thing made} a machinic agency of its own, we can frame the problem of ``pathology'' and ``symptoms'' in GP more constructively.
+A GP system does not resist by ``having the wrong population size'' or by ``having too high a mutation rate''; those are not \emph{behaviors}, but tiny facets of a complex plan instantiated (to some extent) in a complex dynamical system.
+Rather we should say that a GP system is resisting when it is \emph{raising concern or causing dissatisfaction in its human user}.
+It is inevitably that \emph{human} observer who is driven to the insight needed to provide an accommodating response.
 
@@
 
-The broader field of machine learning seems to take a much more traditional stance towards its subject matter.
+The broader field of machine learning seems to take a much more ``linear'' stance towards its subject matter than we do in ours, in the sense that the pastiche of the ``scientific method'' applies.
 The result of training a neural network or even a random forest on a given data set is not expected to be a \emph{surprise} in any real sense, but rather the reliable and robust end-product of applying numerical optimization to a well-specified mathematical programming problem.
@@
 The space under consideration by GP is not some vector of numerical constants or a binary mask over a suite of input variables, but the \emph{power-set} of inputs, functions over inputs, and higher-order functions over those.
-We who work in the field can be glib about the ``open-endedness'' of GP systems, but that open-endedness puts GP at odds with its supposed machine learning cousins.
+We who work in the field can be glib about the ``open-endedness'' of GP systems, but that open-endedness puts GP in a qualitatively different realm from its machine learning cousins.
 While GP can be used to explore arbitrarily close to some parametric model, its more common use case is exactly the production of \emph{unexpected} insights.
@@
 \item run GP ``on your data''
-\item (whatever happens here, it's not our problem)
+\item (unexpected things happen here, but it's not our problem)
 \item you have solved your problem
@@
 
-This does not arise from within the GP community; it is \emph{exactly} the stance expected in any planning or public policy setting, or in any waterfall management or programming project management setting.
-Twenty years and more of skepticism even from early-adopting users shouldn't be unexpected.
-\emph{Being surprising} might be the worst conceivable outcome for any tool used in a traditional project management setting.
-
-Given that pressure, it's no wonder that so much of GP research is focused on constraining tweaks to bring GP ``into line''.
+This is of course exactly the stance expected in any planning or public policy setting, or any workplace using waterfall project management.
+And as we know from those cultures, ``being surprising'' could be the worst imaginable outcome.
+Given that pressure, it's no wonder that so much of GP research is focused on the discovery of constraining tweaks aimed at bringing GP ``into line'' with more predictable machine learning tools.
 If only GP could be ``tamed'' or made ``adaptive'' so that step (4) above \emph{never happens}\ldots{}.
@@
 That is, that an ``AI candidate'' loses  authenticity as soon as it's ``tweaked'' or ``adjusted'' in the course of operation.
-It is as if every new-born ``AI'' must be quickly jammed into an air-tight computational container and isolated \emph{until it learns to reason by itself}, and for that matter without exceeding a finite computational budget.\footnote{If humans creating real intelligences treated them anything like the way computer scientists insist we treat nascent artificial ones, murder charges would be forthcoming.}
+It is as if every new-born ``AI'' must be quickly jammed into an air-tight computational container and isolated \emph{until it learns to reason by itself}---and for that matter without exceeding a finite computational budget.
 
-Maybe the responsibility for this failure lies neither with the stupid user or the shoddily-made GP system, but in our unwillingness to let them give more useful feedback to one another.
+If humans creating real intelligences (for example, other human beings) treated them anything like the way computer scientists insist we treat nascent artificial intelligences, I have no doubt that the resulting murder convictions would be swift and merciless.
+It is my hope with this contribution to suggest that we might be able to do better than the virtualized serial murder that is our legacy to date.
 
-Consider our typical GP user who is ``not interfering'' with her running GP system; she can only peer at a results file after the fact, and can't fiddle with the ``settings'' while the thing is actually working.
+Consider the poor ``GP user'' that most of our research seems aimed at, one who is carefully ``not interfering'' with her running GP system: she can only peer at a results file after the fact, and can't fiddle with the ``settings'' while the thing is actually working.
 But of course during any given run of 100 generations, \emph{all sorts} of dynamics have happened: crossover, mutation, selection, all the many random choices.
 
-Imagine for a moment we were given perfect access to the entire dynamical pedigree of the unsatisfying results she receives at the end, and were able to backtrack to any point in the run and change a single decision.
+Imagine for a moment if she were given perfect access to the entire dynamical pedigree of the unsatisfying results she receives at the end, and were able to backtrack to any point in the run and \emph{change a single decision}.
 Before that point, it's unclear how badly things will actually turn out at the 100-generation mark; at some point after that juncture, it's obvious to anybody watching that the whole thing's a mess.
+If such miraculous insights were available, then surely her strategy would be one of \emph{intervention}: even if she could only decide after the fact, she should roll back the system to that crucial turning point, make a change aimed at avoiding the mess, and then continue from there.
 
-If such miraculous insights were available, then surely the correct approach would be to intervene and adjust the situation when the crucial point was reached\ldots{} and then continue.
-Lacking (as we do) this miraculous insight, \emph{why then does it seem reasonable to stop any run arbitrarily at a pre-ordained time point and begin again from scratch?}
+Lacking (as we do) this miraculous insight, or the tools for understanding the internal dynamics of any particular GP system, on what grounds does it seem reasonable to stop \emph{any} run arbitrarily at a pre-ordained time point and begin again from scratch?
+Replication, in the sense we are prohibited from reaching in and affecting outcomes, is no better than dice-rolling.
 
-Insofar as GP \emph{surprises us}, and since that is its sole strength over more predictable and manageable frameworks, we must inevitably see a good fraction of those surprises at least in part as disappointments rather than encouraging opportunities to change our plans.
-Let's learn new ways to accommodate those disappointments, and stop trying to make them ``go away''.
+I would much rather say this: Insofar as GP \emph{surprises us}, and since that is its sole strength over more familiar and manageable machine learning frameworks, we must learn to recognize and accommodate the surprises that arise in its use.
+Some surprises will always remain disappointments, but the senseless restriction we impose on engaging the systems we build blocks us from seeing others as \textbf{encouraging opportunities to improve our plans before it's too late}.

### 2015-06-29 tex/Book/gpbook-Tozier-2015-03-23/tozier.tex
final revisions

--- 2015-03-27 tex/Book/gpbook-Tozier-2015-03-23/chapter.tex
@@
 
-\title*{GP As If You Meant It: Real and Imaginary User Experience}
+\title*{GP As If You Meant It:\\An Exercise for Mindful Practice}
 \titlerunning{GP As If You Meant It}
@@
 %\institute{William A. Tozier}
+% Not affiliated with any institution!
 
@@
 
-\abstract{(to be written last)}
+\abstract{In this contribution I present a \emph{kata} called ``GP As If You Meant It'', aimed at advanced users of genetic programming.
+Inspired by \emph{code katas} that are popular among software developers, it's an exercise designed to help participants hone their skills through mindful practice.
+Its intent is to surface certain unquestioned habits common in our field: to make the participants painfully aware of the tacit \emph{justification} for certain GP algorithm design decisions they may otherwise take for granted.
+In the exercise, the human players are charged with trying to ``rescue'' an ineffectual but unstoppable GP system (which is the other ``player''), which has been set up to only use ``random guessing''---but they must do so by \emph{incrementally modifying the search process without interrupting it}.
+The exercise is a game for two players, plus a Facilitator who acts as a referee.
+The human ``User'' player examines the state of the GP run in order to make amendments to its rules, using a very limited toolkit.
+The other ``player'' is the automated GP System itself, which adds to a growing population of solutions by applying the search operators and evaluation functions specified by the User player.
+The User's goal is to convince the System to produce ``good enough'' answers to a target supervised learning problem chosen by the Facilitator.
+To further complicate the task, the User must also provide the  Facilitator with convincing justifications, or \emph{warrants}, which explain each move she makes.
+The Facilitator chooses the initial search problem, provides training data, and most importantly is empowered to \emph{disqualify} any of the User's moves if unconvinced by the accompanying warrants.
+As a result, the User is forced to work around our field's most insidious habit: that of ``stopping it and starting over again with different parameters''.
+In the process of working within these constraints, the participants---Facilitator and User---are made mindful of the habits they have already developed, tacitly or explicitly, for coping with ``pathologies'' and ``symptoms'' encountered in their more typical work with GP.}
 
 \begin{keywords}
-keywords to your chapter, these words should also be indexed
+	mindful practice, design process, coding kata, praxis, Mangle of Practice
 \end{keywords}
-\index{keywords to your chapter}
-\index{these words should also be indexed}
-\\
+\index{mindful practice}
+\index{design process}
+\index{coding kata}
+\index{praxis}
+\index{Mangle of Practice}
+
+%% [version of \today\ at \currenttime]
 

### 2015-11-11 markdown/manuscript.tex
making sure it compiles

@@
 
-The game ``board'' is a \emph{Tableau} with two components: A list of search {\tt operators}, and a two-dimensional spreadsheet-like table which uses {\tt answers} as its row labels and {\tt rubrics} as its column labels.
-Initially the Tableau is empty, except for a single entry on the list of {\tt operators} labeled ``random guess''.
+The game ``board'' is a \emph{Tableau} with two components: A list of search \texttt{operators}, and a two-dimensional spreadsheet-like table which uses \texttt{answers} as its row labels and \texttt{rubrics} as its column labels.
+Initially the Tableau is empty, except for a single entry on the list of \texttt{operators} labeled ``random guess''.
 
-The ``random guess'' operator is set up by the Facilitator before play begins; it produces a single random {\tt answer} for the target problem, with no arguments.
+The ``random guess'' operator is set up by the Facilitator before play begins; it produces a single random \texttt{answer} for the target problem, with no arguments.
 Think of it as equivalent to the ``initialization'' function used to produce a single ``random'' individual in a traditional GP setting.
@@
 During the User's turn, she can examine the state of the Tableau, including any or all of its history, and the algorithms in play, and apply any amount of data analysis or statistical work she wants.
-On her turn she is permitted no more than \emph{two} moves: She can (optionally) code and append one new {\tt operator} to the list of {\tt operators}, and she can (optionally) code and append one new {\tt rubric} column to that table.
+On her turn she is permitted no more than \emph{two} moves: She can (optionally) code and append one new \texttt{operator} to the list of \texttt{operators}, and she can (optionally) code and append one new \texttt{rubric} column to that table.
 
-During the System's turn, it will produce a fixed number of new {\tt answers}.
-It creates each new {\tt answer} by first picking an entry from the list of {\tt operators} with uniform probability.
-When an {\tt operator} is chosen, it executes the indicated algorithm, selecting parents from the {\tt answers} table as needed.
-For each required selection of a parent, the scores recorded in the table of {\tt answers} vs {\tt rubrics} are used (see below).
-Once parents are selected, the {\tt operator} is applied and a new {\tt answer} is immediately appended to that part of the Tableau.
+During the System's turn, it will produce a fixed number of new \texttt{answers}.
+It creates each new \texttt{answer} by first picking an entry from the list of \texttt{operators} with uniform probability.
+When an \texttt{operator} is chosen, it executes the indicated algorithm, selecting parents from the \texttt{answers} table as needed.
+For each required selection of a parent, the scores recorded in the table of \texttt{answers} vs \texttt{rubrics} are used (see below).
+Once parents are selected, the \texttt{operator} is applied and a new \texttt{answer} is immediately appended to that part of the Tableau.
 
@@
 
-What I'm calling an {\tt answer} here would probably be called an ``individual'' in the GP literature.
+What I'm calling an \texttt{answer} here would probably be called an ``individual'' in the GP literature.
 In this case, it is a particular \emph{script} or program in the representation language the Facilitator has chosen for the target problem.
@@
 
-Each {\tt operator} is a function which takes as its argument an unordered  collection of zero or more {\tt answers}, and which produces a new collection of one or more {\tt answers} as output.
+Each \texttt{operator} is a function which takes as its argument an unordered  collection of zero or more \texttt{answers}, and which produces a new collection of one or more \texttt{answers} as output.
 
-The initial Tableau includes only a single {\tt operator}, which implements a pre-coded ``random guessing'' algorithm.
-On the User's move, she may build and launch other more complex (and familiar) {\tt operators} like crossover or mutation.
-The only permitted argument is a set of zero or more {\tt answers}; no numerical or other parameters are allowed.
-A wide variety of {\tt operator} algorithms are still possible, and the details of the code in which they are implemented is left to the Facilitator to specify as part of the game setup.
+The initial Tableau includes only a single \texttt{operator}, which implements a pre-coded ``random guessing'' algorithm.
+On the User's move, she may build and launch other more complex (and familiar) \texttt{operators} like crossover or mutation.
+The only permitted argument is a set of zero or more \texttt{answers}; no numerical or other parameters are allowed.
+A wide variety of \texttt{operator} algorithms are still possible, and the details of the code in which they are implemented is left to the Facilitator to specify as part of the game setup.
 
-Except for some unusual edge cases, the \emph{specific} set of {\tt answers} to which an {\tt operator} will be applied cannot be chosen directly by the User.
-All ``parents'' defined for each {\tt operator} are chosen independently (and incrementally) by \emph{lexicase selection}, using the suite of {\tt rubrics} in play when the System takes its turn.
-This lexicase selection process samples every {\tt rubric} in the Tableau with equal probability.
-It may be possible for the User to design a new ``selection'' {\tt operator} which takes as its argument ``all the answers'' and somehow culls that collection down to a subset---but to do so, it must rely on the immediate state of the {\tt answers} it is given; {\tt rubric} scores are ``stored'' only in the Tableau {\tt answer} table itself, not as gettable attributes of the {\tt answers}, making it very difficult for any new {\tt operator} algorithm to use {\tt rubric} scores in its implementation.
+Except for some unusual edge cases, the \emph{specific} set of \texttt{answers} to which an \texttt{operator} will be applied cannot be chosen directly by the User.
+All ``parents'' defined for each \texttt{operator} are chosen independently (and incrementally) by \emph{lexicase selection}, using the suite of \texttt{rubrics} in play when the System takes its turn.
+This lexicase selection process samples every \texttt{rubric} in the Tableau with equal probability.
+It may be possible for the User to design a new ``selection'' \texttt{operator} which takes as its argument ``all the answers'' and somehow culls that collection down to a subset---but to do so, it must rely on the immediate state of the \texttt{answers} it is given; \texttt{rubric} scores are ``stored'' only in the Tableau \texttt{answer} table itself, not as gettable attributes of the \texttt{answers}, making it very difficult for any new \texttt{operator} algorithm to use \texttt{rubric} scores in its implementation.
 
-Note again: no mechanism exists which \emph{removes} {\tt answers} from the Tableau.
+Note again: no mechanism exists which \emph{removes} \texttt{answers} from the Tableau.
 
@@
 
-A {\tt rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for any given {\tt answer}, conditioned (as needed) on the instantaneous Tableau state.
-When a {\tt rubric} is applied to an {\tt answer}, it sets a new value (or ``score'') for that {\tt answer} in the appropriate cell of the ``spreadsheet'' portion of the Tableau.
+A \texttt{rubric} is a function which returns a scalar value (not necessarily a \emph{number}) for any given \texttt{answer}, conditioned (as needed) on the instantaneous Tableau state.
+When a \texttt{rubric} is applied to an \texttt{answer}, it sets a new value (or ``score'') for that \texttt{answer} in the appropriate cell of the ``spreadsheet'' portion of the Tableau.
 
-No score is ever changed for an {\tt answer}, once it has been set by a {\tt rubric}.
-However, if multiple {\tt answers} exist which have the same {\tt script}, each may be scored at different times or in different contexts, and the resulting values may differ.
+No score is ever changed for an \texttt{answer}, once it has been set by a \texttt{rubric}.
+However, if multiple \texttt{answers} exist which have the same \texttt{script}, each may be scored at different times or in different contexts, and the resulting values may differ.
 
-An aggregate or \emph{higher-order} {\tt rubric} can be created, but its existence ``entails'' all of the component {\tt rubrics} from which its score is derived.
-So, for example, if the User constructs a {\tt rubric} like ``maximum absolute error observed over any of these 30 training cases'', on the turn when she adds that {\tt rubric} up to 31 new columns will be added to the spreadsheet portion of the Tableau: one for each of the 30 requisite ``sub-{\tt rubrics}'', which produce as their scores the absolute error for a single training case, and also one \emph{aggregate} {\tt rubric} which calculates and reports ``maximum of those other 30 scores''.
-Whenever the User submits such a higher-order {\tt rubric} on her turn, all of the entailed {\tt rubrics} are added automatically if they do not already exist.
-Note however that there must still be a \emph{warrant} for the {\tt rubric}, which convinces the Facilitator of its potential usefulness.
+An aggregate or \emph{higher-order} \texttt{rubric} can be created, but its existence ``entails'' all of the component \texttt{rubrics} from which its score is derived.
+So, for example, if the User constructs a \texttt{rubric} like ``maximum absolute error observed over any of these 30 training cases'', on the turn when she adds that \texttt{rubric} up to 31 new columns will be added to the spreadsheet portion of the Tableau: one for each of the 30 requisite ``sub-\texttt{rubrics}'', which produce as their scores the absolute error for a single training case, and also one \emph{aggregate} \texttt{rubric} which calculates and reports ``maximum of those other 30 scores''.
+Whenever the User submits such a higher-order \texttt{rubric} on her turn, all of the entailed \texttt{rubrics} are added automatically if they do not already exist.
+Note however that there must still be a \emph{warrant} for the \texttt{rubric}, which convinces the Facilitator of its potential usefulness.
 
-The User can use any information present in the Tableau in building new {\tt rubric} functions (such as row numbers, string values of {\tt answers}, or scores), and can also examine the detailed state of the interpreter before and after the {\tt script} is run.
-It should be clear therefore that the User can specify {\tt rubric} functions which score aspects of the problem such as:
+The User can use any information present in the Tableau in building new \texttt{rubric} functions (such as row numbers, string values of \texttt{answers}, or scores), and can also examine the detailed state of the interpreter before and after the \texttt{script} is run.
+It should be clear therefore that the User can specify \texttt{rubric} functions which score aspects of the problem such as:
 
@@
 \item the absolute error for a single training case
-\item the number of tokens in the {\tt answer}'s script
-\item maximum error measured in any of 35 other {\tt rubrics}
-\item number of {\tt div0} errors produced when running a script with a particular set of inputs
-\item number of stochastic instructions appearing in the {\tt answer}'s script
-\item rewrite difference between the scored script the most common {\tt answer} in the population
+\item the number of tokens in the \texttt{answer}'s script
+\item maximum error measured in any of 35 other \texttt{rubrics}
+\item number of \texttt{div0} errors produced when running a script with a particular set of inputs
+\item number of stochastic instructions appearing in the \texttt{answer}'s script
+\item rewrite difference between the scored script the most common \texttt{answer} in the population
 \item (and so on)
@@
 Various problem-specific aspects are glossed here, and it is left to the Facilitator to be reasonable in the context of the target problem and the representation she has chosen.
-Suffice to say, the construction of useful {\tt rubrics} in response the System's moves is the core of the User player's game strategy.
+Suffice to say, the construction of useful \texttt{rubrics} in response the System's moves is the core of the User player's game strategy.
 
@@
 
-All {\tt operator} inputs are chosen by lexicase selection. Slightly simplifying Spector's original description \citep{Spector:2012:GECCOcompA}, the following algorithm can be used:
+All \texttt{operator} inputs are chosen by lexicase selection. Slightly simplifying Spector's original description \citep{Spector:2012:GECCOcompA}, the following algorithm can be used:
 
 \begin{itemize}
-\item (beginning with a ``set under consideration'' which includes all {\tt answers}\ldots{})
-\item for each {\tt rubric} in a random permutation of all the {\tt rubrics} in the Tableau, discard all {\tt answers} from the set under consideration whose score on this {\tt rubric} is sub-optimal (relative to the current set under consideration)
-\item if multiple {\tt answers} remain after all {\tt rubrics} have been applied, return one {\tt answer} picked randomly, with uniform probability, from the remaining set under consideration
+\item (beginning with a ``set under consideration'' which includes all \texttt{answers}\ldots{})
+\item for each \texttt{rubric} in a random permutation of all the \texttt{rubrics} in the Tableau, discard all \texttt{answers} from the set under consideration whose score on this \texttt{rubric} is sub-optimal (relative to the current set under consideration)
+\item if multiple \texttt{answers} remain after all \texttt{rubrics} have been applied, return one \texttt{answer} picked randomly, with uniform probability, from the remaining set under consideration
 \end{itemize}
 
-At the beginning of the game when no {\tt rubrics} have been added and no {\tt operators} exist which require ``parents'', no selection occurs or needs to occur.
-But note that the algorithm as described would \emph{still} provide a parent, even when no {\tt rubrics} have been specified: an empty set of {\tt rubrics} would be immediately exhausted without eliminating any {\tt answers}, and then a single {\tt answer} would be selected with uniform probability from that complete set.
+At the beginning of the game when no \texttt{rubrics} have been added and no \texttt{operators} exist which require ``parents'', no selection occurs or needs to occur.
+But note that the algorithm as described would \emph{still} provide a parent, even when no \texttt{rubrics} have been specified: an empty set of \texttt{rubrics} would be immediately exhausted without eliminating any \texttt{answers}, and then a single \texttt{answer} would be selected with uniform probability from that complete set.
 
-Lexicase selection has several characteristics which argue for its use here; in particular it is interesting because it seems (anecdotally, but see also Helmuth in this volume) to be a relatively ``slow'' selection method, permitting a diversity of {\tt answers} to coexist in a population at the same time.
+Lexicase selection has several characteristics which argue for its use here; in particular it is interesting because it seems (anecdotally, but see also Helmuth in this volume) to be a relatively ``slow'' selection method, permitting a diversity of \texttt{answers} to coexist in a population at the same time.
 However I admit readily that I've specified it as a core part of this \emph{kata} because it is \emph{unfamiliar to most prospective players}, and therefore more likely to produce unexpected behavior of a useful sort.
@@
 \begin{enumerate}
-\item add one {\tt operator} to the list in the Tableau
-\item add one {\tt rubric} to the Tableau, which can be ``higher-order'' and therefore entail others
+\item add one \texttt{operator} to the list in the Tableau
+\item add one \texttt{rubric} to the Tableau, which can be ``higher-order'' and therefore entail others
 \end{enumerate}
@@
 
-Suppose, for example, that the User has not yet added \emph{any} {\tt rubrics}, and would like to add a new one that scores {\tt answers} for a single training case.
-Since the System is simply making new {\tt answers} with the ``random guess'' {\tt operator}, it seems perfectly reasonable to warrant this new {\tt rubric} simply by pointing out that selection can't drive search towards better answers without at least one training case being used.
+Suppose, for example, that the User has not yet added \emph{any} \texttt{rubrics}, and would like to add a new one that scores \texttt{answers} for a single training case.
+Since the System is simply making new \texttt{answers} with the ``random guess'' \texttt{operator}, it seems perfectly reasonable to warrant this new \texttt{rubric} simply by pointing out that selection can't drive search towards better answers without at least one training case being used.
 
-Suppose further that later in the same game the User has added more {\tt rubrics}, such that more than 100 training cases are being used to select parents for the several {\tt operators} in play.
-In this context, the reasons given for adding a new {\tt rubric} that specifies \emph{just that one particular training case} must surely be very different: perhaps there is a problematic region of the response surface, or a different training case that needs additional ``support'' to differentiate between two close input states (and so forth).
+Suppose further that later in the same game the User has added more \texttt{rubrics}, such that more than 100 training cases are being used to select parents for the several \texttt{operators} in play.
+In this context, the reasons given for adding a new \texttt{rubric} that specifies \emph{just that one particular training case} must surely be very different: perhaps there is a problematic region of the response surface, or a different training case that needs additional ``support'' to differentiate between two close input states (and so forth).
 
 That said, a warrant does \emph{not} need to be ``technical'' or even ``rigorous'', but merely robust enough to be convincing \emph{in the moment}.
-For example, ``I made a crossover {\tt operator} because I think we need to search for new answers `in between' the parents,'' sounds to me like a shoddy excuse that invokes received wisdom.
+For example, ``I made a crossover \texttt{operator} because I think we need to search for new answers `in between' the parents,'' sounds to me like a shoddy excuse that invokes received wisdom.
 On the other hand, ``I made crossover so we can drop the variance on this rubric and foster inbreeding of \emph{these} solutions \emph{here},'' given a glance at the Tableau and charts on hand, should be more convincing.
@@
 That is, for any given algorithmic goal, there should be multiple paths to success.
-So for example if the problem's solution could reasonably be expected to involve ordered lists, then it would be best if the language had at least two different ways to ``use lists'', for example with iterators, recursion, a comprehensive set of second-order functional operators, an explicit {\tt List} type with associated methods, and so forth.
+So for example if the problem's solution could reasonably be expected to involve ordered lists, then it would be best if the language had at least two different ways to ``use lists'', for example with iterators, recursion, a comprehensive set of second-order functional operators, an explicit \texttt{List} type with associated methods, and so forth.
 
@@
 \begin{itemize}
-\item the only {\tt operator} is ``random guess'', which creates one new {\tt answer} with an arbitrary script
-\item no {\tt rubrics} are present
+\item the only \texttt{operator} is ``random guess'', which creates one new \texttt{answer} with an arbitrary script
+\item no \texttt{rubrics} are present
 \item the System player moves first
-\item there is no mechanism for \emph{removing} {\tt answers} from the tableau
-\item the System player \emph{always} uses lexicase selection, \emph{always} choses {\tt operators} with equal probability from the current list, and \emph{always} uses all {\tt rubrics} in the Tableau
-\item a {\tt rubric} can only be \emph{run} on a given {\tt answer} once; stochastic scripts will only be sampled one time, and no {\tt rubric} score is ever recalculated after the first time, though multiple copies of the same stochastic {\tt answer} will probably end up with different scores in the same field
+\item there is no mechanism for \emph{removing} \texttt{answers} from the tableau
+\item the System player \emph{always} uses lexicase selection, \emph{always} choses \texttt{operators} with equal probability from the current list, and \emph{always} uses all \texttt{rubrics} in the Tableau
+\item a \texttt{rubric} can only be \emph{run} on a given \texttt{answer} once; stochastic scripts will only be sampled one time, and no \texttt{rubric} score is ever recalculated after the first time, though multiple copies of the same stochastic \texttt{answer} will probably end up with different scores in the same field
 \end{itemize}
@@
 
-During its turn, the System player will add a specified number of new {\tt answers} to the Tableau, one at a time, using a simple form of lexicase selection.
+During its turn, the System player will add a specified number of new \texttt{answers} to the Tableau, one at a time, using a simple form of lexicase selection.
 Until it reaches its halting state it iterates this cycle:
@@
 \begin{enumerate}
-\item select one {\tt operator} from those in the Tableau, with equal probability
-\item apply lexicase selection to select the required number of input {\tt answers}
-\item apply the chosen {\tt operator} to the inputs {\tt answers} to produce one or more new {\tt answers}, and append those new {\tt answers} immediately to the Tableau
-\item {\tt HALT} if the number of new {\tt answers} meets or exceeds the limit, and delete any extra {\tt answers} that exceed the limit; otherwise, go to step (1)
+\item select one \texttt{operator} from those in the Tableau, with equal probability
+\item apply lexicase selection to select the required number of input \texttt{answers}
+\item apply the chosen \texttt{operator} to the inputs \texttt{answers} to produce one or more new \texttt{answers}, and append those new \texttt{answers} immediately to the Tableau
+\item \texttt{HALT} if the number of new \texttt{answers} meets or exceeds the limit, and delete any extra \texttt{answers} that exceed the limit; otherwise, go to step (1)
 \end{enumerate}
 
-The number of {\tt answers} created in each turn should be enough to have a chance of providing new and useful information to the User, and not so much that the System state grows out of control.
-I would suggest 100 or 500 {\tt answers} per turn; this is about the size of the typical ``population'' for most GP users, and is on a comfortable scale for them.
+The number of \texttt{answers} created in each turn should be enough to have a chance of providing new and useful information to the User, and not so much that the System state grows out of control.
+I would suggest 100 or 500 \texttt{answers} per turn; this is about the size of the typical ``population'' for most GP users, and is on a comfortable scale for them.
 
